# Causal Models {#models}

```{r packagesused02, include = FALSE}
source("_packages_used.R")
```
:::: {.headerbox data-latex=""}
We provide a lay-language primer on the logic of causal models.
::::

<br>

Causal claims are everywhere. Causal knowledge is often the end goal of  empirical social science. It is also a key *input* into causal inference.^[As nicely put by @cartwright1994nature: no causes in, no causes out. We return to the  point  more formally later.] 
 Causal assumptions are also hidden in seemingly descriptive statements: claims that someone is guilty, or exploited, or powerful, or weak, involve beliefs about how things would be were conditions different. Even when scholars carefully try to avoid causal claim-making, causal verbs---depends, drives, produces, influences---tend to surface.     

But while causal claims are commonplace, it is not always clear (1) what exactly is meant by a causal relation and (2) how causal knowledge about one thing can be marshaled to justify causal claims about another. For our purposes, the counterfactual view of causality addresses the first question. Causal models address the second.

<!-- In this chapter we provide a basic introduction to causal models. Subsequent chapters in Part I layer on other foundational components of the book's framework, including a causal-model-based understanding of theory, the definition of common causal estimands within causal models, and the basics of Bayesian inference.  -->
<!-- While here we focus on the formal definition of causal models, in Chapter 10 we discuss strategies for generating them.  -->


## The counterfactual model {#counterfactualmodel}

We begin with what we might think of as a meta-model, the counterfactual model of causation.
<!-- The counterfactual model is perhaps the dominant approach to causal relations in the social sciences.  -->
At its core, a counterfactual understanding of causation captures a simple notion of causation as "difference-making."^[The approach is sometimes attributed to David Hume, whose writing contains ideas both about causality as regularity and causality as counterfactual. On the latter, Hume's key formulation is, "if the first object had not been, the second never had existed" [@hume2000enquiry, Section VIII]. More recently, the counterfactual view has been set forth by @splawa1990application and @lewis1973counterfactuals. See also  @lewis1986causation.] In the counterfactual view, to say that $X$ caused $Y$ is to say: *had* $X$ been different, $Y$ *would have been* different. 

The causal effect, in this view, is the difference between two things that might have happened.  This means that *by definition, causal effects are not measurable quantities*. They are not differences between possible observations in the world, but, at best, differences between outcomes in the world and counterfactual outcomes. They  need to be inferred not measured.

Moreover, in this view, the antecedent, "had $X$ been different,"  imagines a *controlled* change in $X$---an intervention that altered $X$'s value---rather than a naturally arising difference in $X$. The counterfactual claim, then, is not that $Y$ is different in those cases in which $X$ is different; it is, rather, that if one could somehow have *made* $X$ different, $Y$ would have been different. In the terminology of @pearl2000causality, we represent this quantity using a "do" operator:  $Y(do(X=x))$ is the value of $Y$ when $X$ is *set* to $x$.

Consider a simple example.  Students with teacher A perform well without studying. Students with teacher B perform well if they study, and do not perform well if they do not study. Moreover, only students with teacher B in fact study. And all perform well. 

When we say that one of teacher B's students did well *because* they studied, we are comparing the outcome that they experienced to the outcome they would have experienced if they had had teacher B (as they did) but (counterfactually) had not studied. Notably, we are *not* comparing their realized outcome to the outcome they would have experienced if they had been among the people that in fact didn't study (i.e., if they had had teacher A). 

A second example. Consider the claim that Switzerland democratized ($D=1$) because it had a relatively low level of economic inequality ($I=0$) (drawing on the logic of @boix2003democracy). In the counterfactual view, this is equivalent to saying that, had Switzerland *not* had a high level of equality, the country would not have democratized. High economic equality made a difference. The comparison for the causal statement is with the outcome Switzerland would have experienced under an intervention that boosted its historical level of economic inequality---*not* with how Switzerland would have performed if it had been one of the countries that *in fact* had higher levels of inequality, cases that likely differ from Switzerland in other causally relevant ways. 

<!-- AJ: see Slack thread about whether we're expressing the controlled change right. https://gbiqq.slack.com/archives/C02C1DHUJ1M/p1629990240000600 -->
<!-- MH: OK -->


Along with this notion of causation as difference-making, we also want to allow for *variability* in how $X$ acts on the world. $X$ might sometimes make a difference, for some units of interest, and sometimes not. High levels of equality might generate democratization in some countries or historical settings but not in others. Moreover, while equality might make democratization happen in some times in places, it might *prevent* that same outcome in others. We need a language to describe these different types of relations.


### Potential outcomes

The "potential outcomes" framework is useful for describing the different kinds of counterfactual causal relations that might prevail between variables @Rubin1974. In this framework we characterize how a given unit responds to a causal variable by positing the outcomes that it *would* take on at different values of the causal variable.

A setting in which it is quite natural to think about potential outcomes is medical treatment. Imagine some individuals in a diseased population are observed to have received a drug ($X=1$) while others have not ($X=0$). Assume that, subsequently, a researcher observes which individuals become healthy ($Y=1$) and which do not ($Y=0$). Given the assignments of all other individuals,^[We noted that we are conditioning on the assignments of others. If we wanted to describe outcomes as a function of the *profile* of treatments received by others we would have a more complex types space. For instance in an $X \rightarrow Y$ model with 2 individuals we would report how $(Y_1, Y_2)$ respond to $(X_1,X_0)$; each vector can take on four values producing a type space with $4^4$ types rather than $2^2$. The complex type space  could be reduced back down to four types again, however, if we  invoked the assumption that the treatment or non-treatment of one patient has no effect on the outcomes of other patients---an assumption known as the stable unit treatment value assumption (SUTVA).] we can treat each individual as belonging to one of four unobserved response "types," defined by the outcomes that the individual *would have* if they received or did not receive treatment:^[See @copas1973randomization} for an early classification of this form. The literature on probabilistic models also refers to such strata as "canonical partitions" or "equivalence classes."]

<!-- AJ: SUTVA Footnote above needs to be made clearer. -->
<!-- MH Check -->

* **a**dverse: Those individuals who would get better if and only if they do not receive the treatment
* **b**eneficial: Those who would get better if and only if they do receive the treatment
* **c**hronic: Those who will remain sick whether or not they receive treatment
* **d**estined: Those who will get better whether or not they receive treatment

Table \@ref(tab:PO) maps the four types ($a, b, c, d$) onto their respective potential outcomes. In each column, we have simply written down the outcome that a patient of a given type would experience if they are not treated, and the outcome they would experience if they are treated. In each cases we are imagining *controlled* changes in treatment: the responses if treatments are changed without changes to other background conditions about the case.  

<!-- \begin{table}[h!]
\begin{tabular}{l|cccc} \small

& Type a & Type b & Type c & Type d \\

& **a**dverse effects & **b**eneficial Effects & **c**hronic cases & **d**estined cases \\
\hline
Not treated &    Healthy &       Sick &       Sick &    Healthy \\
Treated &       Sick &    Healthy &       Sick &    Healthy \\
\end{tabular}  
\caption{Potential Outcomes: What would happen to each of four possible types of case if they were or were not treated.}
\label{tabPO}
\end{table} -->


| \small      |          Type a          |            Type b           |         Type c         |          Type d         |
|-------------|:------------------------:|:---------------------------:|:----------------------:|:-----------------------:|
|             | **a**dverse | **b**eneficial | **c**hronic | **d**estined  |
| Outcome if not treated |          Healthy         |             Sick            |          Sick          |         Healthy         |
| Outcome if treated     |           Sick           |           Healthy           |          Sick          |         Healthy         |

Table: (\#tab:PO). Potential outcomes: What would happen to each of four possible types of case if they were or were not treated.


We highlight that, in this framework, case-level causal relations are treated as deterministic. A given case has a set of potential outcomes. Any uncertainty about outcomes enters as incomplete knowledge of a case's "type," not from underlying randomness in causal relations.  This understanding of causality---as ontologically deterministic, but empirically imperfectly understood---is compatible with views of causation commonly employed by qualitative researchers (see, e.g., @mahoney2008toward), and with understandings of causal determinism going back at least to @laplace1901philosophical. 

As we will also see, we can readily express this kind of incompleteness of knowledge within a causal model framework: indeed, the way in which causal models manage uncertainty is central to how they allow us to pose questions of interest and to learn from evidence. There are certainly situations we could imagine in which one might want to conceptualize potential outcomes themselves as random (for instance, if individuals in different conditions play different lotteries). But for the vast majority of the settings we condsider, not much of importance is lost if we treat potential outcomes as deterministic but possibly unknown: at the end of the day something will occur or it will not occur, we just do not know which it is.

### Generalization

Throughout the book, we generalize from this simple setup. Whenever we have one causal variable and one outcome, and both variables are binary (i.e., each can take on two possible values, 0 or 1), there are only four sets of possible potential outcomes, or "types." More generally, for variable, $Y$, we will use $\theta^Y$ to capture the unit's "type": the way that $Y$ responds to its potential causes.^[Later, we will refer to these as "nodal types."] We, further, add subscripts to denote particular types. Where there are four possible types, for instance, we use the notation $\theta^Y_{ij}$, where $i$ represents the case's potential outcome when $X=0$ and $j$ is the case's potential outcome when $X=1$.

Adopting this notation, for a causal structure with one binary causal variable and a binary outcome, the four types can be represented as $\{\theta^Y_{10},   \theta^Y_{01},   \theta^Y_{00}, \theta^Y_{11}\}$, as shown in Table \@ref(tab:POGEN):

| \small      |          Type a          |            Type b           |         Type c         |          Type d         |
|-------------|:------------------------:|:---------------------------:|:----------------------:|:-----------------------:|
|             | $\theta^Y=\theta^Y_{10}$ |  $\theta^Y=\theta^Y_{01}$   |$\theta^Y=\theta^Y_{00}$| $\theta^Y=\theta^Y_{11}$|
| Set $X=0$   |     $Y(0)=1$             |       $Y(0)=0$              |      $Y(0)=0$          |      $Y(0)=1$           |
| Set $X=1$   |     $Y(1)=0$             |       $Y(1)=1$              |      $Y(1)=0$          |      $Y(1)=1$           |
Table: (\#tab:POGEN). Generalizing from Table \@ref(tab:PO), the table gives for each causal type the values that $Y$ would take on if $X$ is set at $0$ and if $X$ is set at 1.



<!-- Let $Y(x)$ denote the "potential" outcome (the value $Y$ would take on) when $X=x$. Then, if $X$ is a binary variable, the effect of $X$ on $Y$ is simply defined as $Y(1) -  Y(0)$.  -->

<!-- These types differ in their ''potential outcomes'' | that is on what outcomes, $Y$,  they *would* have depending on their treatment condition $X$ . More formally, we let $Y(x)$ denote a case or type's potential outcome  when $X=x$. Thus, the potential outcomes are $Y(0)=1, Y(1)=0$ for type $a$; $Y(0)=0, Y(1)=1$ for type $b$; $Y(0)=0, Y(1)=0$ for type $c$; and $Y(0)=1, Y(1)=1$ for type $d$.   -->


<!-- For any given response type, the causal effect of $X$ on $Y$ is $Y(1) -  Y(0)$. Thus, the causal effect is $-1$ for $a$ types, $1$ for $b$ types, and $0$ for both $c$ and $d$ types. -->

Returning to the matter of inequality and democratization to illustrate, let $I=1$ represent a high level of economic inequality and $I=0$ its absence; let $D=1$ represent democratization and $D=0$ its absence. A $\theta^D_{10}$ (or $a$) type is a case in which a high level of inequality, if it occurs, *prevents* democratization in a country that would otherwise have democratized. The causal effect of high inequality in a case, $i$, of $\theta^D_{10}$ type is $\tau_i= -1$. A $\theta^D_{01}$ type (or $b$ type) is a case in which high inequality, if it occurs, generates democratization in a country that would otherwise have remained non-democratic (effect $\tau_i= 1$). A $\theta^D_{00}$  type ($c$ type) is a case that will not democratize regardless of the level of inequality (effect $\tau_i = 0$); and a $\theta^D_{11}$ type ($d$ type)  is one that will democratize regardless of the level of inequality (again, effect $\tau_i= 0$).

In this setting, a causal *explanation* of a given case outcome amounts to a statement about its type. The claim that Switzerland's low level of equality was a cause of its democratization is equivalent to saying that Switzerland democratized and is a $\theta^D_{10}$ type. To claim that Sierra Leone democratized because of low inequality is equivalent to saying that Sierra Leone democratized and is a $\theta^D_{01}$ type. To claim, on the other hand, that Malawi democratized for reasons having nothing to do with its level of economic inequality is to characterize Malawi as a $\theta^D_{11}$ type (which implies that Malawi would have been democratic no matter what its level of inequality).

<!-- FLAG: CHECK USE OF THESE CASES -->

Now, let us consider more complex causal relations. Suppose now that there are two binary causal variables $X_1$ and $X_2$. We can specify any given case's potential outcomes for each of the different possible combinations of causal conditions---there now being four such conditions, as each causal variable may take on $0$ or $1$ when the other is at $0$ or $1$. 

As for notation, we now need to expand $\theta$'s subscript since we need to represent the value that $Y$ takes on under each of the four possible combinations of $X_1$ and $X_2$ values.  We construct the four-digit subscript with the ordering (and, in general, mapping any two "parents" alphabetically into $X_1$ and $X_2$). The next equation connects this notation to the "do" notation used to convey the idea that conditions are controlled.

<!-- FLAG: ADD CROSSREF -->

<!--
\begin{eqnarray} 
Y_{hijk} \left\{\begin{array}{ccc}  h& =& Y|do(X_1=0, X_2=0)\\
 i &=& Y|do(X_1=1, X_2=0)\\
 j &=& Y|do(X_1=0, X_2=1)\\
 k &=& Y|do(X_1=1, X_2=1)
 \end{array} \right.
\end{eqnarray}
-->

\begin{align} 
Y_{hijk} \left\{\begin{array}{ccc}  h& =& Y|do(X_1=0, X_2=0)\\
 i &=& Y|do(X_1=1, X_2=0)\\
 j &=& Y|do(X_1=0, X_2=1)\\
 k &=& Y|do(X_1=1, X_2=1)
 \end{array} \right.
\end{align}

Thus, for instance, $\theta^Y_{0100}$ means that $Y$ is 1 if $X_1=1$ and $X_2=0$ and is 0 otherwise. 

We now have 16 causal types: 16 different patterns that $Y$ might display in response to changes in $X_1$ and $X_2$. The full set is represented in Table \@ref(tab:PO16), which also makes clear how we read types off of four-digit subscripts. For instance, the table shows us that for nodal type  $\theta^Y_{0101}$,  $X_1$ has a positive causal effect on $Y$ but $X_2$ has no effect. On the other hand, for type $\theta^Y_{0011}$, $X_2$ has a positive effect while $X_1$ has none. 

We also capture interactions here. For instance, in a $\theta^Y_{0001}$ type, $X_2$ has a positive causal effect if and only if $X_1$ is 1. In that type, $X_1$ and $X_2$ serve as "complements." For $\theta^Y_{0111}$, $X_2$ has a positive causal effect if and only if $X_1$ is 0. In that setup, $X_1$ and $X_2$ are "substitutes."



<!-- let $Y(x_1, x_2)$ denote the outcome when $X_1=x_1$ and  $X_2=x_2$. Then the quantity $\left(Y(1, 1) - Y(0, 1)\right) - \left(Y(1, 0) - Y(0, 0)\right)$ describes the interactive effect of two treatments: it captures how the effect of $X_1$ changing from $0$ to $1$ is different between those situations in which $X_2=1$ and those situations in which $X_2=0$. -->



```{r PO16, echo = FALSE, warning = FALSE}
tab <- get_nodal_types(make_model("X1 -> Y <- X2"), collapse = FALSE)$Y

tab <- cbind(type = paste0("$\\theta^Y_{", rownames(tab), "}$"), tab)

kabble(tab, 
      col.names = c("$\\theta^Y$", "if $X_1=0, X_2=0$",   "if $X_1=1,X_2=0$",  "if $X_1=0,X_2=1$",   "if $X_1=1, X_2=1$" ),
      row.names = FALSE, align=c(rep('c',ncol(tab))),
      caption = "With two binary causal variables, there are 16 causal types: 16 ways in which $Y$ might respond to changes in the other two variables.", format = 'pandoc')
```



This is a rich framework in that it allows for all possible ways in which a set of multiple causes can interact with each other. Often, when seeking to explain the outcome in a case, researchers proceed as though causes are necessarily *rival*, where $X_1$ being a cause of $Y$ implies that $X_2$ was not. Did Malawi democratize because it was a relatively economically equal society *or* because of international pressure to do so? In the counterfactual model, however, causal relations can be non-rival. If two out of three people vote for an outcome under majority rule, for example, then both of the two supporters caused the outcome: the outcome would not have occurred if *either* supporter's vote were different. This typological, potential-outcomes conceptualization provides a straightforward way of representing this kind of complex causation.

Because of this complexity, when we say that $X$ caused $Y$ in a given case, we will generally mean that $X$ was *a* cause, not *the* (only) cause. Malawi might not have democratized if *either* a relatively high level of economic equality *or* international pressure had been absent. For most social phenomena that we study, there will be multiple, and sometimes a great many, difference-makers for any given case outcome.

We will mostly use $\theta^Y_{ij}$-style notation in this book to refer to types. We will, however, occasionally revert to the simpler $a, b, c, d$ designations when that helps ease exposition. As types play a central role in the causal-model framework, we recommend getting comfortable with both forms of notation before going further.

<!-- ## Counter-intuitive implications of the counterfactual model -->

<!-- Although the counterfactual framework is now widely employed, it contains a set of implications that might sit uncomfortably with common conceptions of how causal inference operates. -->

<!-- Second, it is often intuitive to think of causal processes as sets of transitive relations: if $A$ causes $B$ and  $B$ causes $C$, then we might think that $A$ causes $C$. Yet, in the counterfactual model, causal relations are *not* transitive. In a classic illustration, imagine a boulder that rolls down a hill, causing you to duck, and that ducking in turn saves your life. As a counterfactual matter, the boulder clearly caused the ducking and the ducking your survival. But the boulder rolling down the hill did not save your life. To consider an example from the social realm, think about situations in which action begets reaction. For instance, a rebellion causes a military crackdown, and the military crackdown causes the regime to survive; yet the rebellion did not cause the regime to survive. (For discussions see @hall2004two and @paul2013causation.) The insight has implications both for process tracing and for correlational approaches to establishing causation. Finding in a causal link in a case from $A$ to $B$ and a causal link from $B$ to $C$ is not equivalent to finding that $A$ caused $C$. Likewise, identifying a population-level causal effect of $X$ on some suspected mediator $M$, and another effect of $M$ on $Y$, does not establish an $X\rightarrow Y$ causal relationship. -->

<!-- <!-- I do not find the next point convincing. --> 

<!-- Third, notions of causality going back at least to Hume [@hume2000enquiry] treat causal relations as characterized by spatio-temporal contiguity between cause and effect, or at least of the intermediate steps between them. Yet in the counterfactural model, there is no requirement that causes be temporally or spatially connected to their effects. For instance, *potentially* intervening events that did *not* occur can have causal effects, even though they make no spatio-temporal contact with the observable events that seem to lie along the path from $X$ to $Y$. The plague that put Friar John into quarantine meant that he did not deliver the letter to Romeo to inform him that Juliet was not dead, which in turn led to Romeo's death. There is a *causal* path from the plague to Romeo's death, but no *spatio-temporal* one.  -->

<!-- Fourth, hypothesis-testing at the case level sometimes proceeds as though competing explanations amount to rival causes, where $A$ caused $B$ implies that $C$ did not. But in the counterfactual model, causal relations are neither rival nor decomposable. If two out of three people vote for an outcome under majority rule, for example, then both of the two supporters caused the outcome: the outcome would not have occurred if *either* supporter's vote were different. The causes are not rival. Now, imagine that all three of three voters supported an outcome. Then the three votes jointly caused the outcome. However, this joint cause is not decomposable into its component parts: none of the individual votes had *any* effect on the outcome. A change in any one vote would have made no difference. -->

<!-- <!-- Does the latter example above  satisfy minimality? Is a graph minimal that points from all three yesses to the outcome? --> 

<!-- Thus, there appear to be some tensions between the counterfactual model and some common notions of causality. These tensions largely disappear, however, once we properly specify causal models as systems of causal relations. For this work, Directed Acyclic Graphs provide a powerful tool.    -->

Using the same framework, we can generalize to structures in which a unit has any number of causes and also to cases in which causes and outcomes are non-binary. As one might imagine, the number of types increases rapidly (very rapidly) as the number of considered causal variables increases, as it also does as we allow $X$ or $Y$ to take on more than 2 possible values. For example, if there are $n$ binary causes of an outcome, then there can be $2^{\left(2^n\right)}$ types of this form: that is, $k=2^n$ combinations of values of causes to consider, and $2^k$ distinct ways to react to each combination. If causes and outcomes are ternary instead of binary, we have $3^{\left(3^n\right)}$ causal types of this form. Yet, the basic principle of representing possible causal relations as patterns of potential outcomes remains unchanged, at least as long as variables are discrete.

### Summaries of potential outcomes

So far, we have focused on causal relations at the level of an individual case. Causal relations at the level of a population are, however, simply a summary of causal relations for cases, and the same basic ideas can be used. We could, for instance, summarize our beliefs about the relationship between economic equality and democratization by saying that we think that the world is comprised of a mixture of $a$, $b$, $c$, and $d$ types, as defined above. We could get more specific and express a belief about what proportions of cases in the world are of each of the four types. For instance, we might believe that $a$ types and $d$ types are quite rare while $b$ and $c$ types are quite common. Moreover, our belief about the proportions of $b$ (positive effect) and $a$ (negative effect) cases imply a belief about equality's *average* effect on democratization as, in a binary setup, this quantity is simply the proportion of $b$ types minus the proportion of $a$ types. Such summaries allow us to move from discussion of the cause of a single outcome to discussions of average effects, a distinction that we take up again in Chapter \@ref(questions). 

## Causal Models and Directed Acyclic Graphs

So far we have discussed how a single outcome is affected by one or more possible causes. However, these same ideas can be used to describe more complex relations between collections of variables---for example, with one variable affecting another directly as well as indirectly via its impact on some mediating variable.

Potential outcomes tables can be used to describe such complex relations. However, as causal structures become more complex---especially, as the number of variables in a domain increases---a causal model can be a powerful organizing tool. In this section, we show how causal models and their visual counterparts, directed acyclic graphs (DAGs), can represent substantive beliefs about counterfactual causal relationships in the world. The key ideas in this section can be found in many texts (see, e.g., @halpern2005causesa and @galles1998axiomatic), and we introduce here a set of basic principles that readers will need to keep in mind in order to follow the argumentation in this book. 

As we shift to talking about networks of causal relations between variables we will also shift our language. When talking about causal networks, or causal graphs, we will generally refer to variables as "nodes." And we will sometimes use familial terms to describe relations between nodes. For instance, if $A$ is a cause of $B$ we will refer to $A$ as a parent and $B$ as a child and represent this with an arrow from $A$ to $B$. If $X_1$ and $X_2$ have a child in common (both directly affecting the same variable) we refer to them as "spouses." We can also say that $I$ is an "ancestor" of $D$ (a node upstream from $D$'s parent) and conversely that $D$ is a descendant of $I$ (a node downstream from $I$'s child).

Return to our running democratization example, but suppose now that we have more fully specified beliefs about how the level of economic inequality can have an effect on whether a country democratizes. We might believe that inequality affects the likelihood of democratization by generating demands for redistribution, which in turn can cause the mobilization of lower-income citizens, which in turn can cause democratization. We might also believe that mobilization itself is not just a function of redistributive preferences but also of the degree of ethnic homogeneity, which shapes the capacities of lower-income citizens for collective action. In this model $R$ is a parent of $M$, $I$ is an ancestor of $M$ but not a parent. $R$ and $E$ are spouses and $M$ is their child. We can visualize this model as a Directed Acyclic Diagram (DAG) in Figure \@ref(fig:simpleDAG).

```{r simpleDAG, echo = FALSE, out.width='60%', fig.width = 7, fig.height = 4,  fig.align="center", out.width='80%', fig.cap = "A simple causal model in which high inequality ($I$) affects democratization ($D$) via redistributive demands and mass mobilization ($M$), which is also a function of ethnic homogeneity ($E$). The arrows show relations of causal dependence between variables.  The graph does not capture the ranges of the variables and the functional relations between them."}
hj_ggdag(x = c(0, 1, 2, 1, 3, 3, 2, 1, 0, 1),
       y = c(2, 2, 2, 0, 2, 3, 3, 3, 3, -1),
       names = c(
         "I", 
         "R",
         "M",
         "E", 
         "D", 
         expression(paste(theta^D)),
         expression(paste(theta^M)),
         expression(paste(theta^R)),
         expression(paste(theta^I)),
         expression(paste(theta^E)) 
         ),
       arcs = cbind( c(1, 2, 4, 3, 6, 7, 8, 9, 10),
                     c(2, 3, 3, 5, 5, 3, 2, 1, 4)),
       title = "A Model of Inequality's Effect on Democratization",
       padding = .2)

```

Fundamentally, we treat causal models in this book as formal representations of *beliefs* about how the world works---or, more specifically, about causal relations within a given domain. We might use a causal model to capture our own beliefs, a working simplification of our beliefs, or a set of potential beliefs that one might hold. The formalization of *prior* beliefs in the form of a causal model is the starting point for research design and inference in this book's analytic framework. Using the democratization example, we will now walk through the three components of a causal model in which our beliefs get embedded: nodes, functions, and distributions.

### Components of a Causal Model 

The three components of a causal model are (i) the nodes---that is, the set of variables we are focused on and how are they defined (ii) the functional relations---which nodes are "explained" by which other nodes and how, and (iii) probability distributions over unexplained elements of a model. 

#### The nodes

The first component of a causal model is the set of variables (nodes) across which the model characterizes causal relations. On the graph in Figure \@ref(fig:simpleDAG), the five included variables are represented by the five lettered nodes. (In addition we mark $\theta^D$ on the graph though, as will be made clear, we will not think of this as a variable.)

In identifying the nodes, we also need to specify the *ranges* over which they can  vary. We might specify, for instance, that all nodes in the model are binary, taking on the values 0 or 1. We could, alternatively, define a set of categories across which a node ranges or allow a node to take on any real number value or any value between a set of bounds. ^[If we let \(\mathcal{R}\) denote a set of ranges for all nodes in the model, we can indicate $X$'s range, for instance, by writing \(\mathcal{R}(X)=\{0,1\}\). The nodes in a causal model together with their ranges---the triple \((\mathcal{U}, \mathcal{V}, \mathcal{R})\)---are sometimes called a \emph{signature}, \(\mathcal{S}\).]

Notice that some of these nodes have arrows pointing *into* them: $R, M$, and $D$ are endogenous nodes, meaning that their values are determined entirely by other nodes in the model.

Other nodes have arrows pointing out of them but no arrows pointing into them: $I$ and $E$.  $I$ and $E$ are "exogenous" nodes; they influence other nodes in the model but themselves have no causes specified in the model. 

The $\theta$ terms require a little more explanation since they do not describe substantive nodes. In our discussion above, we introduced $\theta$ notation for representing types. Generally we can think of  $\theta$ terms on a causal graph as unobservable and unspecified features of a causal domain that affect outcomes. These might include random processes (noise) or contextual features that we are unable to identify or do not understand. We imagine them pointing into everynode, whether indicated or not. In all cases the work they do is to characterize what the value of a node will be, given the value of its parents.

We note that our notation for representing these unobservable, unspecified influences differs from that commonly found in the literature on causal models. In many treatments, these components are themselves referred to as "exogenous" variables, and often labelled as sets $\mathcal U$, to be distinguished from the endogenous---named--variables often labelled as $\mathcal V$. We will generally use $\theta$ to denote these unobserved, unspecified influences to emphasize their particular role, as direct objects of interest in causal inquiry. As we will show, we can think of $\theta$ nodes as capturing the functional relations between endogeneous variables and as being quantities of direct interest for causal inquiry. We more fully develop this point---returning to the notion of $\theta$ terms as receptacles for causal effects---below.

<!-- For instance, here $\theta^D$ captures *how* and *whether* $D$'s parent produces $D$. In the present example, we believe democratization to be potentially affected by mobilization, and two of $\theta^D$'s possible values allow for such an effect: $\theta^D_{01}$ and $\theta^D_{10}$. We also know that democratization is affected by other things, even if we do not know what they are and have not represented them explicitly in the model. The values $\theta^D_{00}$ (no democratization regardless of mobilization) and $\theta^D_{11}$ (always democratization regardless of mobilization) capture all "other" factors---factors other than mobilization---that may affect democratization.  -->

<!-- Similarly, $\theta^M$ represents how $M$ responds to $R$ and $E$, while $\theta^R$ captures how $R$ responds to $I$. The $\theta^I$ and $\theta^E$ terms, pointing into the two exogenous nodes, represent all processes outside the model that assign these exogenous nodes to their values. -->

<!-- ^[Readers familiar with the literature on causal models will know that $U$ terms are typically used to capture unspecified exogenous influences on included nodes. We could have, here, included a term $U_D$ to indicate an "error" term or uncertainty regarding exactly what value $D$ will take given knowledge of $M$. We use the concept of a type and the $\theta$ notation, instead, to highlight the fact that, in non-parametric models, this "residual" component can be thought of as *the* locus of learning about the questions we are asking.] -->

<!-- Going forward, we set every node in our working example to be binary. -->

#### The functions

Next, we need to specify our beliefs about the causal relations among the nodes in our model. How is the value of one node affected by, and how does it affect, the values of others? For each endogenous node---each node influenced by others in the model---we need to express beliefs about how its value is affected by its parents, its immediate causes.

The DAG already represents a critical part of these beliefs: the arrows, or directed edges, tell us *which nodes we believe to be direct causal inputs into other nodes*. So, for instance, we believe that democratization ($D$) is determined jointly by mobilization ($M$) and some exogenous, unspecified factor (or set of factors), $\theta^D$. As we have said, we can think of $\theta^D$ as all of the other influences on democratization, besides mobilization, that we either do not know of or have decided not to explicitly include in the model. We believe, likewise, that $M$ is determined by $I$ and an unspecified exogenous factor (or set of factors), $\theta^M$. And we are conceptualizing inequality ($I$) and ethnic heterogeneity ($E$) as shaped solely by factors exogenous to the model, captured by $\theta^I$ and $\theta^E$, respectively.

<!-- (For all intents and purposes, $I$ and $E$ behave as  exogenous node here since its value is determined solely by an exogenous node.)  -->

<!-- We emphasize that an arrow indicates a *direct* relationship, given the other nodes in the graph. Likewise, the *absence* of an arrow between two nodes indicates the belief that there is no direct relationship between the them. Consider, for instance, the absence of a an arrow running direcly from $R$ to $D$. This omission represents the belief that any effect of redistributive preferences on democratization work *only* through mass mobilization; or, equivalently, that $R$ has no effect on $D$ *given* $M$. The omission of an $R \rightarrow D$ arrow means we are building quite a strong assumption into the model; the missing arrow rules out, for example, the possibility that $R$ moderates the effect of $M$ on $D$.  -->

Beyond the beliefs captured by the arrows in a DAG, we can express more specific beliefs about causal relations in the form of a causal function.^[The collection of all causal functions in the model can be denoted as $\mathcal{F}$.] Specifying a function means writing down whatever general or theoretical knowledge we have about the direct causal relations between nodes. A function specifies how the value that one node takes on is determined by the values that other nodes---its parents---take on. 

<!-- For each endogenous node, we can specify two kinds of nodes as direct causes:  (i) other endogenous nodes, which we call the node's *parents*;^[For node $V_i$, we write its parents as $PA_i$.] and (ii) an exogenous node. Thus, for instance, the node $Y$ in Figure \@ref(fig:simpleDAG) has as its direct causes the node $R$, its parent (an endogenous node itself) and the random-disturbance, $U_Y$.^[Any node with no parents in $\mathcal V$ must be a function of a member of $\mathcal U$; otherwise, we could not consider it endogenous. A node that is a function of one or more members of $\mathcal V$, however, can be modeled without a $U_i$ term if we believe that it is fully determined by nodes specified in $\mathcal V$.]  -->

We can specify this relationship in a vast variety of ways. It is useful however to distinguish broadly between parametric and non-parametric approaches. We take a non-parametric approach in this book, but it is helpful to juxtapose that approach to a parametric one.

**Parametric approaches.** A parametric approach specifies a functional form that relates parents to children. For instance, we might  model one node as a linear function of another and  write $D=\alpha + \beta M$, where $\beta$ is a parameter that we do not know the value of at the outset of a study but about which we wish to learn. If we believe $D$ to be linearly affected by $M$ but also subject to forces that we do not yet understand and have not yet specified in our theory, then we might write: $D=\alpha + \beta M+\theta^D$. In this function, $\alpha, \beta$ might be the parameters of interest, with $\theta^D$ treated merely as a random disturbance.  We can be still more agnostic by, for example, including parameters that govern how other parameters operate. Consider, for instance the function, $D=\beta M^{\theta^D}$. Here, $D$ and $M$ are linearly related if $\theta^D=1$, but not otherwise. 

Note that functions can be written to be quite specific or extremely general, depending on the state of prior knowledge about the phenomenon under investigation. The use of a structural model *does not require precise knowledge of specific causal relations*, even of the functional forms through which two nodes are related. 


<!-- ^[Here the difference between $\beta$ and $U_D$ is that $\beta$ is a parameter that we believe takes a constant value for all units, even if its value is unknown, while $U_D$ represents some unknown factor or combination of factors the value of which may vary across units, over a pre-specified range.] -->


<!-- * We may even be uncertain about whether or in what direction one node affects another. Or we may believe that their relationship varies across cases for reasons that we do not yet understand. We can capture this type of uncertainty as an interaction such as: $D=M U_D$. Here, the value of our random-disturbance term does not merely represent noise around an $M \rightarrow D$ relationship. Rather, $U_D$ now conditions, or moderates, the strength, sign, or existence of the relationship itself. Which of these $U_D$ conditions will depend on the $U_D$'s specified range. For instance, if we allow $U_D$ to take on the values $-1, 0$ or $1$, then $U_D$ will determine whether $M$ has a negative effect, no effect, or a positive effect on $D$. If instead, $U_D$ is allowed to vary continuously between $0$ and $1$, inclusive, then it conditions the strength and existence of a positive causal effect of $M$ on $D$, though not the sign of that effect.  As we discuss in later chapters, our investigation might then center on drawing inferences about $U_D$ in order to assess $M$'s causal effect. -->


**The non-parametric approach.** With discrete data, causal functions can take fully *non-parametric* form, allowing for *any possible relation* between parents and children. We use this framework for most of this book and thus spend some time developing the approach here.

We begin by returning to the concept of types. Drawing on our original four types from earlier in this chapter, we know that we can fully specify causal relations between a binary $M$ and a binary $D$ by allowing the node $\theta_D$ to range across four possible values $\{\theta^D_{10}, \theta^D_{01}, \theta^D_{00}, \theta^D_{11}\}$. For instance, $\theta^D_{10}$ represents a negative causal effect of $M$ on $D$ while $\theta^D_{00}$ represents $D$ remaining at 0 regardless of $M$. Put differently, $\theta^D$ represents the non-parametric function that relates $M$ to $D$. We can formally specify $D$'s behavior as a function of $M$ and $\theta^D$ in the following way: 

$$D(M, \theta^D_{ij}) = \left\{\begin{array}{ccc} i & \text{if} & M=0 \\ j & \text{if} & M=1 \end{array}\right.$$

Note that $\theta^D$ ranges over *all possible* functional forms between these two binary variables. 

How should we think about what kind of *thing* $\theta^D$ is, in a more substantive sense? It is probably most helpful to think of $\theta^D$ as an unknown and possibly random factor that conditions the effect of mobilization on democratization, determining whether $M$ has a negative effect, a positive effect, no effect with democratization never occurring, or no effect with democratization bound to occur regardless of mobilization.  Importantly, however, while we might think of $\theta^D$ as an unknown or random quantity, in this formulation $\theta^D$ should not be thought of as a nuisance---as "noise" that we would like to get rid of---but as *the quantity that we want to learn about*: we want to know whether $M$ likely had a positive, negative, or no effect on $D$. We elaborate on this point at much greater length in Chapter \@ref(questions).

We can similarly use $\theta$ terms to capture causal relations involving any number of parent nodes. Every substantively defined node, $J$, in a graph can be thought of as having a $\theta^J$ term pointing into it, and the (unobservable) value of $\theta^J$ represents the mapping from $J$'s parents (if it has any) to the value of $J$. 


<!-- With two parent nodes, for instance, we simply use causal types of the form $\theta^Y_{hijk}$, as illustrated above.  -->

<!-- We may, for instance, believe that an outcome occurs when and only when _two_ conditions are present. Redistributive demands and ethnic homogeneity may be individually necessary and jointly sufficient conditions for mobilization. We can express this belief with a slightly more complex function: $M=E R$. According to this function, $M=1$ (mobilization occurs) if _both_ $E=1$ (ethnic homogeneity is present) and $R=1$ (redistributive preferences are present), but $M=0$ (mobilization does not occur) otherwise. Note that this formulation also builds in causal heterogeneity: here, we are saying that redistributive preferences have an effect on mobilization when and only when ethnic homogeneity is present, and vice versa. -->



Applied to the binary nodes in Figure \@ref(fig:simpleDAG), $\theta^J$ ranges as follows:

* **Nodes with no parents**: For an exogenous node, like $I$ or $E$, $\theta^J$ represents an external "assignment" process that can take on one of two values, $\theta^J_{0}$, meaning that $J$ is "assigned" to $0$ or $\theta^J_{1}$, meaning that $J$ is assigned to 1. For instance, $\theta^I_{0}$ describes a case in which exogenous forces have generated low inequality.
* **Binary nodes with 1 binary parent**: For endogenous node $R$, with only one parent ($I$), $\theta^R$ takes on one of four values of the form $\theta^R_{ij}$ (our four original types, $\theta^R_{10}$, $\theta^R_{01}$, etc.).
* **Binary nodes with 2 binary parents**: $M$ has two parent nodes. Thus, $\theta^M$ will take on a possible 16 values of the form $\theta^M_{hijk}$ ($\theta^M_{0000}$, $\theta^M_{0001}$, etc.), using the syntax detailed earlier in this chapter.




**Nodal types and causal types.** For analytic applications later in the book, we will want to be able to think both about the type operating at a particular *node* and about *collections* of types operating across a model. We thus refer to $\theta^J$ as a unit's *nodal* causal type, or simply nodal type, for $J$. We refer to the collection of nodal types across all nodes for a given unit (i.e., a case) as the case's *unit causal type*, or simply *causal type*, denoted by the vector $\theta$. 

<!-- AJ: Should we not be using vector notation above -- i.e.,\Theta? -->
<!-- MH: I prefer \theta here; not shouting. $\theta = (\theta^X, \theta^Y)$ -->

If we hypothetically knew a unit's causal type---all nodal types for all nodes---then we would know everything there is to know about that unit. Since the nodal types of exogenous nodes include *values* of all exogenous nodes, and the nodal types of all endogenous nodes specify how those nodes respond to all of their parents, a unit's causal type fully specifies all nodal values as well as all *counterfactual* nodal values for a unit.

We will sometimes refer to the values of $\theta$ as a unit's *context*. This is because $\theta$ captures all exogenous forces acting on a unit. This includes the assignment process driving the model's exogenous nodes (in our example, $\theta^I$ and $\theta^E$) as well as all contextual factors that shape causal relations between nodes ($\theta^R$, $\theta^M$, and $\theta^D$). Put differently, $\theta$ captures both how a unit reacts to situations and which situations it is reacting to. One implication is that there is no *formal* distinction between a unit's type and a unit's situation---between, say, a hungry person, and a person who has had no food.





**term**          |**symbol** |**meaning**
:-------------:|:---------:|:-----------------------------------------:
nodal type        |$\theta^J$ | The way that node $J$ responds to the values of its parents. Example: $\theta^Y_{10}$: $Y$ takes the value 1 if $X=0$ and 0 if $X=1$.  
causal type       |$\theta$   | A causal type is a concatenation of nodal types, one for each node. Example:  $(\theta^X_0, \theta^Y_{00})$, is a causal type that has $X=0$ and that has $Y=0$ no matter what the value of $X$. 

Table: (\#tab:NTCT). Nodal types, causal types.


<!-- It is thus worth dwelling for a moment on what this kind of function is doing. We have started with a graph in which mobilization can have an effect on democratization and the understanding that this effect, both its existence and its sign, may vary across cases. Cases, in other words, may be of different causal types. Further, we do not know what it is that shapes $D$'s response to $M$---what makes a case one type versus another. We thus use $\theta_D$ as a stand-in for the unknown and unspecified moderators of $M$'s effect. We might, at this stage, wonder what the point is of including $\theta_D$ in the model; are we not essentially just placing a question mark on the graph? We are, and that is precisely the point. As we will see in later chapters, non-substantive, causal-type nodes can play a key role in specifying (a) what we are uncertain about in a causal network and (b) what we would like to find out. Embedding our questions about the world directly into a model of the world, in turn, allows us to answer those questions in ways systematically and transparently guided by prior knowledge. -->


<!-- In fact, we can include factors as another node's "parents" even if we are unsure that those factors matter. Including nodes on the righthand side in a functional equation allows for the possibility that those nodes matter, and in turn sets us up to investigate their possible effect empirically.^[For instance, in $B=AU_B+C$, $A$ will only affect $B$ if $\theta^B$ takes on a non-zero value. ] -->

<!-- ### Interpretation of functional equations -->

A few important aspects of causal functions are worth highlighting. 

1. These functions express *causal* beliefs. When we write $D=\beta M$ as a function, we do not just mean that we believe the values of $M$ and $D$ in the world to be linearly related. We mean that we believe that the value of $M$ *determines* the value of $D$ through this linear function. Functions are, in this sense, *directional* statements, with causes on the righthand side and an outcome on the left.

2. The collection of simple functions that map from the values of parents of a given node to the values of that node are sufficient to represent potentially complex webs of causal relations. For each node, we do not need to think through entire sequences of causation that might precede it. We need only specify how we believe it to be affected by its parents---that is to say, those nodes pointing directly into it. Our outcome of interest, $D$, may be shaped by multiple, long chains of causality. To theorize how $D$ is generated, however, we write down how we believe $D$ is shaped by its parent---its direct cause, $M$. We then, separately, express a belief about how $M$ is shaped by _its_ parents, $R$ and $E$. A node's function must include as inputs all, and only, those nodes that point directly into that node.^[The set of a node's parents is required to be minimal in the sense that a node is not included among the parents if, given the other parents, the child does not depend on it in any state that arises with positive probability.]

3. As in the general potential-outcomes framework, all relations in a causal model are conceptualized as deterministic at the case level. Yet, there is not as much at stake here as one might think at first; by this we simply mean that a node's value is  *determined* by the values of its parents *along with* any stochastic or unknown components. We express uncertainty about causal relations, however, as unknown parameters, such as the causal types $\theta$. 

<!-- 4. *The values of the exogenous nodes*---those with no arrows pointing in to them---*are sufficient to determine the values of all other nodes in the model.*  In other words, context determines all other values. For instance, in Figure \@ref(fig:simpleDAG), knowing the values of $I$, $E$, and $\theta^D$ as well as the causal functions (including the values of any parameters they contain) would tell us the values of $R$, $M$, and $D$. -->

<!-- ]  nodes that have no parents are called *roots*.^[Thus in our usage all elements of $\mathcal{U}$ are roots, but so are nodes in $\mathcal{V}$ that depend on nodes in $\mathcal{U}$ only.]  We will say that $\mathcal{F}$ is a set of *ordered structural equations* if no node is its own descendant and if no element in $\mathcal{U}$ is parent to more than one element of \(\mathcal{V}\).^[This last condition can be achieved by shifting any parent of multiple children in $\mathcal{U}$ to $\mathcal{V}$.] -->

<!-- Do we need to define roots? -->

<!-- For notational simplicity we generally write functional equations in the form $c(a,b)$ rather than $f_c(a,b)$. -->

<!-- **The distributions**. So far, we have specified the nodes in the model and their causal relations, possibly with uncertainty. These relations imply  -->

<!-- In general, $U$ terms---capturing unspecified disturbances---represent features of the world that we are *not* able to directly observe. In some cases, we may not even have a specific conceptualization of the phenomena in the world to which a $U$ term corresponds. Nonetheless, we may be able to make *inferences* about the value of a $U$ term from observed data. Indeed, given the role of the exogenous terms in a model in determining the operation of the world that the model describes, learning about a case's context becomes central to model-based causal inquiry and, as we shall see, lies at the heart of the framework that we are elaborating. -->

#### The distributions

Putting causal structure and causal functions together gives us a *structural causal model.* In a structural causal model, all endogenous nodes are, either directly or by implication, functions of a case's context (the values of the set of exogenous nodes).^[More formally, a **structural causal model** *over* signature $\mathcal{S}=<\mathcal{U},\mathcal{V},\mathcal{R}>$ is a pair $<\mathcal{S}, \mathcal{F}>$, where $\mathcal{F}$ is a set of ordered structural equations containing a function  $f_i$  for each element $Y\in \mathcal{V}$. We say that $\mathcal{F}$ is a set of ordered structural equations if no node is its own descendant and if no element in $\mathcal{U}$ is parent to more than one element of \(\mathcal{V}\). This last condition can be achieved by shifting any parent of multiple children in $\mathcal{U}$ to $\mathcal{V}$. This definition thus includes an assumption of acyclicity, which is not found in all definitions in the literature.] What we have not yet inscribed into the model, however, is beliefs about how *likely* or *common* different kinds of contexts might be. 

Thus, for instance, a structural causal model consistent with Figure \@ref(fig:simpleDAG) stipulates which nodes may have effects on which other nodes. But it says nothing in itself about the distribution of values of either the exogenous nodes or of the causal relations between nodes.^[Thus $P(d|i,e, u_D)$ would defined by this structural model (as a degenerate distribution), but $P(i)$, $P(e)$, $P(u_D)$, and $P(i,e, u_D)$ would not be.] We have not said anything, for instance, about how common high inequality is across the relevant domain of cases or how common ethnic homogeneity is. Put differently, we have said nothing about the *distribution* of $\theta^I$ or of $\theta^E$. Similarly, we have said nothing yet about how commonly mobilization has positive, negative, or null effects of democratization---that is, the distribution of $\theta^D$---or about how commonly $I$ and $E$ have different possible joint causal effects on $M$ (the distribution of $\theta^M$). 

In many research situations, we will have or want to posit a set of prior beliefs about how the world works under different conditions and about what kinds of conditions are more likely than others. We can express these beliefs about context as probability distributions over the model's $\theta^J$ terms. For instance, our structural causal model might tell us that $E$ and $R$ can jointly affect $M$. We might, then, add to this a belief about $\theta^M$ such that, in the population of interest, redistribution rarely has a positive effect on mobilization when ethnic homogeneity is low. We would thus be putting a low probability on the nodal types for $M$ in which $R$ has a positive effect on $M$ when $E=0$, relative to $M$'s other nodal types.^[Ordering the parent nodes alphabetically, the types we would be setting to a low probability would be $\theta^M_{0010}, \theta^M_{0110}, \theta^M_{0111}, \theta^M_{0011}$.] 

We might add to this the belief that $E=1$ in only 10\% of cases in the population of interest, thus setting a 0.1 probability on $\theta^E_1$. Note that these two beliefs jointly imply that $R$ will rarely have a positive effect on $M$. 

As with functions, we can also (and typically would) build uncertainty into our beliefs about the shares of different nodal types in the population. We do this by specifying a *distribution* over possible "share" allocations.^[More strictly our uncertainty is over probabilities. However it is sometimes more intuitive to describe uncertainty over shares. The distinction is not important for the applications later in which we typically assume units are independently drawn from a large population.] For instance, we can specify a distribution over the shares of cases with ethnic homogeneity ($\theta^E_1$), and a distribution over the shares of $\theta^M$ types, with our degrees of uncertainty captured by each distribution's variance. (More on these distributions in Chapter \@ref(bayeschapter).)



In the default setup, we assume that each $\theta$ term ($\theta^I, \theta^E, \theta^R$, etc.) is generated independently of the others.
<!-- ---that each nodal type for a given case is drawn from an independent distribution. -->

<!-- FLAG: Cross ref for boxes -->

While this is not without loss of generality, it is not as constraining as it might at first appear: any graph in which two $\theta$ terms are *not* independent can be replaced by a graph in which these two terms are themselves generated by a common, third $\theta$ term.^[Operationally, in the `CausalQueries` package, we can specify nodal types as having joint distributions.] This independence feature is critical for being able to read off relations of conditional independence from a graph (see Box below). If it cannot be defended then the graph needs to be modified to communicate that $\theta$s are not independent, typically using two headed arrows. More on this in section \@ref(graphing).  
<!-- Note also that one could envision "incomplete probabilistic causal models" in which researchers claim knowledge regarding distributions over *subsets* of $\theta$.  -->

<!-- With our non-parametric representation of functional forms, we let $\lambda_j^X$ denote the probability that $\theta^X = \theta^X_j$. For instance in a simple $X \rightarrow Y$ model, $\lambda^Y_{01}$ denotes the probability that $\theta^Y = \theta^Y_{01}$.  -->


<!-- FLAG: FLESH OUT INCLUDING HOW CONFOUNDING IS TREATED -->

<!-- Am trying to render all the indented paragraphs below as a single, multi-paragraph footnote. Have read up and tried all sorts of things, no luck yet. -->

<!-- Thus, a probabilistic causal model is a structural causal model coupled with a probability distribution over the model's exogenous nodes. A corresponding probabilistic model, however, might support a stronger claim of the form: "Condition $C$ arises with frequency $\pi^C$, and so $X$ causes $Y$ with probability $\pi^C$." -->

:::: {.headerbox data-latex=""}
::: {.center data-latex=""  #markov}
**Technical Note on the Markov Property**
:::

The assumptions that no node is its own descendant and that the $\theta$ terms are generated independently make the model *Markovian*, and the parents of a given node are Markovian parents. Knowing the set of Markovian parents allows one to write relatively simple factorizations of a joint probability distribution, exploiting the fact ("the Markov condition") that all nodes are *conditionally independent* of their nondescendants, conditional on their parents: nodes $A$ and $B$ are "conditionally independent" given $C$ if $P(a|b,c) = P(a|c)$ for all values of $a, b$ and $c$.  
  To see how this Markovian property allows for simple factorization of $P$ for Figure \@ref(fig:simpleDAG), note that $P(X, R, Y)$ can always be written as: 
  $$P(X, R, Y) = P(X)P(R|X)P(Y|R, X)$$ 
  If we believe, as in the figure, that $X$ causes $Y$ only through $R$ then we have the slightly simpler factorization:
  $$P(X, R, Y) = P(X)P(R|X)P(Y|R)$$ 
  Or, more generally:

\begin{equation} 
P(v_1,v_2,\dots v_n) = \prod P(v_i|pa_i)
(\#eq:markov)
\end{equation} 

  The distribution $P$ on $\theta$ induces a joint probability distribution on $\mathcal{V}$ that captures not just information about how likely different states are to arise but also the relations of conditional independence between nodes that are implied by the underlying causal process. For example, if we thought that $X$ caused $Y$ via $R$ (and only via $R$), we would then hold that $P(Y | R) = P(Y | X, R)$: in other words if $X$ matters for $Y$ only via $R$ then, conditional on $R$, $X$ should not be informative about $Y$.   
  In this way, a probability distribution $P$ over a set of nodes can be consistent with some causal models but not others. This does not, however, mean that a specific causal model can be extracted from $P$. To demonstrate with a simple example for two nodes, any probability distribution on $(X,Y)$ with $P(x)\neq P(x|y)$ is consistent both with a model in which $X$ is a parent of $Y$ and with a model in which $Y$ is a parent of $X$.
::::

<br>


Once we introduce beliefs about the distribution of values of the exogenous terms (in our setup, the $\theta$ terms) in a model, we have specified a *probabilistic causal model.* We need not say much more, for the moment, about the probabilistic components of causal models. But to foreshadow the argument to come, our prior beliefs about the likelihoods of different contexts play a central role in the framework that we develop here. We will see how the encoding of contextual knowledge---beliefs that some kinds of conditions and causal effects are more common than others---forms a key foundation for causal inference. At the same time, our expressions of *uncertainty* about context represent scope for learning: it is the very things that we are, at a study's outset, uncertain about that we can update our beliefs about as we encounter evidence.


## Graphing models and using graphs

While we have been speaking to causal graphs throughout this chapter, we want to take some time to unpack their core features and uses. A key benefit of causal models is that they lend themselves to graphical representations; in turn, graphs constructed according to particular rules can aid causal analysis. In the next subsection we discuss a set of rules for representing a model in graphical form. The following subsection then demonstrates how access to a graph facilitates causal inference.

### Rules for graphing causal models {#graphing}

The diagram in Figure \@ref(fig:simpleDAG) is a causal DAG [@hernan2006instruments]. We endow it with the interpretation that an arrow from a parent to a child means that a change in the parent can, under some circumstances, induce a change in the child. Though we have already been making use of this causal graph to help us visualize elements of a causal model, we now explicitly point out a number of general features of causal graphs as we will be using them throughout this book. Causal graphs have their own distinctive "grammar," a set of rules that give them important analytic features.

**Directed, acyclic.** A causal graph represents elements of a causal model as a set of nodes (or vertices), representing nodes, connected by a collection of single-headed arrows (or directed edges).  We draw an arrow from node $A$ to node $B$ if and only if we believe that $A$ can have a direct effect on $B$. The resulting diagram is  a *directed acyclic* graph (DAG) if there are no paths along directed edges that lead from any node back to itself---i.e., if the graph contains no causal cycles. The absence of cycles (or "feedback loops") is less constraining than it might appear at first. In particular if one thinks that $A$ today causes $B$ tomorrow which in turn causes $A$ today, we can represent this as $A_1 \rightarrow B \rightarrow A_2$ rather than $A \leftrightarrow B$. That is, we timestamp the nodes, turning what might informally appear as feedback into a non-cyclical chain.

**Meaning of missing arrows.**  The *absence* of an arrow between $A$ and $B$ means that $A$ is not a direct cause of $B$.^[By "direct" we mean that the $A$ is a parent of $B$: i.e., the effect of $A$ on $B$ is not fully mediated by one or more other nodes in the model.] Here lies an important asymmetry: drawing an $A \rightarrow B$ arrow does not mean that we know that $A$ *does* directly cause $B$; but omitting such an arrow implies that we know that $A$ does *not* directly cause $B$. We say more, in other words, with the arrows we omit than with the arrows that we include.

Returning to Figure \@ref(fig:simpleDAG), we have here expressed the belief that redistributive preferences exert no direct effect on democratization; we have done so by *not* drawing an arrow directly from $R$ to $D$. In the context of this model, saying that redistributive preferences have no direct effect on democratization is to say that any effect of redistributive preferences on democratization *must* run through mobilization; there is no other pathway through which such an effect can operate. This might be a way of encoding the knowledge that mass preferences for redistribution cannot induce autocratic elites to liberalize the regime absent collective action in pursuit of those preferences. 

The same goes for the effects of $I$ on $M$, $I$ on $D$, and $E$ on $D$: the graph in Figure \@ref(fig:simpleDAG) implies that we believe that these effects also do not operate directly, but only along the indicated, mediated paths.

**Sometimes-causes.** The existence of an arrow from $A$ to $B$ does not imply that $A$ always has a direct effect on $B$. Consider, for instance, the arrow running from $R$ to $M$. The existence of this arrow requires that $R$ appears somewhere in $M$'s functional equation, as a node's functional equation must include all nodes pointing directly into it. Imagine, though, that $M$'s causal function were specified as: $M = RE$. This function would allow for the *possibility* that $R$ affects $M$, as it will whenever $E=1$. However, it would also allow that $R$ will have no effect, as it will when $E=0$. 

<!-- This example also, incidentally, demonstrates another important consequence of context, the values of the exogenous nodes: a case's context determines not just the settings on the endogenous nodes, but also the causal *effects* that prevail among the nodes. Under the functional equation $M=RE$, a case's ethnic-compositional context determines whether or not redistributive preferences will have an effect on mobilization.  -->

<!-- **Representing $U$ on the graph.** As a matter of convention, explicitly including $U$ terms is optional. In practice, $U$'s are often excluded from the visual representation of a model on the understanding that every node on the graph is subject to some unaccounted-for influence and thus, implicitly, has a $U$ term pointing into it. In this book, we will generally draw the $U$ terms where they are of particular theoretical or analytic interest but will otherwise omit them. Whether we include or omit $U$ terms, we will generally treat those nodes in a graph that have no arrows pointing into them as the exogenous nodes that define the context. -->

<!-- AJ: Have commented out above because I think we've sufficiently dealt with the $U$'s and it's going to be very confusing to say U's are implicit when we've said we're using theta's instead of U;s. -->
<!-- MH OK -->

**No excluded common causes.** Any cause common to multiple nodes on a graph must itself be represented on the graph. If $A$ and $B$ on a graph are both affected by some third node, $C$, then we must represent this common cause. Thus, for instance, the graph in Figure \@ref(fig:simpleDAG) implies that $I$ and $E$ have no common cause. If in fact we believed that a country's level of inequality and its ethnic composition were both shaped by, say, its colonial heritage, then this DAG would *not* be an accurate representation of our beliefs about the world. To make it accurate, we would need to add to the graph a node capturing that colonial heritage and include arrows running from colonial heritage to both $I$ and $E$. 

This rule ensures that the graph captures all potential correlations among nodes that are implied by our beliefs. If $I$ and $E$ are in fact driven by some common cause, then this means not just that these two nodes will be correlated but also that each will be correlated with any consequences of the other. For instance, a common cause of $I$ and $E$ would also imply a correlation between $R$ and $E$. $R$ and $E$ are implied to be independent in the current graph but would be implied to be correlated if a common node pointed into both $I$ and $E$.

Of particular interest in Figure \@ref(fig:simpleDAG) is the implied independence of $\theta^J$'s from one another. Imagine, for instance, that the distribution of $\theta^D$ were different if $I=0$ or $I=1$. This would represent a classic form of confounding: the assignment of cases to values on the explanatory node would be correlated with the case's potential outcomes on $D$. The omission of any such pathway is precisely equivalent to expressing the belief that $I$ is exogenous, i.e., (as if) randomly assigned.

**Representing unobserved confounding.**  It may be however that there are common causes for nodes that we simply do not understand. We might believe that some unknown factor (partially) determines both $I$ and $D$, which is the same as saying that $\theta^I$ and $\theta^D$ are not independently distributed. If we were to represent the $\theta$ terms on the graph we might then want to represent a single term $(\theta^I, \theta^D)$ that points into both $I$ and $D$. Usually however the $\theta$ terms are omitted from graphs and in this case we would  represent the unobserved confounding by adding a dotted line, or a two headed arrow, connecting nodes whose unknown components are not independent. Figure \@ref(fig:simpleDAGb) illustrates. We address this kind of unobserved confounding later in the book and show how we can seek to learn about the joint distribution of nodal types in such situations. 

```{r simpleDAGb, echo = FALSE, fig.cap="A DAG with unobserved confounding", out.width = "60%", fig.height = 4, fig.width = 6, fig.align="center"}
hj_ggdag(model=make_model("I -> D; I <-> D"))
```


**Licence to exclude nodes.** The flip side of the "no excluded common causes" rule is that a causal graph, to do the work it must do, does not need to include everything we know about a substantive domain of interest. We may know quite a lot about the causes of economic inequality, for example. But we can safely omit any other factor from the graph as long *as it does not affect multiple nodes in the model.* Indeed, $\theta^I$ in Figure \@ref(fig:simpleDAG) already implicitly captures all factors that affect $I$, just as $\theta^D$ captures all factors *other than* mobilization that affect democratization. We may be aware of a vast range of forces shaping whether countries democratize, but choose to bracket them for the purposes of an examination of the role of economic inequality. This bracketing is permissible as long as none of these unspecified factors also act on  other nodes included in the model. 


**We can't read functional equations from a graph.** As should be clear, a DAG does not represent all features of a causal model. What it does record is which nodes enter into the structural equation for every other node: what can directly cause what. But the DAG contains no other information about the form of those causal relations. Thus, for instance, the DAG in Figure \@ref(fig:simpleDAG) tells us that $M$ is function of both $R$ and $E$, but it does not tell us whether that joint effect is additive ($R$ and $E$ separately increase mobilization) or interactive (the effect of each depends on the value of the other), or whether either effect is linear, concave, or something else. This lack of information about functional forms often puzzles those encountering causal graphs for the first time: surely it would be convenient to visually differentiate, say, additive from interactive effects. As one thinks about the variety of possible causal functions, however, it quickly becomes clear that there would be no simple visual way of capturing all possible functional relations. Moreover, causal graphs do not require functional statements to perform their main analytic purpose---a purpose to which we now turn.


### Conditional independence from DAGs

If we encode our prior knowledge using the grammar of a causal graph, we can put that knowledge to work for us in powerful ways. In particular, the rules of DAG-construction allow for an easy reading of the *conditional independencies* implied by our beliefs. (For another, somewhat more extended treatment of the ideas in this section, see @rohrer2018thinking.)

To begin thinking about conditional independence, it can be helpful to conceptualize dependencies between nodes as generating *flows of information*. Let us first consider a simple relationship of dependence. Returning to Figure \@ref(fig:simpleDAG), the arrow running from $I$ to $R$, implying a direct causal dependency, means that we expect $I$ and $R$ to be correlated. Put differently, observing the value of one of these nodes also gives us information about the value of the other. If we measured redistributive preferences, the graph implies that we would also be in a better position to infer the level of inequality, and vice versa. Likewise, $I$ and $M$ are also linked in a relationship of dependence: since inequality can affect mobilization (through $R$), knowing the the level of inequality would allow us to improve our estimate of the level of mobilization and vice versa.

In contrast, consider $I$ and $E$, which are in this graph indicated as being *independent* of one another. Learning the level of inequality, according to this graph, would give us no information whatsoever about the degree of ethnic homogeneity, and vice-versa.

Moreover, sometimes what we learn depends on *what we already know.* Suppose that we already knew the level of redistributive preferences. Would we then be in a position to learn about the level of inequality by observing the level of mobilization? According to this graph we would not: since the causal link---and, hence, flow of information between $I$ and $M$---runs through $R$, and we already know $R$, there is nothing left to be learned about $I$ by also observing $M$. Anything we could have learned about inequality by observing mobilization is already captured by the level of redistributive preferences, which we have already seen. In other words, if we were not to include $R$ in the causal model, then $I$ and $M$ would be dependent and informative about each other. When we do include $R$ in the causal graph, $I$ and $M$ are independent of one another and, hence, uninformative about each other. We can express this idea by saying that $I$ and $M$ are *conditionally independent given $R$*. 

We say that two nodes, $A$ and $C$, are "conditionally independent" given a set of nodes $\mathcal B$ if, once we have knowledge of the values in $\mathcal B$, knowledge of $A$ provides no information about $C$ and vice-versa. Taking $\mathcal B$ into account thus "breaks" any relationship that might exist unconditionally between $A$ and $C$. 

To take up another example, suppose that war is a cause of both military casualties and price inflation, as depicted in Figure \@ref(fig:warDAG). Casualties and inflation will then be (unconditionally) correlated with one another because of their shared cause. If we learn that there have been military casualties, this information will lead us to think it more likely that there is also war and, in turn, price inflation (and vice versa). However, assuming that war is their only common cause, we would say that military casualties and price inflation are *conditionally independent given war.* If we already know that there is war, then we can learn nothing further about the level of casualties (price inflation) by learning about price inflation (casualties). We can think of war, when observed, as blocking the flow of information between its two consequences; everything we would learn about inflation from casualties is already contained in the observation that there is war. Put differently, if we were just to look at cases where war is present (i.e., if we hold war constant), we should find no correlation between military casualties and price inflation; likewise, for cases in which war is absent.  


```{r warDAG, echo = FALSE, fig.width = 8, fig.height = 4,  fig.align="center", out.width = "60%", fig.cap = "This graph represents a simple causal model in which war ($W$) affects both military casualties ($C$) and price inflation ($P$)."}
hj_ggdag(x = c(0, 1, 1),
       y = c(1, 2, 0),
       names = c(
         "W", 
         "C",
         "P"
         ),
       arcs = cbind( c(1, 1),
                     c(2, 3)),
       title = "A Model of War's Effect on Casualties and Prices",
       padding = .2) 

```

Relations of conditional independence are central to the strategy of statistical control, or covariate adjustment, in correlation-based forms of causal inference, such as regression. In a regression framework, identifying the causal effect of an explanatory node, $X$, on a dependent node, $Y$, requires the assumption that $X$'s value is conditionally independent of $Y$'s potential outcomes (over values of $X$) given the model's covariates. To draw a causal inference from a regression coefficient, in other words, we have to believe that including the covariates in the model "breaks" any biasing correlation between the value of the causal node and its unit-level effect.

As we will explore, however, relations of conditional independence are also of more general interest in that they tell us, given a model, *when information about one feature of the world may be informative about another feature of the world, given what we already know*. By identifying the possibilities for learning, relations of conditional independence can thus guide research design. We discuss these research-design implications in Chapter \@ref(pt), but focus here on showing how relations of conditional independence operate on a DAG.


<!-- Some possibilities are excluded by the framework, however: for example, one cannot represent uncertainty regarding whether $A$ causes $B$ or $B$ causes $A$. -->

<!-- It would be nice to make a less general statement than the above as it sounds like none of this has any relevance if we think there's reciprocal causation in the causal system of interest; and that sounds like it excludes A LOT of problems political scientists are interested in. Pearl has a bit of discussion of the fact that one can compute the effect of interventions for models with cyclic features as well. So can we put this point more narrowly? Have been looking into this a bit but not yet sure exactly how to do that. -->




<!-- The above point is something I want to get more clarity around. Also, why we are defining roots as we are. -->

<!-- In Figure \@ref(fig:simpleDAG) we show a simple DAG that represents a situation in which $X$ is a parent of $M$, and $M$ is a parent of $Y$. In this example, the three nodes $\theta^X$, $\theta^M$, and $\theta^Y$ are all exogenous and thus elements of \(\mathcal{U}\). $X$, $M$, and $Y$ are endogenous and members of  \(\mathcal{V}\). If these three nodes were binary, then there would be eight possible realizations of outcomes, i.e., of \(\mathcal{V}\). In the underlying model,  $\theta^X$ is an ancestor of $X$, $M$, and $Y$ which are all descendants of $\theta^X$. The elements of $\mathcal{U}$ are all roots, though $X$ is also  a root as it has no parent in $\mathcal{V}$.   -->

 
<!-- In addition we will usually omit nodes from a graph only if they are single parents---this has the advantage of clarifying that all uncertainty is over the value of roots, and not over functional forms given roots; this is without loss of generality as parameters for functional equations can themselves be represented as roots.      -->

<!-- As a very simple example one might imagine that $A$ an $B$ are independently generated binary nodes; $C$ is an indicator for whether $A$ and $B$ have the same value. Then obviously if you know $C$, then knowing $A$ tells you everything about $B$. -->

To see more systematically how a DAG can reveal conditional independencies, it is useful spell out three elemental structures according to which which information can flow across a causal graph:


```{r pathsexp, echo = FALSE, fig.width = 6, fig.height = 4,  fig.align="center", out.width='70%', fig.cap = "Three elemental relations of conditional independence."}
p1 <- hj_ggdag(x = c(-1, 0, 1), y = c(1,1,1), names = c( "A", "B", "C" ),
       arcs = cbind( c(1, 2),
                     c(2, 3)),
       title = "(a) A path of arrows pointing in the same direction", padding = .2)
p2 <- hj_ggdag(x = c(-1, 0, 1), y = c(1,1,1), names = c( "A", "B", "C" ),
       arcs = cbind( c(2, 2),
                     c(1, 3)),
       title = "(b) A forked path", padding = .2) 
p3 <- hj_ggdag(x = c(-1, 0, 1), y = c(1,1,1), names = c( "A", "B", "C" ),
       arcs = cbind( c(1, 3),
                     c(2, 2)),
       title = "(c) An inverted fork (collision)", padding = .2)

plot_grid(plotlist = list(p1,p2,p3),ncol=1)

```


(1a) Information can flow unconditionally along a path of arrows pointing in the same direction. In Panel 1 of Figure \@ref(fig:pathsexp), information flows across all three nodes. Learning about any one will tell us something about the other two. 

(1b) Learning the value of a node along a path of arrows pointing in the same direction *blocks* flows of information across that node. Knowing the value of $B$ in Panel 1 renders $A$ no longer informative about $C$, and vice versa: anything that $A$ might tell us about $C$ is already captured by the information about $B$.

(2a) Information can flow unconditionally across the branches of any forked path. In Panel 2 learning only $A$ can provide information about $C$ and vice-versa.

(2b) Learning the value of the node at the forking point blocks *flows* of information across the branches of a forked path. In Panel 2, learning $A$ provides no information about $C$ if we already know the value of $B$.^[Readers may recognize this statement as the logic of adjusting for a confound that is a cause of both an explanatory node and a dependent node in order to achieve conditional independence.]

(3a) When two or more arrowheads collide, generating an inverted fork, there is no unconditional flow of information between the incoming sequences of arrows. In Panel 3, learning only $A$ provides no  information about $C$, and vice-versa. 

(3b) Collisions can be sites of *conditional* flows of information. In the jargon of causal graphs, $B$ in Panel 3 is a "collider" for $A$ and $C$.^[In the familial language of causal models, a collider is a child of two or more parents.] Although information does not flow unconditionally across colliding sequences, it does flow across them *conditional* on knowing the value of the collider node or any of its downstream consequences. In Panel 3, learning $A$ *does* provide new information about $C$, and vice-versa, *if* we also know the value of $B$ (or, in principle, the value of anything that $B$ causes). 

The last point is somewhat counter-intuitive and warrants further discussion. It is easy enough to see that, for two nodes that are correlated unconditionally, that correlation can be "broken" by controlling for a third node. In the case of collision, two nodes that are *not* correlated when taken by themselves *become* correlated when we condition on (i.e., learn the value of) a third node, the collider. The reason is in fact quite straightforward once one sees it: if an outcome is a joint function of two inputs, then if we know the outcome, information about one of the inputs can provide information about the other input. For example, if I know that you have brown eyes, then learning that your mother has blue eyes makes me more confident that your father has brown eyes. 

Looking back at our democratization DAG in Figure \@ref(fig:simpleDAG), $M$ is a collider for $R$ and $E$, its two inputs. Suppose that we again have the functional equation $M=RE$. Knowing about redistributive preferences alone provides no information whatsoever about ethnic homogeneity since the two are determined independently of one another. On the other hand, imagine that we already know that there was no mobilization. Now, if we observe that there *were* redistributive preferences, we can figure out the level of ethnic homogeneity: it must be 0. (And likewise in going from homogeneity to preferences.)

Using these basic principles, conditional independencies can be read off of any DAG. We do so by checking every path connecting two nodes of interest and ask whether, along those paths, the flow of information is open or blocked, given any other nodes whose values are already observed. Conditional independence is established when *all* paths are blocked given what we already know; otherwise, conditional independence is absent.

Following @pearl2000causality, we will sometimes refer to relations of conditional independence using the concept of *d-separation.* We say that variable set $\mathcal C$ $d-$separates variable set $\mathcal A$ from variable set $\mathcal B$ if $\mathcal A$ and $\mathcal B$ are conditionally independent given $\mathcal C$. We say that $\mathcal A$ and $\mathcal B$ are $d-$connected given $\mathcal C$ if $\mathcal A$ and $\mathcal B$ are *not* conditionally independent given $\mathcal C$. 


### Simplifying models

It is very easy to write down a model that is too complex to use effectively. In such cases we often seek simpler models that are consistent with models we have in mind but contain fewer nodes or more limited variation. In general this is possible but caution has to be taken to ensure that simplified models are indeed consistent with the original model. 

Fortunately the mapping between graphs and relations of conditional independence give guidance for determining when and how it is possible to simplify models.  We focus discussion on simplifications that involve node elimination and conditioning on nodes.

#### Eliminating nodes

If we want to eliminate a node the key rule is that the new model (and graph) must take into account: 

(a) all *dependencies* among remaining nodes and 
(b) all *variation* generated by the eliminated node. 

We can work out what this means, separately, for eliminating *endogenous* nodes and for eliminating *exogenous* nodes.

*Eliminating endogenous nodes*

Eliminating an endogenous node means removing a node with parents (direct causes) represented on the graph. If the node also has one or more children, then the node captures a dependency: it links its parents to its children. When we eliminate this node, preserving these dependencies requires that all of the eliminated node's parents adopt---become parents of---all of the eliminated node's children. Thus, for instance if we had a model in which $A \rightarrow M \leftarrow B$ and $M \rightarrow Y$, if we were to eliminate $M$, $M$'s parents ($A$ and $B$) would need to adopt $M$'s child, $Y$. 

<!-- As for $\theta^M$, it too must now point directly into $Y$---though we can use a bit of shorthand to make this happen. Recall that $\theta^M$ represents the part of $M$ that is randomly determined. Rather than drawing two separate disturbance ($\theta$) terms pointing into $Y$, however, we more simply represent the combined disturbance term as $\theta^Y_{\text{higher}}$, with the ''higher'' signaling the aggregation of roots. (This is, of course, simply reversing the disaggregation that we undertook earlier to move from the higher- to the lower-level model.) -->

More intuitively, when we simplify away a mediator, we need to make sure that we preserve the causal relationships being mediated---both those among substantive variables and any random shocks at the mediating causal steps.^[Eliminating endogenous nodes may also operate via "encapsulated conditional probability distributions" [@koller2009probabilistic] wherein a system of nodes, $\{Z_i\}$  is represented by a single node, $Z$,  that takes the parents of  $\{Z_i\}$ not in $\{Z_i\}$ as parents to $Z$ and issues the children of $(Z_i)$ that are not  in $(Z_i)$ as children. However, this is not a fundamental alteration of the graph.]

*Eliminating exogenous nodes*

What about eliminating exogenous nodes---nodes with no parents? For the most part, exogenous nodes cannot be eliminated, but must either be replaced by or incorporated into $U$ (or $\theta$) terms. The reason is that we need preserve any dependencies or variation generated by the exogenous node.  Figure \@ref(fig:elimrules) walks through four different situations in which we might want to simplify away the exogenous node, $X$. (Here we use the more generic $U$ notation, though the same principles apply if these are type-receptacles $\theta$.)

```{r elimrules, echo = FALSE, fig.width = 10, fig.height = 15,  fig.align="center", out.width='.9\\textwidth', fig.cap = "Basic principles for eliminating exogenous nodes."}
p1 <- hj_ggdag(
  x = c(1,2,2),
  y = c(2,3,1),
       names = c(
         expression(paste(X)),
         expression(paste(W)),
         expression(paste("Y"))
         ),  
       arcs = cbind( c(1, 1),
                     c(3, 2)),
       padding = .2
)

p2 <- hj_ggdag(x = c(1,2,2),
       y = c(2,3,1),
       names = c(
         expression(paste(theta^{WY})),
         expression(paste(W)),
         expression(paste(Y))
       ),
       arcs = cbind( c(1, 1),
                     c(3, 2)),
       padding = .2
)


p3 <- hj_ggdag(x = c(1,1,2),
       y = c(3,1,2),
       names = c(
         expression(paste(X)),
         expression(paste(W)),
         expression(paste("Y"))
       ),
       arcs = cbind( c(1, 2),
                     c(3, 3)),
       padding = .2
)


p4 <- hj_ggdag(x = c(1,1,2),
       y = c(3,1,2),
       names = c(
         expression(paste(theta^Y)),
         expression(paste(W)),
         expression(paste("Y"))
       ),
       arcs = cbind( c(1, 2),
                     c(3, 3)),
       padding = .2
)

p5 <- hj_ggdag(x = c(1,1,2),
       y = c(3,1,2),
       names = c(
         expression(paste(X)),
         expression(paste(theta^Y)),
         expression(paste("Y"))
       ),
       arcs = cbind( c(1, 2),
                     c(3, 3)),
       padding = .2
)

p6 <- hj_ggdag(x = c(1,2),
       y = c(1,1),
       names = c(
         expression(paste(theta^Y[mod])),
         expression(paste("Y"))
       ),
       arcs = cbind( c(1),
                     c(2)),
       padding = .2
)


p7 <- hj_ggdag(x = c(1,3,2),
       y = c(1,1,1),
       names = c(
         expression(paste(X)),
         expression(paste(W)),
         expression(paste("Y"))
       ),
       arcs = cbind( c(1, 3),
                     c(3, 2)),
       padding = .2
)

p8 <- hj_ggdag(x = c(3,2),
       y = c(1,1),
       names = c(
         expression(paste(W)),
         expression(paste("Y"))
       ),
       arcs = cbind( c(2),
                     c(1)),
       padding = .2
)

titles =c("(a1) Original model: X has two children",
          "(a2) Simplification: Dependency between \nW and Y must be preserved",
          "(b1) Original model: X has a substantive spouse",
          "(b2) Simplification: X must be replaced with U term",
          "(c1) Original model: X has a random spouse",
          "(c2) Simplification: X absorbed into error term",
          "(d1) Original model: X has one child and no spouse",
          "(d2) Simplification: X can be safely removed")

plot_grid(plotlist = list(p1,p2,p3,p4,p5,p6,p7,p8),
          labels = titles,
          ncol=2,label_size = 10,hjust=0, scale = 0.85)


```



*  *Multiple children.* In (a1), we start with a model in which $X$ has two children, thus generating a dependency between $W$ and $Y$. If we eliminate $X$, we must preserve this dependency. We can do so, as pictured in (a2), by replacing $X$ with a $\theta$ term that  points into both $W$ and $Y$. By convention, we could, alternatively, convey the same information with a dashed, undirected line between $W$ and $Y$. Though we are no longer specifying what it is that connects $W$ and $Y$, the correlation itself is retained.
*  *Substantive spouse.* In (b1), $X$ has a spouse that is substantively specified, $W$. If we eliminate $X$, we have to preserve the fact that $Y$ is not fully determined by $W$; *something* else also generates variation in $Y$. We thus need to replace $X$ with a $\theta$ term, $\theta^Y$, to capture the variation in $Y$ that is not accounted for by $W$.
* *$\theta$-term spouse.* In (c1), $X$ has a spouse that is *not* substantively specified, $U^{Y}$. Eliminating $X$ requires, again, capturing the variance that it generates as a random input. As we already have a $\theta$ term pointing only into $Y$, we can substitute in $\theta^{Y}_\text{mod}$, which represents both $U^{Y}$ and the variance generated by $X$.^[This aggregation cannot occur if $\theta^{Y}$ also has another child, $W$, that is not a child of $X$ since then we would be representing $Y$'s and $W$'s random components as identical, which they are not in the original graph.]
*  *One child, no spouse.* In (d1), $X$ has only one child and no spouse. Here we can safely eliminate $X$ with no loss of information. It is always understood that every exogenous node has some cause, and there is no loss of information in simply eliminating a node's causes if those causes are exogenous and do not affect  other endogenous nodes in the model. In (d2) we are simply not specifying $Y$'s cause, but we have not lost any dependencies or sources of variance that had been expressed in (d1).


One interesting effect of eliminating a substantive exogenous node can be to render seemingly deterministic relations effectively probabilistic. In moving from (b1) to (b2), we have taken a component of $Y$ that was determined by $X$ and converting it into a random disturbance. Just as we can explain a more probabilistic claim with a less probabilistic theory, we can derive claims from simplified models with greater probabilism from theories with greater determinism.



```{r lowercomplexdem, echo = FALSE, fig.width = 10, fig.height = 6,  fig.align="center", out.width='.5\\textwidth', fig.cap = "A model  from which multiple simpler models can be derived."}

x = c(0,0, 1, 1, 2)
y = c(2,0, 2, 0, 1)

names = c(
         "I: Inequality",
         "E: Ethnic\nhomogeneity",
         "R: Redistributive\npreferences",
         "M: Mobilization",
         "D: Democratization"
         )

set.seed(1)
hj_ggdag(x =  x,
         y = y,
         names = c(names),
         arcs = cbind( c(1,2,2,3,4,3),
                       c(3,3,4,5,5,4)),
         title = "A Model of Inequality and Ethnic Homogeneity's Effects on Democratization",
         padding = -0.1,
         labels = T,direction='x')

# hj_dag(x =  c(x, 0, 0),
#        y = c(y, 0.25, 1.75),
#        names = c(names),
#        arcs = cbind( c(1,2,2, 3, 4, 3),
#                      c(3,3,4, 5, 5, 4)),
#        title = "Free Press and Government Survival",
#        add_functions = 0,
#        contraction = .15,
#        add_functions_text = "Structural Equations: Y = CR, R = CX, C = 1-XS",
#        padding = .2)

# text(c(0,0), c(.25, 1.75), c(expression(paste(theta^X[1])), expression(paste(theta^S[1]))))

```

<!-- ```{r lowercomplexdem, echo = FALSE, fig.width = 10, fig.height = 6,  fig.align="center", out.width='.5\\textwidth', fig.cap = "A lower-level model  from which multiple higher level models can be derived."} -->


<!-- par(mfrow = c(1,1)) -->
<!-- par(mar=c(1,1,3,1)) -->
<!-- hj_dag(x = c(1,1,2,2,3), -->
<!--        y = c(2,1,2,1,1.5), -->
<!--        names = c( -->
<!--          expression(paste("I: Inequality")), -->
<!--          expression(paste("E: Ethnic\nhomogeneity")), -->
<!--          expression(paste("M: Mobilization")), -->
<!--          expression(paste("R: Redistributive\npreferences")), -->
<!--          expression(paste("D: Democratization")) -->
<!--          ),   -->
<!--        arcs = cbind( c(1, 3, 2, 2, 3, 4), -->
<!--                      c(3, 4, 4, 3, 5, 5)), -->
<!--        add_functions = 0,  -->
<!--        contraction = .16,  -->
<!--        padding = .2 -->
<!-- ) -->
<!-- ``` -->

(ref:figref) \@ref(fig:lowercomplexdem)


```{r runningsubs, echo = FALSE, fig.width = 12, fig.height = 13.5, fig.cap = "Simplifications of the model of Figure (ref:figref). Nodes that are eliminated are marked in grey; circles denote exogenous nodes that are replaced in subgraphs by unidentified variables. (A circled node pointing into two other nodes could equivalently be indicated as an undirected edge connecting the two.)"}

if(do_diagnosis){
x = c(0,0, 1, 1, 2)
y = c(2,0, 2, 0, 1)
# names = c("S", "X", "C", "R", "Y")
names = c("I", "E", "R", "M", "D")

M <- matrix(0, 5, 5)
M[1, c(3)] <-1
M[2, c(3,4)] <-1
M[3, c(4,5)] <-1
M[4, 5] <-1

colnames(M)<-names
rownames(M)<-names

matrix_remove <- function(M, remove = NULL){
  M2 <- M
if(!is.null(remove)) {
  for(j in remove) {M2 <- (M2 + outer(M[, j], M[j,]))}  # connect parents to children
  for(j in remove) {if(sum(M2[,j]>0)) {M2[j,] <- 0; M2[,j] <- 0}}  # disconnect non-roots (roots may have to be renamed)
}
M2[M2>1] <- 1
M2
}

removes <- perm_bb(c(2,2,2,2,2))[-1,]
removes <- removes[rowSums(removes)<=3,]
removes <- removes==1
removes <- removes[order(rowSums(removes) - (1:nrow(removes))/100),] # for each of the graphs, which arrows to remove?

plist <- list()

for(j in 1:nrow(removes)){
  GO <- (1:5)[removes[j,]]
  arcs <- which(matrix_remove(M, GO)==1, arr.ind = TRUE)
  ind1 <- rownames(M)[arcs[,1]]
  ind2 <- colnames(M)[arcs[,2]]

  from <- names[!names %in% c(ind1,ind2)]
  to <- 1:length(from)
  if (length(from)!=0){
    for (k in 1:length(from)){
      to[k] <- sample(names[!names %in% from[k]],1)
}
    arcs <- rbind(arcs,
                  cbind(match(from,names),
                        match(to,names)))
    obscure <- paste(from,to,sep='->')
    } else {
      obscure <- NULL
      }
  nodecol = rep('white',5)
  textcol = rep('black',5)
  textcol[GO] <- 'grey'
  for(i in 1:ncol(M)) if((sum(M[,i])==0) & (i%in% GO)) nodecol[i] <- 'grey'
  
  plist[j] <- hj_ggdag(
       x = x,
       y = y,
       names = names,
       arcs = arcs,
       add_points = FALSE,
       obscure = obscure,
       shape = 21,
       textcol = textcol,
       nodecol = nodecol,
       textsize = 4,
       nodesize = 12
       ) %>% list()
}

cowplot::plot_grid(plotlist=plist,ncol=5,labels=1:nrow(removes),
                   label_size = 10) %>%
  write_rds("saved/02_simplifications.rds")
}

read_rds("saved/02_simplifications.rds")
```


We can apply these principles to a model of any complexity. We illustrate a wider range of simplifications by starting with Figure \@ref(fig:lowercomplexdem). In Figure \@ref(fig:runningsubs), we show all permissible reductions of the more elaborate model. We can think of these reductions as the full set of simpler claims (involving at least two nodes) that can be derived from the original model. In each subgraph, 

* we mark eliminated nodes in grey; 
* those nodes that are circled must be replaced with $\theta$ terms; and
* arrows represent the causal dependencies that must be preserved. 

Note, for instance, that neither $I$ (because it has a spouse) nor $E$ (because it has multiple children) can be simply eliminated; each must be replaced with a $\theta$ term. Also, the simplified graph with nodes missing can contain arrows that do not appear at all in the graph: eliminating $R$, for instance, forces an arrow running from $E$ to $M$ (though that is there already) and another running from $E$ to $D$, as $E$ must adopt $R$'s children. The simplest elimination is of $Y$ itself since it does not encode any dependencies between other variables.
<!-- ^[Put differently, and in language that we introduce below, colliding arrowheads do not represent a path in DAG analysis.]  -->



#### Conditioning on nodes

Another way to simpify a model is to condition on the value of a node. When we condition on a node, we are restricting the model in scope to situations in which that node's value is held constant. Doing so allows us to eliminate the node as well as all arrows pointing into it or out of it. Consider three different situations in which we might condition on a node:


* *Exogenous, with multiple children.* In simplifying (a1) in Figure \@ref(fig:elimrules), we need to be sure we retain any dependence that $X$ generates between $W$ and $Y$. However, recalling the rules of conditional independence on a graph (see Chapter \@ref(models)), we know that $W$ and $Y$ are *independent* conditional on $X$. Put differently, if we restrict the analysis to contexts in which $X$ takes on a constant value, the model implies that $Y$ and $W$ will be uncorrelated across cases. As fixing $X$'s value breaks the dependence between $Y$ and $W$, we can drop $X$ (and the arrows pointing out of it) without having to represent that dependence. 
*  *Exogenous, with spouse.* In simplifying (b1) or (c1) in Figure \@ref(fig:elimrules), we need to account for the variation generated by $X$. If we fix $X$'s value, however, then we eliminate this variation by assumption and do not need to continue to represent it (or the arrow pointing out of it) on the graph.
*  *Endogenous.* When we condition on an endogenous node, we can eliminate the node as well the arrows pointing into and out of it. We, again, leverage relations of conditional independence here. If we start with model $X \rightarrow M \rightarrow Y$ and we condition on the mediator, $M$, we sever the link between $Y$ and $X$, rendering them conditionally independent of one another. We can thus remove $M$, the arrow from $X$ to $M$, and the arrow from $M$ to $Y$. In the new model, with $M$ fixed, $Y$ will be entirely determined by the random disturbance $\theta^{Y}$.

### Retaining probabilistic relations

We have highlighted the graphical implications of node elimination or node conditioning but importantly the distribution over $\theta$ also needs to be preserved faithfully in a move to a simpler model.


In sum, we can work with models that are simpler than our causal beliefs: we may believe a model to be true, but we can derive from it a sparser set of claims. There may be intervening causal steps or features of context that we believe matter, but that are not of  interest for a particular line of inquiry. While these can be removed, we nonetheless have to make sure that their *implications* for the relations remaining in the model are not lost. Understanding the rules of reduction allow us to undertake an important task: checking which simpler claims are and are not consistent with our full belief set.


## Conclusion {#conc2}

In this chapter, we have shown how we can inscribe causal beliefs, rooted in the potential outcomes framework, into a causal model. In doing so, we have now set out the foundations of the book's analytic framework. Causal models are both the starting point for analysis in this framework and the object about which we seek to learn. Before moving on to build on this foundation, we aim in the next chapter to offer further guidance by example on the construction of causal models, by illustrating how a set of substantive social scientific arguments from the literature can be represented in causal model form. 


<!-- ### A simple running example -->

<!-- We can illustrate these core ideas with a simple example of a model of government corruption and survival, one that we will return to in future chapters. -->

<!-- We begin with two binary features of context. Consider, first, that a country may or may not have a free press ($X$). Second, the country's government may or may not be sensitive to public opinion ($S$).^[Government sensitivity here can be thought of as government sophistication (does it take the actions of others into account when making decisions?) or as a matter of preferences (does it have a dominant strategy to engage in corruption?).] Let us then stipulate what follows from these conditions. The government will engage in corruption ($C=1$) unless it is sensitive to public opinion and there is a free press. Moreover, if and only if there is both government corruption and a free press, the press will report on the corruption ($R=1$). Finally, the government will be removed from office ($Y=1$) if it has acted corruptly and this gets reported in the press; otherwise, the government remains in office. -->

<!-- As a set of equations, this simple causal model may be written as follows: -->

<!-- $\begin{array}{ll} -->
<!-- C = 1-X\times S &  \mbox{Whether the government is corrupt}\\ -->
<!-- R = C\times X &  \mbox{Whether the press reports on corruption}\\ -->
<!-- Y = C\times R & \mbox{Whether the government is removed from office} -->
<!-- \end{array}$ -->

<!-- One thing that these equations make clear is that the nodes in our model function in various places as causal-type nodes for one another. For instance, we can see from equation for $C$ that the causal effect of a free press ($X$) on corruption ($C$) depends on whether the government is sensitive to public opinion ($S$): $S$ determines $C$'s response to $X$ (as does $X$ for $S$'s effect on $C$). A similar relationship holds for $C$ and $X$ in their effect on $R$ and for $C$ and $R$ in their effect on $Y$. As we will see below, the model also implies more complex causal-type relationships. We can, further, substitute through the causal processes to write down the functional equation for the outcome in terms of the two initial causal nodes: $Y=(1-S)X$.^[In Boolean notation (but preserving a structural equation interpretation), where $Y$ stands for the occurrence of government removal, $Y= \neg S \land X$; and the function for the outcome "government retained" can be written  $\neg Y = (S\land X) \lor (S\land\neg X) \lor (\neg S \land \neg X)$ or, equivalently, $\neg Y = S + \neg S \neg X$.] -->

<!-- Let us, further, allow the two primary causal nodes---the existence of a free press and the existence of a sensitive government---to vary probabilistically. In particular, we represent the probability of a free press with the population parameter $\lambda^X_1$ and the probability of a sensitive government with the parameter $\lambda^S_1$.  -->

<!-- Note that in this model, only the most "senior" specified nodes, $X$ and $S$, have a stochastic component (i.e., $\lambda^X_1$ and $\lambda^S_1$ lie between 0 and 1). All other endogenous nodes are deterministic functions of other specified nodes (put differently: each node has only a single nodal type). -->


<!-- ```{r running, echo = FALSE, fig.width = 11, fig.height = 9, fig.align="center", out.width = "70%", fig.cap = "The figure shows a simple causal model. $S$ and $X$ are stochastic, other nodes fully determined by their parents, as shown in bottom right panel.", fig.align="center", warning = FALSE, eval = FALSE} -->


<!-- x = c(0,0, 1, 1, 2) -->
<!-- y = c(2,0, 2, 0, 1) -->

<!-- names = c("S:\nSensitive\ngovernment\n\n", "\nX:\nFree Press", "C:\n Corruption", "R:\n Media report", "Y:\nGovernment\nreplaced") -->

<!-- hj_dag(x =  x, -->
<!--        y = y, -->
<!--        names = c(names), -->
<!--        arcs = cbind( c(1,2,2, 3, 4, 3), -->
<!--                      c(3,3,4, 5, 5, 4)), -->
<!--        title = "Free Press and Government Survival", -->
<!--        add_functions = 0, -->
<!--        contraction = .15, -->
<!--        add_functions_text = "Structural Equations: Y = CR, R = CX, C = 1-XS", -->
<!--        padding = .2) -->

<!-- # hj_dag(x =  c(x, 0, 0), -->
<!-- #        y = c(y, 0.25, 1.75), -->
<!-- #        names = c(names), -->
<!-- #        arcs = cbind( c(1,2,2, 3, 4, 3), -->
<!-- #                      c(3,3,4, 5, 5, 4)), -->
<!-- #        title = "Free Press and Government Survival", -->
<!-- #        add_functions = 0, -->
<!-- #        contraction = .15, -->
<!-- #        add_functions_text = "Structural Equations: Y = CR, R = CX, C = 1-XS", -->
<!-- #        padding = .2) -->

<!-- # text(c(0,0), c(.25, 1.75), c(expression(paste(theta^X[1])), expression(paste(theta^S[1])))) -->

<!-- ``` -->


<!-- The corresponding causal diagram for this model is shown in Figure \@ref(fig:running). -->

<!-- In later chapters we will develop this model and use it to illustrate different estimands and different strategies for case level inference. -->


\newpage

## Chapter Appendix

### Steps for constructing causal models

:::: {.headerbox data-latex=""}
::: {.center data-latex=""}
**Steps for constructing causal models**
:::

1. Identify a set of variables in a domain of interest. These become the nodes of the model.

  * Specify the range of each node: is it continuous or discrete?
  * Each node should have an associated $\theta$ term representing unspecified other influences (not necessarily graphed)

2. Draw a causal graph (DAG) representing beliefs about causal dependencies among these nodes

  * Include arrows for direct effects only
  * Arrows indicate *possible* causal effects
  * The absence of an arrow between two nodes indicates a belief of *no* direct causal relationship between them
  * Ensure that the graph captures all correlations among nodes. This means that either (a) any common cause of two or more nodes is included on the graph (with implications for Step 1) or (b) correlated nodes are connected with a dashed or curved, undirected edge.
  
3. Write down one causal function for each endogenous node

  * Each node's function must include all nodes directly pointing into it on the graph as well as the $\theta$ terms
  * Functions may express arbitrary amounts of uncertainty about causal relations
  
4. State probabilistic beliefs about the distributions of the $\theta$s.

  * How common or likely to do we think different values of the exogenous nodes are?
  * Are they independently distributed? If in step 2 you drew an undirected edge between nodes then you believe that the connected nodes are not independently distributed.
::::


### Model construction in code

Our `CausalQueries` package provides a set of functions to implement all of these steps concisely for *binary* models -- models in which all nodes are dichotomous.

```{r modelconstr, message = FALSE, eval = FALSE}

# Steps 1 and 2 
# We define a model with three binary nodes and 
# specified edges between them:
model <- make_model("X -> M -> Y")

# Functional forms are unrestricted. Restrictions can 
# be added. Here we impose monotonicity at each step 
# by removing one type for M and one for Y
model <- set_restrictions(model, labels = list(M = "10", Y="10"))

# Step 4
# Set priors over the distribution of (remaining) causal types.
# Here we set "jeffreys priors"
model <- set_priors(model, distribution = "jeffreys")

# We now have a model defined as an R object. 
# Later we will update  and query this model

```

These steps are enough to fully describe a binary causal model. Later in this book we will see how we can ask questions of a model like this but also how to use data to train it. 

<!-- If we want to know whether two nodes, $A$ and $C$, are conditionally independent given some set of nodes, $\mathcal B$, we need to find out whether any paths between $A$ and $C$ are "active" given $\mathcal B$. Put slightly differently, the question is whether nodes in $\mathcal{B}$ block information flows from $A$ to $C$---or rather allow, or even create, such flows. This can be assessed as follows. For each possible path between $A$ and $C$, we check whether there are three connected nodes $X, Y, Z$^[Where $X$ and/or $Y$ may be $A$ and/or $B$.] on that path that either: -->

<!-- (a) form a "chain" $X\rightarrow Y \rightarrow Z$ (going either direction) or "fork" $X\leftarrow Y\rightarrow Z$, with $Y \subset \mathcal{B}$,^[For instance, under this criterion, if we have $A\rightarrow W\rightarrow Y \rightarrow B$, and $Y$ is an element of the set $\mathcal{B}$, then this path is not active given $\mathcal{B}$. Similarly, if we have $A \leftarrow X\leftarrow Y\rightarrow Z \rightarrow B$, and $Y$ is in $\mathcal{B}$, then the path is not active given $\mathcal{B}$.] or  -->

<!-- (b) form an "inverted fork" $X \rightarrow Y \leftarrow Z$, with neither $Y$ nor its descendants in $\mathcal C$.^[For instance, if we have $A \rightarrow W \rightarrow Y \leftarrow Z \leftarrow B$, and $Y$ is *not* in $\mathcal{C}$, then the path is not active given $\mathcal{C}$. In other words, "inverted fork" paths are not unconditionally active. Knowing only $A$, in this setup, tells us nothing about $B$, and vice-versa.]  -->

<!-- If either of these conditions holds, then the path is blocked (not active) given $\mathcal C$. In the first case, any possible information flows along the path are *blocked* by a node in $\mathcal C$. In the second case, *no* node in $\mathcal C$ is *creating* an information flow that would not otherwise be present. If there are no active paths, then $A$ and $B$ are conditionally independent given $\mathcal C$. In the graph-analytic language of Pearl and others, $A$ and $B$ are said to be "$d$-separated" by $\mathcal{C}$.^[There are multiple techniques for establishing $d-$separation. Pearl's guide "$d-$separation without tears" appears in an appendix to @pearl2009causality.]   -->

<!-- Thus, in Figure \@ref(fig:simpleDAG), we can readily see that $X$ and $Y$ are conditionally independent given $R$: $X$ and $Y$ are *d-*separated by $R$.  Conversely, $R$ and $\theta^Y$ are unconditionally independent. However, conditioning on $Y$ $d-$*connects* $R$ and $\theta^Y$, generating a dependency between them. If and only if we know the outcome, $Y$, then learning $R$ yields information about $\theta^Y$. To foreshadow a point that we develop further later in the book, this analysis reveals how we can, and cannot, learn empirically about elements of our models. For instance, if $\theta^Y$ is a node of interest but not directly observable, then information on $R$ is unhelpful if we have not observed the outcome, $Y$, but *is* informative if we have. -->


<!-- ###Interventions in causal graphs -->

<!-- A second advantage of causal graphs is that they provide a useful structure for thinking through causal effects. In a causal-model framework, we think of a node's causal effect as being the effect of an imagined *intervention*: a manipulation of some node, $X$, that provides $X$ with a value that is *not* determined by its parents.  In an intervention, it is as if the causal function for $X$ is replaced by the function $X=x$, with $x$ being the constant value to which we have set $X$. If we take $X$ to be binary, to keep things simple, then the causal effect of $X$ on $Y$ is the difference between the value $Y$ would take on under the intervention $X=0$ (which can be written as $do (X)=0$) and the value $Y$ would take on under the intervention $X=1$ ($do(X)=1$).  -->

<!-- As this manipulative account of causation makes clear, analyzing the effect of $X$ requires us we to set aside the "natural causes" of $X$ itself in a particular way. We can readily see how this works graphically by considering a modified version of our free press/government survival model as displayed in Figure \ref{fig:DAGdirect}. As compared with Figure \ref {fig:simpleDAG}, this DAG represents a model in which $X$ has effects on $Y$ both indirectly through $R$ and directly. We might imagine, for instance, that part of the effect of a free press ($X$) on government turnover ($Y$) runs via media reports of official corruption ($R$) while part of the effect runs through a deterrent effect of a free press that reduces graft and thus leaves greater public resources for investment in public goods (neither of which mediating nodes are represented on the graph). -->

<!-- Suppose that we want to estimate the effect of an increase in media reports of corruption, $R$, on government survival, $Y$. We thus want to know the difference between the value that $Y$ would take on if the number of media reports were set at a  high level and the value $Y$ would take on if media reports were set to low. As should be clear from the graph, we cannot estimate this difference by comparing the observed value of $Y$ when media report levels are high to the observed value of $Y$ when media report levels are low. The reason is that $R$ has an ancestor, $X$, that affects $Y$ via a path that does not run through $R$. Thus, the value that $Y$ takes on when $R$ is observed to be high (or low) will be determined not just by $R$ but also by $X$, which is itself systematically correlated with $R$ by being its ancestor. The difference between $Y$'s values at two values of $R$ will thus itself be a combination of $R$'s effect and of $X$'s effect. -->

<!-- Instead, the query, "What is the effect of $R$ on $Y$?", must be conceived of in a way that separates out the effect of $X$ on $R$. We can represent the empirical conditions that we would need to estimate this effect in Figure \ref{fig:DAGdirectmut}. We have "mutilated" the original DAG by removing all arrows pointing into our causal node, $R$ (here, simply the arrow running from $X$ to $R$). This graph represents a world in which $R$ has been manipulated, rather than naturally caused. (Equivalently, it is a world in which $R$ is purely exogenous.) And what the mutilated graph tells us is that, in an empirical situation in which the  dependency of $R$ on $X$ could be removed, $R$'s causal effect on $Y$ *could* be estimated by comparing $Y$'s values at high and low levels of $R$.^[In the formulation used in  @pearl2009causality, an intervention involving an endogenous node $V_i$ can be written as $do(V_i)=v_i'$, or for notational simplicity $\hat{v}_i'$ (meaning $V_i$ is forced to take the particular value $v_i'$). The resulting distribution can be written as a modified version of Equation \@ref(eq:markov): -->

<!-- \begin{equation}  -->
<!-- P(v_1,v_2,\dots v_n|\hat{v}_i) = \prod_{-i}P(v_j|pa_j)\mathbb{1}(v_i = v_i')(\#eq:eqdo) -->
<!-- \end{equation} -->

<!-- where $-i$ indicates that the product is formed over all nodes $V_j$ other than $V_i$, and the indicator function ensures that probability mass is only placed on vectors (or worlds) with $v_i = v_i'$. This new distribution has a graphical interpretation, representing the probability distribution over a graph in which all arrows into $V_i$ are removed.  The difference between Equation \@ref(eq:markov) and \@ref(eq:eqdo) is the difference between an *observed* probability distribution and the effect of an *intervention*. The key differences is that, when we intervene to manipulate a node, we break the link between the node and its "natural" causes.] -->

<!-- It is a separate question how such an empirical situation might be generated. But it is not hard to see from Figure \ref{fig:DAGdirect} that the  dependency of $R$ on $X$ could be removed either via (i) directly manipulating $R$ in a manner orthogonal to $X$ or (ii) controlling for $X$, since any variation in $R$ conditional on $X$ would itself be independent of $X$. Equivalently, using graph-analytic logic, we can also see that conditioning on $X$ blocks the path between $R$ and $Y$ that generates the confound (called a "backdoor"), thus expunging the confounding effect from the observed correlation between $R$ and $Y$. -->


<!-- ```{r, echo = FALSE, fig.width = 5, fig.height = 3,  fig.align="center", out.width='.5\\textwidth', fig.cap = "\\label{fig:DAGdirect} As compared with Figure \ref {fig:simpleDAG}, this DAG represents a model in which $X$ has effects on $Y$ both indirectly through $R$ and directly. We might imagine, for instance, that part of the effect of a free press ($X$) on government turnover ($Y$) runs via media reports of official corruption ($R$) while part of the effect runs through a deterrent effect of a free press that reduces graft and thus leaves greater public resources for investment in public goods (neither of which mediating nodes are represented on the graph)."} -->
<!-- par(mar=c(1,1,3,1)) -->
<!-- hj_dag(x = c(0, 1, 2), -->
<!--        y = c(2, 3, 2), -->
<!--        names = c("X", "R", "Y"), -->
<!--        arcs = cbind( c(1, 2, 1), -->
<!--                      c(2, 3, 3)), -->
<!--        title = "A DAG with Indirect and Direct Effects", -->
<!--        padding = .4, contraction = .15)  -->

<!-- ``` -->


<!-- ```{r, echo = FALSE, fig.width = 5, fig.height = 3,  fig.align="center", out.width='.5\\textwidth', fig.cap = "\\label{fig:DAGdirectmut} This DAG represents a 'mutilated' version of the previous graph in which the causes of $R$ have been removed, and thus captures the empirical relations that would be required to hold to estimate the effect of $R$ on $Y$."} -->
<!-- par(mar=c(1,1,3,1)) -->
<!-- hj_dag(x = c(0, 1, 2), -->
<!--        y = c(2, 3, 2), -->
<!--        names = c("X", "R", "Y"), -->
<!--        arcs = cbind( c(2, 1), -->
<!--                      c(3, 3)), -->
<!--        title = "A 'Mutilated' DAG for Estimating the Effect of R on Y", -->
<!--        padding = .4, contraction = .15)  -->

<!-- ``` -->





<!-- Suppose, for instance, that democracy ($D$) causes higher levels of private investment ($I$) and greater public-goods provision ($P$), and that both of these cause faster economic growth ($G$). If we want to know the joint distribution of public-goods provision and democracy, we would use Equation \ref{eqmarkov}, conditioning on the values of the parents of these two nodes. If, however, we want to know what level of growth would be produced by an increase public-goods provision---a causal question---we have to ask  -->



<!-- ```{r, echo = FALSE, fig.width = 11, fig.height = 11.5, fig.align="center", out.width='\\textwidth', fig.cap = "\\label{fig:intervention} The main panel shows a simple causal model. $S$ and $X$ are stochastic, other nodes determined by their parents, as shown in bottom right panel. Other panels show four possible histories that can arise depending on values taken by $S$ and $X$, along with causal relations in each case. The equations for $S$ and $X$ are written with indicator nodes, which take a value of 1 whenever the $u$ value is less than the $\\pi$ value.", fig.align="center", warning = FALSE} -->

<!-- hj_dag(x = c(1, 2, 2, 3), -->
<!--        y = c(1, 2, 0, 1), -->
<!--        names = c("D","I", "P", "G"), -->
<!--        arcs = cbind( c(1, 1, 3, 2), -->
<!--                      c(2, 3, 4, 4)), -->
<!--        add_functions = 0, -->
<!--        contraction = .2 -->
<!--        ) -->
<!-- title("A causal model") -->
<!-- ``` -->



<!-- Not completely following the logic this next paragraph or sure if it is helpful here -->

<!-- So far, although not completely general, the focus on causal DAGs is consistent with most approaches used in qualitative work on process tracing, in qualitative case analysis, and in econometric approaches. Some of these approaches commonly assume simple functional forms but these impositions are not implied by the general approach. For example econometric models often impose linear assumptions---for example in work on linear structural equations. Qualitative case analysis often assume all units are binary and that outcomes are deterministic. Under some representations the latter assumption implies conditional independencies that cannot be read from the graph, and thus violate stability conditions commonly assumed of the probability distributions that graphs are meant to represent (though it is still always that case that one can tell from the graph whenever two sets of nodes are not conditionally independent given some other set). In our running example described below we give an example of such deterministic relations. -->


### Rules for moving between levels
:::: {.headerbox data-latex=""}
::: {.center data-latex=""}
**Rules for moving between levels**
:::

*Moving down levels*: 

All (conditional) independencies represented in a higher-level model must be preserved in the lower-level model. 

When we disaggregate or add nodes to a model, new conditional independencies can be generated. But any variables that are independent or conditionally independent (given a third variable) in the higher-level model must also be independent or conditionally independent in the lower-level model.

*Moving up levels*: 

We can move up levels by eliminating an exogenous node, eliminating an endogenous node, or conditioning on a node. When we eliminate a node from a model, we must preserve any variation and dependencies that it generates: 

1. When eliminating an endogenous node, that nodes parents adopt (become direct causes of) that nodes children. 
2. When eliminating an exogenous node, we must usually replace it with a $\theta$ term. If the node has more than one child, it must be replaced with a $\theta$ term pointing into both children (or an undirected edge connecting them) to preserve the dependency between its children. If the node has a spouse, the eliminated nodes variation must also be preserved using a $\theta$ term. Where the spouse is (already) a $\theta$ term with no other children, $\theta$ terms can be combined.  
3. Since conditioning on a node blocks the path through which it connects its children, we can simply eliminate the node and the arrows between it and its children.
4. An exogenous node with no spouse and only one child can be simply eliminated.
::::


### Reading conditional independence from a graph

We illustrate how to  identify the relations of conditional independence between $A$ and $D$ in Figure \ref{fig:exercise}. 

```{r exercise, echo = FALSE, fig.width = 5, fig.height = 2,  fig.align="center", out.width='80%', fig.cap = "\\label{fig:CItest} An exercise: $A$ and $D$ are conditionally independent, given which other node(s)?"}
hj_ggdag(x = 1:4, y = c(1, 1,1,1), names = c( "A", "B", "C", "D" ),
       arcs = cbind( c(1, 3, 3),
                     c(2, 2, 4)),
       title = " ", padding = .2) 

```

Are A and D independent:

* unconditionally?

Yes. $B$ is a collider, and information does not flow across a collider if the value of the collider node or its consequences is not known. Since no information can flow between $A$ and $C$, no information can flow between $A$ and $D$ simply because any such flow would have to run through $C$.

* if you condition on $B$?

No. Conditioning on a collider opens the flow of information across the incoming paths. Now, information flows between $A$ and $C$. And since information flows between $C$ and $D$, $A$ and $D$ are now also connected by an unbroken path. While $A$ and $D$ were independent when we conditioned on nothing, they cease to be independent when we condition on $B$.

* if you condition on $C$?

Yes. Conditioning on $C$, in fact, has no effect on the situation. Doing so cuts off $B$ from $D$, but this is irrelevant to the $A$-$D$ relationship since the flow between $A$ and $D$ was already blocked at $B$, an unobserved collider. 

* if you condition on $B$ and $C$?

Yes. Now we are doing two, countervailing things at once. While conditioning on $B$ opens the path connecting $A$ and $D$, conditioning on $C$ closes it again, leaving $A$ and $D$ conditionally independent.

Analyzing a causal graph for relations of independence represents one payoff to formally encoding our beliefs about the world in a causal model. We are, in essence, drawing out implications of those beliefs: given what we believe about a set of direct causal relations (the arrows on the graph), what must this logically imply about other dependencies and independencies on the graph, conditional on having observed some particular set of nodes? We show in a later chapter how these implications can be deployed to guide research design, by indicating which parts of a causal system are potentially informative about other parts that may be of interest.
