
```{r setup-ch2, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packagesused02a, include = FALSE}
source("_packages_used.R")
```


# Illustrating Causal Models {#illustratemodels}

:::: {.headerbox data-latex=""}
::: {.center data-latex=""}
:::
We use three  arguments from published political science research to illustrate how to represent theoretical ideas as structural causal models.
::::

<br>

In this short chapter, we provide more of a sense of how we can encode prior knowledge in a causal model by asking how we might construct models in light of extant scholarly works. We undertake this exercise by drawing on three well-known publications in comparative politics and international relations: Paul Pierson's seminal book on welfare-state retrenchment [@pierson1994dismantling]; Elizabeth Saunders' research on leaders' choice of military intervention strategies [@saunders2011leaders]; and Przeworski and Limongi's work on democratic survival [@przeworski1997modernization], an instructive counterpoint to Boix's [@boix2003democracy] argument about a related dependent variable. For each, we represent the causal knowledge that we might plausibly think we take away from the work in question in the form of a causal model. 

Readers might represent these knowledge bases differently;  our aim here is only to illustrate how causal models are constructed, rather than to defend a particular representation (much less the works in question) as accurate.

For each exercise below, we focus on a specific argument in the literature in order to fix in place a relatively clear set of background causal beliefs and simplify the exposition. We emphasize, however, that *in general* a causal model should be thought of as a representation of our state of knowledge or beliefs about causal relations within a domain, rather than as a representation of a specific argument. Suppose, for instance, that we are interested in testing a specific argument in which $X$ affects $Y$ through the mediator $M$. In constructing a causal model to guide our empirical analysis, we cannot simply draw that argument in DAG form ($X \rightarrow M \rightarrow Y$) and leave it at that. In line with the principles relating to conditional independence outlined in Chapter \@ref(models), we must consult our beliefs about this causal domain in a broader sense. For instance, given what we know about the domain from prior observations or studies, is it plausible that $X$ could affect $Y$ through a pathway that does not go through $M$? If we believe it is possible, then we must also draw a direct $X \rightarrow Y$ arrow, or our causal model will steer us wrong --- even if our primary aim is to examine the pathway through $M$. Otherwise, our DAG will contain a relation of conditional independence ($X$ being conditionally independent of $Y$ given $M$) that we do not believe holds. Thus, while we draw on specific works in the illustrations in this chapter, we urge readers to remember that in practice one would want to characterize a broader prior knowledge base in relation to a causal domain in generating a causal model.

We aim to illuminate a number of features of causal models and their construction with these exercises. The examples that we work through variously illustrate how graphs capture beliefs about relations of conditional independence; the potential causal complexity embedded in the causal structures implied by common social-scientific arguments; and the elements of a causal model that cannot be read from a graph. For each work, we discuss both a parametric rendering of the causal functions and a non-parametric formulation built on nodal types. 

## Welfare state reform

The argument in Pierson's 1994 book  *Dismantling the Welfare State?* challenged prior notions of post-1980 welfare-state retrenchment in OECD countries as a process driven primarily by socioeconomic pressures (slowed growth, rising unemployment, rising deficits, aging populations) and the rise of market-conservative ideologies (embodied for instance by the ascendance of Thatcher and Reagan). Pierson argues that socioeconomic and ideological forces put retrenchment on the policy agenda, but do not ensure its enactment because retrenchment is a politically perilous process of imposing losses on large segments of the electorate. Governments will only impose such losses if they can do so in ways that allow them avoid blame for doing so---by, for instance, making the losses hard to perceive or the responsibility for them difficult to trace. These blame-avoidance opportunities are themselves conditioned by the particular social-program structures that governments inherit. 

 <!-- $C$=conservative government; $S$=socioeconomic pressures; $P$=program structure; $A$=retrenchment being on the agenda; $B$=blame-avoidance opportunities; $R$=retrenchment; $\theta^R$=random influences on retrenchment -->

<!-- ```{r pierson, echo = FALSE, fig.width = 12, fig.height = 7, fig.align="center", out.width='.7\\textwidth', fig.cap = "A graphical representation of Pierson (1994).", warning = FALSE, message = FALSE} -->

<!-- par(mfrow = c(1,1)) -->
<!-- par(mar=c(1,1,3,1)) -->
<!-- hj_dag(x = c(1,2,3,1,1,2,3), -->
<!--        y = c(1,1,2,3,2,2,3), -->
<!--        names = c( -->
<!--          expression(paste("P:\nProgram\nstructure")), -->
<!--          expression(paste("B:\nBlame-avoidance\nopportunity")),   -->
<!--          expression(paste("R:\nRetrenchment")), -->
<!--          expression(paste("C:\nConservative\ngovt")), -->
<!--          expression(paste("S:\nSocioeconomic\npressures")), -->
<!--          expression(paste("A:\nOn Agenda")), -->
<!--          expression(paste(theta^R: "\nRandom influence\n on retrenchment"))), -->
<!--        arcs = cbind( c(1,2, 4,5,6,7), -->
<!--                      c(2,3, 6,6,3,3)), -->
<!--        title = "Welfare-State Rentrenchment (Pierson, 1994)", -->
<!--        add_functions = 0,  -->
<!--        contraction = .16,  -->
<!--        padding = .2 -->
<!-- ) -->


<!-- ``` -->

```{r pierson, echo = FALSE, fig.width = 12, fig.height = 7, fig.align="center", out.width='.7\\textwidth', fig.cap = "A graphical representation of Pierson (1994).", warning = FALSE, message = FALSE}

par(mfrow = c(1,1))
par(mar=c(1,1,3,1))
hj_dag(x = c(1,2,3,1,1,2),
       y = c(1,1,2,3,2,2),
       names = c(
         expression(paste("P:\nProgram\nstructure")),
         expression(paste("B:\nBlame-avoidance\nopportunity")),  
         expression(paste("R:\nRetrenchment")),
         expression(paste("C:\nConservative\ngovt")),
         expression(paste("S:\nSocioeconomic\npressures")),
         expression(paste("A:\nOn Agenda"))
         ),
       arcs = cbind( c(1,2, 4,5,6),
                     c(2,3, 6,6,3)),
       title = "Welfare-State Rentrenchment (Pierson, 1994)",
       add_functions = 0, 
       contraction = .16, 
       padding = .2
)


```

While the argument has many more specific features (e.g., different program-structural factors that matter, various potential strategies of blame-avoidance), its essential components can be captured with a relatively simple causal model. We propose such a model in graphical form in Figure \@ref(fig:pierson). Here, the outcome of retrenchment ($R$) hinges on whether retrenchment makes it onto the agenda ($A$) and on whether blame-avoidance strategies are available to governments ($B$). Retrenchment emerges on the policy agenda as a consequence of both socioeconomic developments ($S$) and the ascendance of ideologically conservative political actors ($C$). Inherited program structures ($P$), meanwhile, determine the availability of blame-avoidance strategies. To avoid cluttering the graph, we do not represent the $\theta$ terms, but it is implied that every node on this graph has a $\theta$ node pointing into it.

A few features of this graph warrant attention. As we have discussed, it is the omitted arrows in any causal graph that imply the strongest statements. The graph implies that $C$, $S$, and $P$---which are neither connected along a directed path nor downstream from a common cause---are independent of one another. This implies, for instance, that whether conservatives govern is independent of whether program structures will allow for blame-free retrenchment. Thus, as Pierson argues, a Reagan or Thatcher can come to power but nonetheless run up against an opportunity structure that would make retrenchment politically perilous. Given the absence of bidirectional arrows indicating confounding, the graph similarly implies that the nodal types for all nodes are independent of one another.

Further, this graph represents the belief that any effect of program structures on retrenchment *must* run through their effects on blame-avoidance opportunities. One could imagine relaxing this restriction by, for instance, drawing an arrow from $P$ to $A$: program structures might additionally affect retrenchment in other ways, such as by conditioning the fiscal costliness of the welfare state and thus helping to determine whether reform makes it onto the agenda. If the current state of knowledge suggested that program structures could affect retrenchment via a pathway other than blame-avoidance opportunities, then we would indeed want to include a direct $P \rightarrow A$ arrow. 

Where two variables *are* connected by an arrow, moreover, this does not imply that a causal effect will always operate. Consider, for instance, the arrow pointing from $A$ to $R$. The fact that $A$ sometimes affects $R$ and sometimes does not is, in fact, central to Pierson's argument: conservatives and socioeconomic pressures forcing retrenchment on the agenda will *not* generate retrenchment if blame-avoidance opportunities are absent. 

The graph also reflects a choice about where to begin. We could, of course, construct a causal account of how conservatives come to power, how socioeconomic pressures arise, or why programs were originally designed as they were. Yet it is perfectly permissible for us to bracket these antecedents and start the model with $C$, $S$, and $T$, as long as we do not believe that these variables have any antecedents in common. If they do have common causes, then this correlation should be captured in the DAG.^[In DAG syntax, this correlation can be captured by placing the common cause(s) explicitly on the graph or by drawing a dashed line between the correlated nodes, leaving the source of the correlation unspecified.]

The DAG itself tells us about the possible direct causal dependencies but is silent on the ranges of and functional relations among the variables. How might we express these? With three endogenous variables, we need three functions indicating how their values are determined. Moreover, every variable pointing directly into another variable must be part of that second variable's function. 

Let us assume that all variables (including the implied $\theta$ terms) are binary, with each condition either absent or present. 

One option would be to take a parameteric approach and imagine specific functions connecting parents to children, with $\theta$ terms representing exogenous "noise." For instance, we can capture quite a lot of Pierson's theoretical logic with the following quite simple functional equations:

* $A=CS\theta^C$, capturing the idea that retrenchment makes it on the agenda only if conservatives are in power *and* socioeconomic pressures are high. 
* $B=P\theta^P$, implying that blame-avoidance opportunities arise only when program structures take a particular form.
* $R=AB\theta^R$, implying that retrenchment will occur only if it is on the agenda and blame-avoidance opportunities are present.

In each equation, the $\theta$ term allows for exogenous forces that might block the outcome from occurring. In the last functional equation, for instance, retrenchment will only occur if retrenchment is on the agenda and blame-avoidance opportunities are present, but even if both are present, the effect on retrenchment also hinges on the value of $\theta^R$. When $\theta^R=1$, the $AB$ combination has a positive causal effect on retrenchment. When $\theta^R=0$, $AB$ has no causal effect: retrenchment will not occur regardless of the presence of $AB$. We can think of $\theta^R$ as capturing a collection of features of a case's context that might render the case susceptible or not susceptible to an $AB$ causal effect. For instance, Pierson's analysis suggests that a polity's institutional structure might widely diffuse veto power such that stakeholders can block reform even when retrenchment is on the agenda and could be pursued without electoral losses. We could think of such a case as having a $\theta^R$ value of 0, implying that $AB$ has no causal effect. A $\theta^R=1$ case, with a positive effect, would be one in which the government has the institutional capacity to enact reforms that it has the political will to pursue.

Alternatively, we could take a non-parameteric approach, as we generally do in the remainder of this book. In a non-parametric setup, each node's $\theta$ term captures that node's nodal type. Each value of a $\theta$ term's range represents a possible way in which the node might respond to its parents. We would define $\theta^A$ as taking on one of 16 values (16 types, given 2 parent nodes); $\theta^B$ as taking on on one of four values; and $\theta^R$ as taking on one of 16 values; with $\theta^C$ and $\theta^S$ each taking on one of two values. 

Then choices on the probability distributions fully reflect the ways these variables relate to each other. Note that the parametric argument given above can be thought  of as a special case of the non-parametric representation with all probability mass placed on a small set of possible nodal types. Thus the central thrust of Pierson's argument could then be represented in nodal-type form as:

* $\theta^A=\theta^A_{0001}$ 
* $\theta^B=\theta^B_{01}$
* $\theta^R=\theta^R_{0001}$. 

In practice, however we would allow for a richer probability  *distribution* over each $\theta$, representing beliefs over the assignment process or causal relations operating at each node. Beliefs about the distribution of exogenous conditions would be captured in distributions over the values of $\theta^C, \theta^S$, and $\theta^P$. How we handle distributions over $\theta^A, \theta^B$, and $\theta^R$ depends on the degree of confidence that we want to express in Pierson's argument. To represent the belief that Pierson's argument is correct with certainty and operates in uniform, deterministic fashion across units, we would simply have degenerate distributions for $\theta^A, \theta^B$, and $\theta^R$, with a probability of 1.0 placed on the respective nodal types shown above. To capture uncertainty about the functional relations on any graph or if we believe that there is some heterogeneity of effects across units we would disperse  probability density across types for each $\theta$. For instance, for $\theta^R$ we might want to put some weight on $\theta^R_{0011}$ (blame-avoidance opportunities alone are enough to generate retrenchment),  $\theta^R_{0101}$ (conservative leaders alone are enough), $\theta^R_{0111}$ (either is enough), and $\theta^R_{0000}$ (retrenchment will not happen even when both conditions are present), while perhaps putting greatest weight on $\theta^R_{0001}$.^[In notation that we use later, these beliefs would be represented with a $\lambda^R$ vector.]


## Military Interventions

@saunders2011leaders asks why, when intervening militarily abroad, do leaders sometimes seek to transform the *domestic* political institutions of the states they target but sometimes seek only to shape the states' *external* behaviors. 

Saunders' central explanatory variable is the nature of leaders' causal beliefs about security threats. When leaders are "internally focused," they believe that threats in the international arena derive from the internal characteristics of other states. Leaders who are "externally focused," by contrast, understand threats as emerging strictly from other states' foreign and security policies. 

These basic worldviews, in turn, affect the cost-benefit calculations leaders make about intervention strategies---in particular, about whether to try to transform the internal institutions of a target state---via two mechanisms. First, an internal focus (as opposed to an external focus) affects leaders' perceptions of the likely security gains from a transformative intervention strategy. Second, internal vs. external focus affects the kinds of strategic capabilities in which leaders invest over time (do they invest in the kinds of capabilities suited to internal transformation?); and those investments in turn affect the costliness and likelihood of success of alternative intervention strategies. Calculations about the relative costs and benefits of different strategies then shape the choice between a transformative and non-transformative approach to intervention. 

At the same time, leaders can only choose a transformative strategy if they decide to intervene at all. The decision about whether to intervene depends, in turn, on at least two kinds of considerations. The first is about fit: a leader is more likely to intervene against a target when the nature of the dispute makes the leader's preferred strategy appear feasible in a given situation. Second, Saunders allows that forces outside the logic of her main argument might also affect the likelihood of intervention: in particular, leaders may be pushed to intervene by international or domestic audiences.

Figure \ref{fig:DAGSaunders} depicts the causal dependencies in Saunders' argument in DAG form (again, with all $\theta$ terms implied). Working from left to right, we see that whether or not leaders are "internally focused" ($F$) affects the expected net relative benefits of transformation ($B$), both via a direct pathway and via an indirect pathway running through investments in transformative capacities ($T$). Characteristics of a given dispute or target state ($D$) likewise influence the benefits of transformation ($B$). The decision about whether to intervene ($I$) is then a function of three factors: internal focus ($F$), the expected relative net benefits of transformation ($B$), and audience pressures ($A$). Finally, the choice of whether to pursue a transformative strategy ($S$) is a function of whether or not intervention occurs at all ($I$), and of cost-benefit comparisons between the two strategies ($B$).

<!-- ```{r saunders, echo = FALSE, fig.width = 12, fig.height = 7, fig.align="center", out.width='.7\\textwidth', fig.cap = "\\label{fig:DAGSaunders} A graphical representation of Saunders' (2011) argument.", warning = FALSE, message = FALSE} -->

<!-- par(mfrow = c(1,1)) -->
<!-- par(mar=c(1,1,3,1)) -->
<!-- hj_dag(x = c(1,2,2,1, 3, 3,3,3.5), -->
<!--        y = c(1,0,2,2, 2, 1,0, .5), -->
<!--        names = c( -->
<!--          expression(paste("C: Causal beliefs")), -->
<!--          expression(paste("P: Preparedness\ninvestment")),   -->
<!--          expression(paste("B: Benefits expected\nfrom transfers")),   -->
<!--          expression(paste("T: Target\ncharacteristics")), -->
<!--          expression(paste("A: Audience pressures")), -->
<!--          expression(paste("I: Intervention")), -->
<!--          expression(paste("S: Intervention strategy")), -->
<!--          expression(paste(U[S]: "Random influence\non strategy"))  -->
<!--          ), -->
<!--        arcs = cbind( c(1,1, 2,1, 4, 3, 3,5,6,8), -->
<!--                      c(2,3, 3,6, 3, 6, 7,6,7,7)), -->
<!--        title = "(b) Military intervention strategies (Saunders, 2011)", -->
<!--        add_functions = 0,  -->
<!--        contraction = .16,  -->
<!--        padding = .2 -->
<!-- ) -->


<!-- ``` -->


```{r saunders, echo = FALSE, fig.width = 12, fig.height = 7, fig.align="center", out.width='.7\\textwidth', fig.cap = "\\label{fig:DAGSaunders} A graphical representation of Saunders' (2011) argument.", warning = FALSE, message = FALSE}

par(mfrow = c(1,1))
par(mar=c(1,1,3,1))
hj_dag(x = c(1,2,2,1, 3, 3,3),
       y = c(1,0,2,2, 2, 1,0),
       names = c(
         expression(paste("F: Internal focus")),
         expression(paste("T: Invest in\ntransformation\ncapablities")),  
         expression(paste("B:\nExpected benefits\nfrom transformation")),  
         expression(paste("D: Dispute\ncharacteristics")),
         expression(paste("A: Audience pressures")),
         expression(paste("I: Intervene")),
         expression(paste("S: Transformative strategy")) 
         ),
       arcs = cbind( c(1,1, 2,1, 4, 3, 3,5,6),
                     c(2,3, 3,6, 3, 6, 7,6,7)),
       title = "Military intervention strategies (Saunders, 2011)",
       add_functions = 0, 
       contraction = .16, 
       padding = .2
)


```


This DAG illustrates how readily causal graphs can depict the multiple pathways through which a given variable might affect another variable, as with the multiple pathways linking $F$ to $I$ and $B$ (and, thus, all of its causes) to $S$. In fact, this graphical representation of the dependencies in some ways throws the multiplicity of pathways into even sharper relief than does a narrative exposition of the argument. For instance, Saunders draws explicit attention to how causal beliefs operate on expected net benefits via both a direct and indirect pathway, both of which are parts of an indirect pathway from $F$ to the outcomes of interest, $I$ and $S$. What is a bit easier to miss without formalization is that $F$ also acts *directly* on the choice to intervene as part of the feasibility logic: when leaders assess whether their generally preferred strategy would be feasible if deployed against a particular target, the generally preferred strategy is itself a product of their causal beliefs. The DAG also makes helpfully explicit that the two main outcomes of interest---the choice about whether to intervene and the choice about how---are not just shaped by some of the same causes but are themselves causally linked, with the latter depending on the former.

Omitted links are also notable. For instance, the lack of an arrow between $D$ and $A$ suggests that features of the dispute that affect feasibility have no effect on audience pressures. If we instead believed there could be other connections---for instance, that audiences take feasibility into account in demanding intervention---then we would want to include a $D \rightarrow A$ arrow.

Turning to variable ranges and functional equations, it is not hard to see how one might readily capture Saunders' logic in a fairly straightforward set-theoretic manner. All variables except $S$ could be treated as binary with, for instance, $F=1$ representing internally focused causal beliefs, $T=1$ representing investments in transformative capabilities, $B=1$ representing expectations that transformation will be more net beneficial than non-transformation, $D=1$ meaning that a dispute has characteristics that make transformation a feasible strategy, and so on. Although there are two strategies, we in fact need three values for $S$ because it must be defined for all values of the other variables---i.e., it must take on a distinct categorical value if there is no intervention at all. We could then define functions, such as:

* $B=FTD$, implying that transformation will only be perceived to be net beneficial in a case if and only if the leader has internally focused causal beliefs, the government is prepared for a transformative strategy, and the dispute has characteristics that make transformation feasible.
* $I=(1-|B-F|)+(1-(1-|B-F|))A$, implying that intervention can occur under (and only under) either of two alternative sets of conditions: if the generally preferred strategy and the more net-beneficial strategy in a given case are the same (i.e., such that $B-F=0$) or, when this alignment is absent (i.e., such that $|B-F|=0$), where audiences pressure a leader to intervene.

As illustrated in the Pierson example, in a non-parametric framework, each parametric functional equation represents one nodal type for the relevant $\theta$. For instance, though we spare the reader the complexities of the corresponding subscript notation, there is a single value of $\theta^B$ under which the conditions $F=1, P=1,$ and $T=1$ generate $B=1$, and we get $B=0$ otherwise.  Likewise, there exists a single value of $\theta^I$ under which $B=1, F=1$ and $B=0, F=0$ produce $I=1$, for either value of $A$; and $A$ has a positive effect on $I$ whenever $B \neq F$. To work with this model, we would specify a probability distribution over all possible nodal types for each node on the graph.

This example also nicely illustrates how much potential causal complexity a moderately intricate argument and causal graph implies. The number of *possible* nodal types at each node depends on how many parents that node has. Looking at the endogenous nodes here, we have one node with one parent ($T$), implying 4 nodal types; one node with two parents ($S$), implying 16 nodal types; and two nodes with 3 parents ($B$ and $I$), implying 256 nodal types each. If we now conceptualize the set of possible "causal types" as containing all distinct combinations of nodal types---all ways in which a case might behave across all of its nodes (see Chapter \@ref(models))---then this graph implies about 4 million different ways in which the values of exogenous nodes ($D$, $F$, and $A$) might jointly produce patterns of outcomes. Saunders' argument effectively represents one of these 4 million possible sets of relations. 

The framework that we outline in this book allows for updating on arguments like Saunders': we can ask how likely the specific causal type implied by this argument is relative to other causal types. Yet, as we will see, the approach lends itself to a much broader view of causal inquiry. In general, we will use data to update beliefs over *all* causal types allowed for in a model, and then use these updated beliefs to answer any number of causal questions about relations in the model. For instance, we can use the same data and updated model to ask about the average effect of internal focus on intervention; the relative importance in this effect of the expected-benefits pathway over the direct pathway; about individual steps in the causal chain, such as the effect of expected benefits on choice of strategy; and so on.

## Development and Democratization

@przeworski1997modernization argue that democratization occurs for reasons that are, with respect to socioeconomic or macro-structural conditions, largely idiosyncratic; but once a country has democratized, a higher level of economic development makes democracy more likely to survive. Economic development thus affects whether or not a country is a democracy, but only after a democratic transition has occurred, not before. Thus, in their description---and contrary to @boix2003democracy ---democratization is "exogenous": it is not  determined by other variables in the model. The dynamic component of Przeworski and Limongi's argument---the fact that both the presence of democracy and the causal effect of development on democracy depend on whether a democratic transition occurred at a previous point in time---forces us to think about how to capture over-time processes in a causal model. 

We represent Przeworski and Limongi's argument in the DAG in Figure \@ref(fig:DAGPL). The first thing to note is that we can capture dynamics by considering democracy at different points in time as separate nodes. According to the graph, whether a country is a democracy in a given period ($D_t$) is a function, jointly, of whether it was a democracy in the previous period ($D_{t-1}$) and of the level of per capita GDP in the current period (as well as of other unspecified forces $\theta^{D_t}$, not pictured).




```{r DAGPL, echo = FALSE, fig.width = 10, fig.height = 6, fig.align="center", out.width='.7\\textwidth', fig.cap = "A graphical representation of Przeworski and Limongi's argument, where $D_{t-1}$=democracy in the previous period; $GDP_t$=per capita GDP in the current period; $D_t$=democracy in the current period.", warning = FALSE, message = FALSE}

par(mfrow = c(1,1))
par(mar=c(1,1,3,1))
hj_dag(x = c(1,1,2),
       y = c(2,0,1),
       names = c(
         expression(paste(D[t-1]: "Democracy,\nlast period")),
         expression(paste(Y[t]: "GDP per capita,\nthis period")),  
         expression(paste(D[t]: "Democracy,\nthis period"))
         ),
       arcs = cbind( c(1,2),
                     c(3,3)),
       title = "Democratization (Przeworski and Limongi, 1997)",
       add_functions = 0, 
       contraction = .16, 
       padding = .2
)


```


Second, the arrow running from $GDP_{t-1}$ to $D_t$ means that $GDP$ *may* affect democracy, not that it always does. Indeed, Przeworski and Limongi's argument is that development's effect depends on a regime's prior state: GDP matters for whether democracies continue to be democracies, but not for whether autocracies go on to become democracies. The absence of an arrow between $D_{t-1}$ and $GDP_{t-1}$, however, implies a (possibly incorrect) belief that democracy and $GDP$ in the last period are independent of one another.

Inspection of this figure highlights, we think, a curious feature of this argument. The key claim---that the switch *to* democracy does not depend on income---is not readable from the graph. The reason is simply that, given this argument, being a non democracy in one period given you were a democracy in a previous period does depend on income. Your  state in the second period does depend on income. Specifically there is a *counterfactual* dependence---income causes a state to be democratic even if it does not cause it to *transition to* democracy. The effect of income may well be asymmetric depending on whether you start as a democracy or start as an autocracy but this asymmetry has to be be captured by the specification of the functional relations; it is not captured by the graph. 

For a parametric representation of this asymmetric relationship we can specify a function in which  $GDP$ can reduce the likelihood of a transition *away* from democracy but does affect the probability of a transition *to* democracy, which should be exogenously determined. One possible translation of the argument into functional terms is:


$$D_t = \mathbb{1}(\theta^{D_t} + (1-D_{t-1})p + D_{t-1}Y_{t-1}q) > 0)$$

<!-- FLAG: Get proper indicator function \mathbb{1} -->

where

* $D_t$ and $D_{t-1}$ are binary, representing current and last-period democracy, respectively
* $p$ is a parameter representing the probability that an autocracy democratizes
* $q$ is a parameter representing the probability that a democracy with a GDP of 1 remains democratic
* $Y_{t-1}$ represents national per capita GDP, scaled to 0-1.
* $\theta^{D_t}$ represents a random, additional input into democracy
* the indicator function, ${1}$, evaluates the inequality and generates a value of $1$ if and only if it is true

Unpacking the equation, the likelihood that a country is a democracy in a given period rises and falls with the expression to the left of the inequality operator. This expression itself has two parts, reflecting the difference between the determinants of *transitions to* democracy (captured by the first part) and the determinants of democratic *survival* (captured by the second). The first part comes into play---i.e., is non-zero---only for non-democracies. For non-democracies, the expression evaluates simply to $p$, the exogenous probability of democratization. The second part is non-zero only for democracies, where it evaluates to $q$ times $Y_{t-1}$: thus, remaining democratic is more likely as national income rises. The inequality is then evaluated by asking whether the expression on the left passes a threshold. Thus, higher values for the expression increase the likelihood of democracy while the randomness of the $\theta^{D_t}$ threshold captures the role of other, idiosyncratic inputs. The mean and variance of $\theta^{D_t}$ capture the overall likelihood of being a democracy as well as the importance of unspecified factors.^[Note how, while the functional equation nails down certain features of the process, it leaves others up for grabs. In particular, the parameters $p$ and $q$ are assumed to be constant for all autocracies and for all democracies, respectively, but their values are left unspecified. And one could readily write down a function that left even more openness---by, for instance, including an unknown parameter that translates $y$ into a change in the probability of reversion or allowing for non-linearities, with unknown parameters, in this effect.] In a model like this it would be natural to seek to estimate parameters $p$ and $q$ as well as trying to understand the distribution of $\theta^{D_t}$.

We can also represent the asymmetry in the binary set up with causal types that we developed in the last chapter. 

Type $\theta^{D_t}_{0001}$ is a type for which the regime type *will stay as they are* if they are wealthy, but will become authoritarian if they are not wealthy.  To be clear, wealth still affects whether or not a state is a democracy, rather than an autocracy, in this period--counterfactually---but wealth does not make a non democracy become a democracy. In other words it causes a case to *be* a democracy but not to *become* a democracy. This  type can be distinguished from a $\theta^{D_t}_{0011}$  type in which a non democracy becomes a democracy when income is high. 

Although we do not engage with dynamic models in this book, it is instructive to think through the implications of a distribution of causal types for a dynamic process. Say we were to imagine that income were constant but that in each period one half of units were of type  $\theta^{D_t}_{0001}$, and one half of type $\theta^{D_t}_{1111}$ (with types drawn afresh each period).  Say that in an initial period, half the units were democracies and half had high income and there was no relation between these two features. Then in the next period we would have that 
half of cases would be democracies (regardless of income), half of which would be surviving democracies and half new democracies; of the other half, one quarter would be surviving democracies (surviving *because of* their income), and the other three quarters would be autocracies, one third of which would be "backsliders" because of their poverty. Similar transitions occur in future periods until eventually the wealthy states are all stable democracies and the  poorer states transition between democracy and autocracy back and forth randomly each period.

In this approach there are no parameters $p$ or $q$ to be estimated. Rather the focus is entirely on the distribution of nodal types. 

<!-- 	T1	A	D	T2	A	D	T3 -->
<!-- AP	16	8	8	16	8	8	16 -->
<!-- AR	16	8	8	8	4	4	4 -->
<!-- DP	16	8	8	16	8	8	16 -->
<!-- DR	16	0	16	24	0	24	28 -->
