# Case selection as a Decision Problem {#caseselection}

```{r packagesused13, include = FALSE}
source("_packages_used.R")
do_diagnosis = FALSE
```

:::: {.headerbox data-latex=""}
::: {.center data-latex=""}
:::
With a causal model in hand, together with priors over parameters, you can assess in advance what conclusions you will draw from different observations and assess what kinds of observations are most worth seeking. We draw out the implications of this idea for case selection.
::::
<br>


A critical decision for scholars employing mixed methods is to determine which cases are most valuable for within-case analysis.  

A host of different strategies have been proposed for selecting cases for in-depth study based on the observed values of $X$, $Y$ data. Perhaps the most common strategy is to select cases in which $X=1$ and $Y=1$ and look to see whether in fact $X$ caused $Y$ in the case in question (using some more or less formal strategy for inferring causality from within-case evidence). But many other strategies have been proposed, including strategies to select cases "on the regression line" or, for some purposes, cases "off the regression line" (e.g., @Lieberman2005nested). Some scholars suggest ensuring variation in $X$ (most prominently, @king1994designing), while others have proposed various kinds of matching strategies. Some have pointed to the advantages of random sampling of cases, either stratified or unstratified by values on $X$ or $Y$ (@FL2008, @HerronQuinn). 

Which cases you should choose will likely depend on the purposes to which you want to put them. 

A matching strategy for instance---selecting cases that are comparable on many features but that differ on $X$---replicates at a small scale the kind of inference done by matching estimators with large-$n$ data. The strategy emphasize the inferences to be made from $X,Y$ variation rather than inferences drawn specifically from within case information beyond what is available in the measurement of $X$ and $Y$. (Citations needed.)

Other treatments seek to use qualitative information to check assumptions made in $X, Y$ analysis: for example, is the measurement of $X$ and $Y$ reliable in critical cases? (Citations needed) For such questions with limited resources, it might make sense to focus on cases for which validation plausibly makes a difference to the $X,Y$ inferences: for example influential cases that have unusually extreme values on $X$ and $Y$.^[Note: We can say more about why these would be good choices from a Bayesian perspective, based on the idea that measurement is more likely to be wrong in such cases and shifting them to more typical values would make a big difference.] Similar arguments are made for checking assumptions on selection processes, though we consider this a more complex desideratum since this requires making case level causal inferences and not simply measurement claims.

A third purpose is to use a case to generate alternative or richer theories of causal processes, as in Lieberman's "model-building" mode of "nested analysis" (@Lieberman2005nested). Here it may be cases off the regression line that are of interest.

Weller and Barnes (CITE article) on case selection focus on (a) X/Y relations and (b) whether the cases are useful for hypothesis generation. 

In what follows, we focus on a simpler goal: given existing $X, Y$ data for a set of cases and a given clue (or set of clues) that we can go looking for in the intensive analysis of some subset of these cases, for which cases would process tracing yield the greatest learning about the population-level causal effect of $X$ on $Y$?

The basic insight of this chapter is simple enough: *the optimal strategy for case selection for a model-based analysis can be determined by the model and the query*, just as we saw for the optimal clue-selection strategy in Chapter \@ref(Clues). Using this strategy yields guidance that is consistent with some common advice but at odds with other advice. The main principles that emerge from the analysis can be summarized as:

* go where the probative value is, and
* sample from $X$ and $Y$ values in proportion to their occurrence in the population,
* invest in collections of cases that provide complementary learning. 

Beyond these general principles, other patterns are more complex and thus more difficult to neatly summarize. The most general message of this chapter is about the general approach: that is, that we can use a causal model to tell us what kinds of cases are likely to yield the greatest learning, given the model and a strategy of inference. We provide a tool for researchers to undertake this analysis, at least for simple problems with $X, Y, K$ data.


Most closely related to our analysis in this chapter is the contribution of @HerronQuinn, who build on @SeawrightGerring2008. While Seawright and Gerring provide a taxonomy of approaches to case selection, they do not provide a strategy for assessing the relative merits of these different approaches.  As we do, @HerronQuinn focus on a situation with binary $X,Y$ data and assess the gains from learning about causal type in a set of cases (interestingly in their treatment causal type, $Z_i$  is called a confounder rather than being an estimand of direct interest; in our setup, confounding as normally understood arises because of different probabilities of different causal types of being assigned to "treatment", or an $X=1$ value). @HerronQuinn assume that in any given case selected for analysis a qualitative researcher is able to infer the causal type perfectly.  

Our setup differs from that in @HerronQuinn in a few ways.  @HerronQuinn paramaterize differently, though this difference is not important.^[@HerronQuinn have a parameter $\theta$ that governs the distribution of data over $X$ and $Y$ and then, conditional on $X,Y$ values, a set of parameters $\psi_{xy}$ that describe the probability of a case's being of a given causal type. We take both $\theta$ and $\psi_{xy}$ to derive from the fundamental distribution of causal types and assignment probabilities. Thus, for example, $\psi_{00}$ from @HerronQuinn corresponds to $\frac{(1-\pi_b)\lambda_b}{(1-\pi_b)\lambda_b + (1-\pi_c)\lambda_c}$ in our notation. The difference in  paramaterization does have implications for interpretations of the priors. For example flat priors over $\theta$ and $\psi$ implies a tighter distribution that a uniform prior over the causal types. In fact @HerronQuinn use priors with greater variance than uniform in any event.] Perhaps the most important difference between our analysis and that in @HerronQuinn  is that we connect the inference strategy to process-tracing approaches. Whereas @HerronQuinn assume that causal types can be read directly, we assume that these are inferred *imperfectly* from clues. As in our baseline model, our ability to make inferences for causal types can differ by type and as a function of $X$. And, as in the baseline model, not only can we have uncertainty about the probative value of clues, but researchers can learn about the probative value of clues by examining cases.

<!-- Are Herron and Quinn's priors Jeffrey priors? -->

Here we assume that the case selection decision is made after observing the $XY$ distribution and we explore a range of different possible contingency tables. In  @HerronQuinn the distribution from which the contingency tables are drawn is fixed, though set to  exhibit an expected  observed difference in means (though not necessarily a true treatment effect) of 0.2. They assume large $XY$ data sets (with 10,000) units and case selection strategies ranging from 1 to 20 cases.

Another important difference, is that in many of their analyses, @HerronQuinn take the perspective of an outside analyst who knows the true treatment effect; they then assess the expected bias generated by a research strategy over the possible data realizations. We, instead, take the perspective of a researcher who has *beliefs* about the true treatment effect that correspond to their priors, and for whom there is therefore no *expected* bias. This has consequences also for the assessment of expected posterior variance, as in our analyses the expectation of the variance is taken with respect to the researcher's beliefs about the world, rather than being made conditional on some specific world (ATE). We think that this setup is addressed to the question that a researcher must answer when deciding on a strategy: given what they know now, what will produce the greatest reduction in uncertainty (the lowest expected posterior variance)?

Finally, we proceed somewhat differently in our identification of strategies from Herron and Quinn: rather than pre-specifying particular sets of strategies (operationalizations of those identified by @SeawrightGerring2008) and evaluating them, we define a strategy as the particular distribution over $XY$ cells to be examined and proceed to examine *every possible strategy* given a choice of a certain number of cases in which to conduct process tracing. We thus let the clusters of strategies---those strategies that perform similarly---emerge from the analysis rather than being privileged by past conceptualizations of case-selection strategies.

Despite these various differences, our results will agree in key ways with those in @HerronQuinn.


## Case selection logics depends on probative value and queries

Although it might be tempting to seek general rules of the form "examine cases in which $X=1$ and $Y=1$" or, "ignore cases in which $X=0$ and $Y=0$", it is easily demonstrated that which choices are more informative depend on how waht can be learned within a case contributes to infernecnce, and that hte contributions depend on the query in question.

Say we know in a given population that:

  * $X \rightarrow Y \leftarrow K$
  * $\Pr(Y=1|X=0, K = 0) = 1$
  * $\Pr(Y=1|X=1, K = 0) = .5$
  * $\Pr(Y=1|X=0, K = 1) = 0$
  * $\Pr(Y=1|X=1, K = 1) = .9$

We do not know, however how common $K$ is. Thus we do not know either unit level causal effects or population level effects.

Given the information above we can see that if $X=Y=1$, then $K$ is a "doubly decisive" clue for assessing whether, in a given case, $X$ causes $Y$. In particular we have that for an $X=Y=1$ case, seeing $K=1$ implies  $X$ caused $Y$ Since otherwise $Y$ would have been 0) and seeing $K=0$ implies   $X$ did not cause $Y$ (since othereise $Y$ would have been 1 in any event.

However, if we had a case in which $X=Y=0$ then learning $K$ would be entirely uninformative for the case. In particular, we already know that $K=1$ in this case as we know there are not cases in which  $X=Y=0$ and $K=0$. And this knowledge is not enough in this case to know whether $X=0$ caused $Y=0$.

For the same reason, if, off the diagonal, $X=0, Y=1$ we learn nothing from $K$, since we learn about $K$ from $Y$. If however we had a case in which $X=1, Y=0$, if $K$ were 1 then we would know that $X=1$ did not cause $Y=0$; if $K$ were 0 we would know that  $X=1$ caused $Y=1$.


```{r, eval= FALSE, echo = FALSE}
py_k0 = .5
py_k1 = .9

py_k1 / (py_k0 + py_k1)

(1-py_k1) / (1-py_k0 + 1-py_k1)

```

Say now we were interested in the population estimand: the average effect of $X$ on $Y$. We can see that this is equal to $\Pr(K=1)\times.9 + (1-\Pr(K-0))\times(-.5)) = 1.4\Pr(K=1)-.5$. For this estimand we are only interested in the prevalence of $K$ in the population. It might seem that this means that it is irrelevant what type of case you choose; however, as noted above, we have more information about the likely value of $K$ in some cases than in others. So for this estimand also, in this case, selecting an $X=Y=1$ case  is informative for the population effect; selecting $X=Y=0$ case is not informative. 

Say now we had a case in which $X=1$. Should we choose: a case in which $Y=1$ or a case in which $Y=0$. In both cases $K$ is doubly decisive for the case level estimand. However for a $Y=1$ case we think it likely that $K=1$ (specifically, assuming a prior of $\Pr(K=1)$ we think $\Pr(K=1 | X=1, Y=1) = \frac{.9}{.9+.5}=.64$); for the  $Y=0$ case we think  $\Pr(K=1 | X=1, Y=0) = \frac{.1}{.5+.1}=.17$. Thus we are much more uncertain about the value of $K$ in the $X=Y=1$ case. We would learn more for the avergae treatment effect then by choosing off the diagonal.

Specifically let $\kappa$ denote $\Pr(K=1)$. say we begin thinking it equally likely that $\kappa=\kappa^H = .5$ and $\kappa=\kappa^L=0$. Say $\Pr(X=1) = .5$. Say we observe one case with $X=Y=1$ and another with $X=1, Y=0$. From that informaation alone we can update over $\kappa$. Specifically, conditioning on the probability that $X=1$ for both cases and focuseing only on the probability that Y=0 in one case and Y=1 in the other:

$$p(\kappa = \kappa^L|D) =  \frac{p(D|\kappa^H)}{p(D|\kappa^H)+p(D|\kappa^L)}=\frac{.5}{.5 + .25\times(2\times.5\times.5 + 2\times.9\times.1 + 2\times(.9\times.5 +.1\times.5))}$$



In summary in this case it makes sense to select and $X=Y=1$ case if you are interested in teh population estimand or if you are interested in teh case level estimand for an $X=Y=1$ case. If you are interested in the case level estimand for an $X=Y=0$ case then there are no gains from either selection strategy.

Sometimes however you learn more in a $X=Y=0$ than in a  $X=Y=1$ case. Say instead that you knew that: 

  * $X \rightarrow Y \leftarrow K$
  * $\Pr(Y=1|X=0, K = 0) = .5$
  * $\Pr(Y=1|X=1, K = 0) = 0$
  * $\Pr(Y=1|X=0, K = 1) = .5$
  * $\Pr(Y=1|X=1, K = 1) = 1$

In this case you learn nothing from observing a case in which $X=Y=1$---in this case you already expect that $K=1$. In contrast if $X=Y=0$ then if you learn that $K=1$ you know that were $X=1$ the $Y$ would have been 1; but if you learn that $K=0$ you know that *were* $X=1$ then $Y$ would have (still) been 0. So in this case $K$ is doubly decisive for an $X=Y=0$ case but not for a   $X=Y=1$ case. 


This logic also holds for the off diagonals.

Say instead that you knew that: 

  * $X \rightarrow Y \leftarrow K$
  * $\Pr(Y=1|X=0, K = 0) = 0$
  * $\Pr(Y=1|X=1, K = 0) = .5$
  * $\Pr(Y=1|X=0, K = 1) = 1$
  * $\Pr(Y=1|X=1, K = 1) = .5$

Say you have a case in which $X=1, Y=0$, then if $K=1$ you know that if $X$ were 0, Y would have been 1; but if $K=0$ you know that if $X$ were 0, $Y$ would have been 0. In that case $K$ would be doubly decisive for $X=1$ causing $Y=0$.  

Say you have a case in which $X=0, Y=1$, then it must be that $K=0$.   
  
Say instead that you knew that: 

  * $X \rightarrow Y \leftarrow K$
  * $\Pr(Y=1|X=0, K = 0) = .5$
  * $\Pr(Y=1|X=1, K = 0) = 0$
  * $\Pr(Y=1|X=0, K = 1) = .5$
  * $\Pr(Y=1|X=1, K = 1) = 1$

It may be better to select a case that is not "like" the cases you want to make inferences about.


Which cases to learn $K$ from.

## New section (Dec 2020)

We turn now to a more general approach for deriving case-selection guidance from a causal model, given the causal query we wish to address and any data that we have already observed. This approach can be implemented using `CausalQueries` together with `CQTools`.

### Procedure

We imagine a situation in which we have already observed some data (the values of some nodes from the causal model in some set of cases) and must now decide in which cases we should gather additional data. To simplify the setup, we will be assuming that we are considering gathering additional observations in cases for which we already have *some* data. In other words, we are deciding which cases to investigate more *deeply*. (This is distinct from the question of "wide vs. deep", where we might decide to observe cases we have not yet seen at all.) 

The general intuition of the case-selection approach that we develop here is that we can use our causal model and any previously observed data to estimate what observations we are more or less likely to make under a given case-selection strategy, and then figure out how far off from the (under the model) true estimand we can expect to be under the strategy, given whatever causal question we seek to answer. 

We proceed as follows:

**DAG**. We start, as always, with a DAG representing our beliefs about which variables we believe to be direct causes of other variables. For the current illustrations, we consider two different DAGS: a simple mediation (or "chain") model, $X \rightarrow M \rightarrow Y$, and a two-path model, $X \rightarrow M \rightarrow Y \leftarrow X$.

**Priors**. As when conducting mixed-method inference, we can set qualitative restrictions and/or differential quantitative weights on the (possibly conditional) nodal types in the model. We can also indicate our uncertainty over the latter, by setting the $\alpha$ parameters of the relevant Dirichlet distributions. For the current example, we start by setting flat priors over all nodal types and assume no unobserved confounding.

**Given data.** If we have already made observations of any of the model's nodes in some set of cases, we can use this information to condition our strategy for searching for further information. For instance, if we have observed $X$'s and $Y$'s value in a set of cases, we might select cases for process tracing based on their values of $X$ and/or $Y$. And, importantly, what we have already observed in the cases will affect the inferences we will draw when we observe additional data, including how *informative* a particular new observation is likely to be. For the present examples, we assume that we have already observed $X$ and $Y$ in a set of cases and found a positive correlation.

**Query**. We define our query. This could, for instance, be the share of cases in the population in which $X$ has a positive effect on $Y$, or it might be $X$'s average effect on $Y$. We can use the general procedure to identify case-selection strategies for any causal query that can be defined on a DAG. And, importantly, the optimal case-selection strategy may depend on the query. For instance, the best case-selection strategy for estimating the average causal effect of $X$ on $Y$ may not be the same as the best strategy for figuring out for what proportion of the population $X$ has a positive effect on $Y$. In the example below, we focus on a set of conditional queries, asking what the effect of $X$ on $Y$ is in cases with different $X$, $Y$, and $M$ values. 

FLAG: MAke sure above query definition lines up with what we end up doing.

**Define one or more strategies**. A strategy is defined, generically, as the search for data on a given set of *nodes*, in a given *number* of cases that are randomly selected *conditional* on some information we already have about potential cases. Let us assume here that our strategy will involve uncovering $M$'s value in 1 or 2 cases. What we are wondering is how to choose this one or two cases for deeper analysis. For illustrative purposes, we focus in on four possible strategies, conditional on the $X$ and $Y$ values that we already know: 

1 examine $M$ in 1 $X=Y=1$ case; 
2 examine $M$ in 1 $X=1, Y=0$ case; 
3 examine $M$ in 2 cases, an $X=Y=1$ and an $X=Y=0$; and 
4 examine $M$ in 2 cases, an $X=1, Y=0$ and an $X=0, Y=1$.

In light of the positive $X,Y$ correlation already observed, we can think of strategies 1 and 3 as "on-the-regression line" strategies, while strategies 2 and 4 are "off-the-regression line" strategies.

**Possible data**. For each strategy, there are multiple possible sets of data that we could end up observing. In particular, the data we could end up with will be the $X,Y$ patterns we have already observed plus some pattern of $M$ observations. Thus, for instance, for strategy 1, we could end up observing the initial $X,Y$ pattern plus $M=0$ in one of the $X=1, Y=1$ cases, or the initial $X,Y$ pattern plus $M=1$ in one of the $X=1, Y=1$ cases. For strategy 3, we could observe four possible combinations of $M$ values, with $M$ being 0 or 1 in each of the cases. 

**Probability of the data**. We now calculate a probability of each possible data realization, given the model and the data that we have already observed. In practice, we do this in `CQtools` via simulation. Starting with the model together with our priors, we update our beliefs about $\lambda$ based on the previously observed data. This posterior now represents our *prior* for the purposes of the process tracing. In the present example, we are using the already-observed $X,Y$ correlation to update our beliefs about causal-type share allocations in the population, having seen the $X,Y$ data only. We then use this posterior to draw a series of $\lambda$ values. 

Given that the ambiguity matrix gives us the mapping from causal types to data realizations, we can calculate for each $lambda$ draw the probability of each data possibility given that particular $\lambda$ and the strategy. We then average across repeated $\lambda$ draws. Since $\lambda$'s are being drawn from our prior, we are automatically weighting more heavily those $\lambda$'s that we believe to be most likely.

**Posterior on estimate given the data**. For each data possibility, we can then use `CQtools` to ask what inference we would get from each data possibility, given whatever query we seek to answer, as well as the variance of that posterior. Examining the inferences from possible data-realizations, as we do below, can help us understand how the learning unfolds for different strategies. 

**Expected posterior variance under each strategy**. The quantity of ultimate interest is the posterior variance that we expect to end up with under each *strategy*. Calculating this expectation is now elementary as we have both the posterior variance arising from each data possibility and the probability of each data possibility (given our prior beliefs and the data already observed). The expected posterior variance is simply an average of the posterior variances under each data possibility, weighted by the probability of each data possibility. We can think of the expected gains from a strategy as the expected *reduction* in posterior variance arising from that strategy.

Complicated as the procedure might seem, it can all be done in code using a single command from the `CQtools` package.  


### Inferences from Strategies and Data Realizations

In Tables XXXXX, we examine how different process-tracing case-selection strategies will play out by looking at the inferences we would draw about different kinds of causal effects depending on the results of the process tracing.

Starting with the chain model, for the analyses displayed in Table XXXX, we assume that we have previously observed a positive $X,Y$ correlation in 6 cases (with 4 cases on the regression line, 2 cases off). Now we see what can happen for different queries when we implement each of the four case-selection strategies. The set of queries we are interested in are, what is the effect of $X$ on $Y$ for:

* any case in the population
* any $X=1, Y=1$ case in the population
* any $X=1, M=1, Y=1$ case in the population
* any $X=0, Y=1$ case in the population
* any $X=0, M=1, Y=1$ case in the population

#### Priors (Chain model)

In the top row, we see our pure prior on these queries before we have seen the $X,Y$ correlation --- i.e., based strictly on the DAG and flat priors over nodal types. The story here is straightforward. We start out believing that, for the $X \rightarrow M$ relationship, the population is evenly divided across the four types, $\theta^M_{00}, \theta^M_{01}, \theta^M_{10}, \theta^M_{11}$ (so 0.25 share of each). Likewise for $\theta^Y$ in relation to the $M \rightarrow Y$ relationship. We can get a positive $X,Y$ effect from a causal type involving $\theta^M_{01}$ and $\theta^Y_{01}$ or from a causal type involving $\theta^M_{10}$ and $\theta^Y_{10}$. Each of those causal types has a 0.125 probability, implying a 0.25 share of positive effects. A mirror-image of that logic gets us to a 0.25 share of negative effects, with the remaining 0.5 of the population comprising causal types that yield null effects (involving either a null effect at one or both steps in the chain). Thus, the expected effect of $X$ on $Y$ for a randomly drawn case from the population, given nothing but the model, is 0. 

Moving across the row, we see our prior beliefs for cases with more specific features. For cases with $X=1, Y=1$, our priors imply a positive effect 25\% of the time.^[This belief arises from a straightforward calculation involving the different combinations of nodal types that could generate $X=1, Y=1$ and their prior probabilities.] As we have flat priors over $M$'s type and its effects on $Y$, conditioning $M$'s value leaves this belief unaffected. By a parallel logic, the model and flat priors imply the belief that 0.25 of all $X=0, Y=1$ cases to involve a negative effect of $X$ on $Y$.

In the next row, we see how the initial set of 6 $X,Y$ observations shift our beliefs. In this model (as there is no confounding) a positive $X,Y$ correlation is evidence of a higher relative share of cases in which $X$ has a positive effect on $Y$. This implies a larger expected effect for a randomly selected case as well as for an $X=1, Y=1$ case since we now think there are more cases where $X$ has a positive effect on $Y$ than there are cases in which $Y$ is fixed at 1. (Again, conditioning on $M$'s value has no effect since we still have no information to shift us off our flat priors relating to $M$'s causes or effects.) Notably, however, our beliefs for an $X=0, Y=1$ case are unaffected by observing the positive correlation. To return for a moment to our handy $a,b,c,d$ typology, the reason is that the positive $X,Y$ correlation affects our beliefs about the share of $b$ types relative to $a,c,$ and $d$ types (with reference to the overall $X rightarrow Y$ effect), but it is uninformative about the relative shares of $a$ vs. $d$ types, the two types in contention for an $X=0, Y=1$ case. In the $X,Y$ data, the $X=0, Y=1$ case is equally likely to be an $a$ or a $d$, which is exactly what our prior belief was about a random $X=0, Y=1$ case.^[Put differently, the observed $X,Y$ correlation is equally consistent with all $X=0, Y=1$ cases being $a$'s, with them all being $d$'s, or with anything in between.]

#### $N=1$ strategies (Chain model)

Now, what can we learn by selecting a single case for process tracing, that is, the observation of the additional clue, $M$? We assume here, specifically, that we observe $M=1$ when we go looking for $M$. We can see in the third and fourth rows how our beliefs shift if we observe $M=1$, respectively, in one case *on* the regression line (an $X=Y=1$ case) or in one case *off* the regression line (an $X=1, Y=0$ case). We see that we cannot learn anything about the causal effect of $X$ on $Y$---whether unconditionally, conditioning on $X=Y=1$, or conditioning on $X=0, Y=1$---from observing $M=1$ in an $X=Y=1$ case. Nor can we learn about these quantities from process-tracing in a single off-the-line case. To see why, let's first take the on-the-line strategy. Not having observed $M$ previously, we still have flat priors over the nodal types governing $X$'s effect on $M$ and $M$'s effect on $Y$. That is to say, we still have no idea whether $X$'s positive effect on $Y$ (where present) more commonly operates through a chain of positive effects or a chain of negative effects. Thus, the observation $M=1$ in an $X=1, Y=1$ case is equally consistent with a positive $X \rightarrow Y$ (to the extent that effect operates via linked positive effects) and with no $X \rightarrow Y$ effect (to the extent positive effects operate through linked negative effects). Observing $M=1$ in an $X=1, Y=1$ case therefore tells us nothing about the causal effect in the typical case or in the typical $X=Y=1$ case. Similarly, we have no idea whether $X$'s negative effect on $Y$ (where present) operates through a positive-negative chain or a negative-positive chain, making $M=1$ in an $X=1, Y=0$ case equally consistent with a negative or null $X \rightarrow Y$ effect, yielding no information about whether a random $X=0, Y=1$ case has a negative or null effect. (By a similar logic, observing $M=1$ in the $X=1, Y=1$ case is uninformative about negative effects in an $X=0, Y=1$ case, and observing $M=1$ in an $X=1, Y=0$ case tells us nothing about positive effects in an $X=1, Y=1$ case.)

Interestingly, however, observing $M=1$ in a single case *can* tell us something about effects in *another* case in which $M=1$. Recall that the $X,Y$ pattern tells us that (with respect to the $X \rightarrow Y$ relationship) there are relatively more $b$'s than other types in the population. This means, in turn, that an $X=1, Y=1$ case is more likely to be a $b$ than a $d$. If we now observe $M=1$ in an $X=1, Y=1$---a case we think is probably a $b$---we now have reason to believe that positive $X \rightarrow Y$ effects operate through a positive $X \rightarrow M$ effect followed by a positive $M \rightarrow Y$ effect. (And had we seen $M=0$ in this case, we'd believe positive effects more likely operate through a negative effect followed by a positive effect.) Put differently, we now believe that $M=1$ in an $X=1, Y=1$ case is a clue that that case is likely a $b$ type---and our belief about the causal effect in an $X=1, M=1, Y=1$ case shifts upward. It shifts upward in two senses. 

Thus, while we do not learn about the overall population from an N=1 process-tracing exercise, process-tracing in that single case does make $M$ *informative* about *other* cases. We can see this by comparing our beliefs about an $X=1, M=1, Y=1$ case in the row in which we have only observed the $X,Y$ correlation to our beliefs about that same kind of case in the row in which we have process-traced in an on-the-line case. While in the first instance conditioning on $M=1$ has no impact on the estimated causal effect (because we haven't previously learned about $M$), it does have an impact in the latter instance.

We also see that our belief about the causal effect in an $X=0, M=1, Y=1$ case shifts upward. Why? Updating toward the belief that positive effects --- which we believe to be relatively common --- more often operate via linked positive positive effects implies that $X$ more often has a positive than a negative effect on $M$. Yet for an $X=0, M=1, Y=1$ to be a case with a negative causal effect, $X$'s effect on $M$ would have to be negative. Thus, interestingly, evidence from *on* the regression line can inform inferences about cases *off* the regression line via information about mediating processes.

FLAG: Haven't figured out why seeing M=1 in an X=1, Y=0 case reduces the estimated causal effect for an X=1, M=1, Y=1 case relative to just seeing the X,Y pattern. If we see M=1 in this off-line case, we have evidence in favor of X having a positive effect on M, and M having a negative effect on Y. So this should generally speak against positive effects in  X=1, M=1, Y=1 cases. But if we'd seen M=0, wouldn't this have the same effect? It would suggest negative X->M effects and positive M->Y effects, which would also speak against a positive effect in an  X=1, M=1, Y=1 case. So how does observing M help?

We see even stronger learning about an off-the-line $X=0, M=1, Y=1$, however, if we process trace off-the-line. The observation of $M=1$ in an $X=1, Y=0$ case counts as evidence against the operation of *both* steps in the causal chain through which $X$ could have a negative effect on $Y$ in an $X=0, M=1, Y=1$ case: it suggests that $X$'s effect on $M$ is more likely positive than negative *and* that $M$'s effect on $Y$ is more likely negative---precisely the opposite of what would have to be true for a negative effect to emerge in this kind of case. Thus, we now think this kind of case is less likely to involve a causal effect.

#### $N=2$ Strategies (Chain model)

Next, we consider the process tracing of *two* of our cases. Now, because we are observing $M$ in two cases, we can learn from the variation in $M$ across these cases---or, more specifically, from its covariation with $X$ aand with $Y$. In other words, there is a critical difference between process tracing one case and process tracing two cases. When we only process trace one case in a setup like this one, we do not learn about causal effects in the cases we process-trace because we have no information about intermediate causal effects (e.g., whether they are more likely positive or negative). Thus, while we have pointed out that observing $M$ within a case can lend $M$ probative value for *other* cases in which we might choose to observe $M$, $M$ has no probative value *for* that one case. In contrast, if we observe $M$ in two or more cases, we *do* learn about causal effects for those cases because of the leverage provided by observing covariation between the process-tracing clue and other variables.

We consider first, in the 5th row, a situation in which we look at $M$ in two on-the-line cases, an $X=Y=0$ case and an $X=Y=1$ case, and where we observe $M=0$ in the first case and $M=1$ in the second case. What we are seeing, then, is that $M$ covaries with both $X$ and $Y$. Seeing that the mediator covaries across cases with both the cause and the outcome is evidence of a causal effect of $X$ on $Y$. Since the observed pattern covariation is consistent with a positive effect (linked positive effects), our belief about the unconditional effect of $X$ on $Y$ goes up. For the same reason, so too does our belief about the causal effect conditional on $X=1, Y=1$. 

We see an even bigger boost in our estimate of the effect in an $X=1, Y=1$ case conditional on $M$ also being 1. Two things are generating this upward updating. First, just as for the unconditional estimate and the estimate conditional on $X=1, Y=1$, the process tracing generates a stronger belief that $X$ generally has a positive effect on $Y$. Second, as for the $N=1$ process tracing, seeing $M=1$ in two on-the-line cases (which we already believe are mostly positive-effect cases) makes $M$ informative as a clue in favor of a positive effect---indeed, more informative than if we'd only seen $M=1$ in one on-the-line case.

Moving across the columns, we also see that process-tracing the on-the-line cases and finding evidence consistent with positive effects in those cases leads us to update in favor of *negative* effects as well, conditional on $X=0, Y=1$ (including conditional on $M=1$). Why is that? 

Our answer is somewhat speculative, but we think there are two things going on. One is simply that the correlations between $M$ and $X$ and between $M$ and $Y$ across the two process-traced cases constitutes evidence of *effects*, period, at the intervening steps---a necessary condition for *any* $X \rightarrow Y$ effect, positive or negative. So, for an $X=0, Y=1$ case, a stronger belief that $X$ has some effect on $M$ and that $M$ has some effect on $Y$ implies a strong belief in a negative $X \rightarrow Y$ effect. 

The second thing that is probably going on is constraint imposed by prior beliefs. Having seen the $X,Y$ data, we have previously formed a belief that the *average* effect of $X$ on $Y$ is 0.05. When we find evidence in favor of positive effects, we update to a set of beliefs that are a compromise between the new data and prior beliefs. On the one hand, we upwardly update on the unconditional $X \rightarrow Y$ effect. On the other hand, our beliefs are constrained from moving *too* far away from an unconditional effect of 0.05. Recall that, in a binary setting, the average effect in a population is the share of cases with positive effects minus the share with negative effects. Thus, the conservative influence of our priors means that, as we upwardly update on positive effects, we also upwardly update our beliefs about the prevalence of *negative* effects, though not by as much, with the result that our beliefs about the unconditional effect do not shift as much as they would if we had had flat priors. By anchoring our beliefs about average effects, then, the prior $X,Y$ evidence means that any evidence in favor of positive effects also constitutes evidence in favor of negative effects. 

To put this point another way, while we learn *directly* about positive effects from process-tracing the on-the-line cases since these are cases in which positive effects can possible occur. And we then learn *indirectly* about negative effects from these cases *via* the prior-imposed constraint on the average effect.

Now, supppose we chose the same case-selection strategy---we examine two on-the-line cases, from opposite cells---but instead of observing $M$ covary with $X$ and $Y$, $M=1$ in both cases. We see the results in the 6th row. This data-realization, under the same strategy, represents clear evidence *against* positive effects, while also counting against *any* effects. We have observed that $M$ is unresponsive to changes in $X$, and that $Y$ varies without $M$ varying---null effects at both stages of the chain. Now we see that our posterior on the unconditional effect drops to 0, and we see a sharp movement toward zero conditional on $X=1, Y=1$, representing our updated belief that there are few positive effects. Interestingly, whereas conditioning additionally on $M=1$ generated in a *higher* estimated effect when we observed $M$ correlated with $X$ and $Y$, it generates a *lower* estimated effect when we observe $M=1$ in both cases. One reason is likely that we have updated our belief about $\theta^M$ such that we now believe that $X=1, M=1$ cases are more likely to be $\theta^M_{11}$'s (no effect) than $\theta^M_{01}$'s. We have not similarly updated with respect to $X=1, M=0$ cases and the $\theta^M_{00}$'s than $\theta^M_{10}$ shares. Thus, for an $X=1,Y=1$ case, additionally specifying that $M=1$ implies an even higher likelihood that the case contains a null effect. 

For $X=0, Y=1$ cases, we see similar movements toward zero, reflecting the fact the $M$ pattern speaks against any effects. And, again, the movement is stronger when we condition on $M=1$.

Finally, we consider an off-the-line $N=2$ strategy and two possible data-realizations under that strategy. We select for process tracing one $X=1, Y=0$ case and one $X=0, Y=1$ case. The two possible data-realizations we consisder are that $M$ varies positively with $X$ across the two cases (7th row) or that $M=1$ in both cases (8th row). One thing we can observe immediately is that we get less leverage from this strategy: across all columns, the difference in our posterior beliefs between the two data-realizations is smaller than it is for the two data-realizations of the on-the-line strategy. For instance, while we see $0.08$ swing in the unconditional effect under the on-the-line strategy, we see only a $0.02$ swing under the off-the-line strategy.

Unpacking this further, we know that we can learn directly about *negative* $X \rightarrow Y$ effects through process-tracing off-the-line cases. If we see $M$ covary positively with $X$ across these two cases---which also means covarying negatively with $Y$---we have found evidence in favor of the operation of negative $X \rightarrow Y$ effects. Hence the movement in our posterior away from 0 when we condition on $X=0, Y=1$. We also see updating away from zero for positive effects (conditioning on $X=1, Y=1$), though by a smaller amount (using the $X,Y$-data only as our baseline). Again, some of this movement is likely attributable to the constraint imposed by the prior generated from the $X,Y$ data. Some of this updating is also likely attributable to the fact that we have found evidence against null intermediate effects.

Curiously, under this strategy and data-realization, our beliefs about effects conditional on $X=0, M=1, Y=1$ move in the *opposite* direction from our beliefs conditional just on $X=0,Y=1$. Why might this be? This is likely because of the way in which the process-tracing has lent probative value to $M$. The pattern of $M$ data is most consistent with any negative $X \rightarrow Y$ effect operating via a *positive* $X \rightarrow M$ effect followed by a *negative* $M \rightarrow Y$ effect. Yet the intermediate effects would have to have the opposite sign for $X=0, M=1, Y=1$ to be consistent with an $X \rightarrow Y$ effect. In other words, $M=1$ rules out the likeliest way in which an $X=0, Y=1$ case could contain an effect. 

If on the other hand we observe $M=1$ in both off-the-line cases, we have found "direct" evidence against the operation of negative effects. Hence the the strong movement of our posterior toward 0 for effects conditional on $X=0, Y=1$ and, even more so, if also conditioning on $M=1$ (since we now have evidence that $M=1$ cases are more likely $\theta^M_{11}$'s than $\theta^M_{01}$'s). Likewise, we see upward movement in our posterior on the unconditional effect and, for reasons discussed earlier, in our posterior on positive effects (conditioning on $X=1, Y=1$).

The remaining puzzle is why the learning from the off-the-line strategy is weaker than the learning from the on-the-line strategy. Both strategies generate symmetrical opportunities to observe an $M$ pattern that is consistent or inconsistent with causal effects. We believe there are two reasons, both relating to the fact that we learn "directly" about positive effects from the on-the-line strategy (and "indirectly" about negative effects) and that we learn "directly" about negative effects (and "indirectly" about positive effects) from the off-the-line strategy. 

The first, more mechanical reason is that we start, given the $X,Y$ data, with a higher and higher-variance posterior on the prevalence of positive effects as compared to negative effects. So there is simply more scope for the data to influence our beliefs about positive effects, and the on-the-line cases are those in which we can learn "directly" about positive effects. Note, for instance, the $0.17$ differential in beliefs about effects conditional on $X=1, Y=1$ depending on what we observe under the on-the-line strategy, as compared to the $0.1$ differential for effects conditional on $X=0, Y=1$ under the off-the-line strategy.

The second reason is prevalence: given the $X,Y$ correlation, we believe that there are more cases out there that potentially contain positive effects than there are cases that potentially contain negative effects. Thus, *whatever* it is we learn about positive effects is going to have a bigger impact on our beliefs about the population as a whole than is whatever we learn about negative effects. Since on-the-line case yield the most direct learning about positive effects, they are therefore the most informative about the population.

Note, importantly, that this finding is consistent with recommendations in the literature to investigate on-the-line cases (@Lieberman2005nested, @goertz2017multimethod).^[We set aside here Lieberman's recommendation to choose off-the-line cases for discovery-oriented "model-building" because we see that enterprise as distinct from the estimation endeavor to which our analysis is oriented.] Yet, the reason is very different. Lieberman and Goertz emphasize the importance of examining cases where the theorized mechanism connecting $X$ to $Y$ might potentially play out, arguing that this would be in the set of cases that conform to theory in their $X,Y$ values. We would emphasize that, in the causal-model framework, one *can* learn about both overall effects and intermediate effects from off-the-line cases. The reason why on-the-line cases is simply that they are more *representative* about the universe about which we aim to learn (assuming we want to learn about the population from which the $X,Y$ pattern is drawn).  



## Logic of strategy comparison

```{r selectioncode, echo = FALSE}

# We define a model
model <- make_model("X->M->Y")  %>%
 set_restrictions(c("(Y[M=1]<Y[M=0])", "(M[X=1]<M[X=0])")) %>%
 set_parameter_matrix() %>%
 set_parameters(type = "flat")

# We imagine some preexisting data we have observed
data  <-  data.frame(X = c(0,0,0,1,1,1), M = NA, Y = c(0,0,1,0,1,1))
observed <-  collapse_data(data, model, drop_family = TRUE)

# We can then imagine what data we might observe if we examine M inside some subset of cases
A_on_regression_line <- 
  make_possible_data(
    model, 
    observed = observed, vars = "M", 
    N = c(1,1), 
    condition = c("X==1 & Y==1", "X==0 & Y==0"), 
    prefix = "A")
names(A_on_regression_line)[-c(1:2)] <- c("A1", "A2", "A3", "A4")

B_off_regression_line <- 
  make_possible_data(model, observed = observed, vars = "M", N = c(1,1), 
                     condition = c("X==1 & Y==0", "X==0 & Y==1"), prefix = "B")
names(B_off_regression_line)[-c(1:2)] <- c("B1", "B2", "B3", "B4")

C_X1Y1_only <- 
  make_possible_data(model, observed = observed, vars = "M", N = 2, 
                     condition = c("X==1 & Y==1"), prefix = "C")
names(C_X1Y1_only)[-c(1:2)] <- c("C1", "C2", "C3")

```

Consider a situation in which one has access to $X,Y$ data on just six cases of the form:

```{r observed, echo = FALSE}
kable(observed)
```

We want to examine data on $M$ for two of these cases and are wondering about what strategy we should use to select the cases. We are considering three strategies:

* Strategy $A$ chooses two cases on the regression line (one in the $X=Y=0$ cell and one in the $X=Y=1$ cell)
* Strategy $B$ chooses off the regression line 
* Strategy $C$ selects on the dependent variable -- choosing cases with $X=1, Y = 1$.

Different strategies yield different possible types of data. Each one of these is likely to arise with a different probability and is associated with a different inference. Drawing on the prior beliefs embedded in a causal model, we can thus assess the *expected* posterior variance associated with each strategy. 


```{r dtypes, echo = FALSE}
# Combine data types
df <- A_on_regression_line[,-2] %>% 
  merge(B_off_regression_line[,-2], by = "event", all = TRUE) %>%
  merge(C_X1Y1_only[,-2], by = "event", all = TRUE) %>%
  dplyr:::mutate_if(is.numeric, ~replace(., is.na(.), 0))

# table(duplicated(t(df)))
```

For each of these possible strategies we can assess the posterior that we would obtain for each type of data we might observe.

```{r illustration, echo = FALSE}

if(do_diagnosis){
  
 if(!exists("fit")) fit  <- update_model(make_model("X->Y"), keep_fit = TRUE)
 
 strategies <- 
   list(A_on_regression_line, B_off_regression_line, C_X1Y1_only)

 	write_rds(strategies, "saved/illustration_strategies.rds")
  
 
 estimates_dbs <- lapply(strategies, function(s) 
    make_estimates_database(model, observed, possible_data = s[, -2], queries = "Y[X=1]-Y[X=0]")
    )
    
	write_rds(estimates_dbs, "saved/illustration_estimates.rds")

# The reference model is the original model updated with the XY data we've already seen.
 reference_model <- update_model(model, data, stan_model = fit)
 
	write_rds(reference_model, "saved/reference_XMY_updated_model.rds")

 probabilities <- lapply(strategies, function(s) 
    average_data_probabilities(reference_model, s, using = "posteriors", sims = 3000)
    )
   
	write_rds(probabilities, "saved/illustration_probabilities.rds")

}

  reference_model <- read_rds("saved/reference_XMY_updated_model.rds")
	probabilities   <- read_rds("saved/illustration_probabilities.rds")
  strategies      <- read_rds("saved/illustration_strategies.rds")
  estimates_db    <- read_rds("saved/illustration_estimates.rds")
  
  digits <- 3

  df <- rbind(df, 
          c("Probability", round(unlist(probabilities),digits)),
          c("Posterior mean",   (unlist(lapply(estimates_db, function(j)   unlist(unlist(lapply(j, function(k) k$mean))))))),
          c("Posterior variance",   round(unlist(lapply(estimates_db, function(j)   unlist(unlist(lapply(j, function(k) k$mean^2))))), digits)))
```

The data possibilities, probabilities of each, and inferences given each, are shown in Table \@ref(tab:chselillustration).
  
```{r chselillustration, echo = FALSE}  
kable(df, caption = "Each column shows a possible distribution of data that can be generated from a given strategy. We calculate the probability of each data possibility, given the data seen so far, and the posterior variance associated with each one.", digits = 2)
```

Each of the first two strategies generates one of four different data patterns. The third data strategy generates up to three data patterns. None of these data patterns overlap across strategies. 
From the calculated  probability of each data type, given the data seen so far, and the posterior variance given each data realization, the implied *expected* variance is  easily calculated. These are summarized below:

```{r exppostvar, echo = FALSE}
expected_posterior_var <- sapply(1:3, function(i) 
  sum(probabilities[[i]]*unlist(lapply(estimates_db[[i]], function(x) x$sd^2)))
 )
kable(data.frame(Strategy = c("Online", "Offline", "X=1, Y=1"), Variance = expected_posterior_var), digits = 3)
```


In this example, we see that we would expect to be better off---in the sense of having less posterior uncertainty---by focusing her process-tracing efforts where a greater share of the population of cases lies: on the regression line.  

Why is this?


```{r test, include = FALSE}
# source("_packages_used.R")

model <- make_model("X -> M -> Y")
  
gen_observed_data <- function(n00=1, n01=1, n10=1, n11=1,times = 1) {
  data.frame(X = c(0,0,1,1), Y =c(0,1,0,1), M = NA, ns = c(n00, n01, n10, n11)*times) %>%
    tidyr::uncount(ns) %>%
    collapse_data(model, drop_family = TRUE)}


# Data strategies

data_strats <- function(N = 4) {
  if(N%%2 != 0) stop("even Ns only please")
  list(
    all_on = list(N=c(N,N)/2, withins = TRUE, vars = "M",
                  conditions = list("X==0 & Y==0", "X==1 & Y==1")),
    all_off = list(N=c(N,N)/2, withins = TRUE, vars = "M",
                   conditions = list("X==0 & Y==1", "X==1 & Y==0")),
    all_x1 = list(N=c(N,N)/2, withins = TRUE, vars = "M",
                   conditions = list("X==1 & Y==0", "X==1 & Y==1")),
    all_y1 = list(N=c(N,N)/2, withins = TRUE, vars = "M",
                   conditions = list("X==0 & Y==1", "X==1 & Y==1"))
  )
}

# Queries

# if(!exists("fit")) fit <- fitted_model()

# Diagnosis function 

if(do_diagnosis){
test_corr <- 
  diagnose_strategies(
    analysis_model = model,
    data_strategies = data_strats(2),
    given = gen_observed_data(6,2,2,6),
    queries = list(ATE = "Y[X=1]-Y[X=0]", 
                   a = "Y[X=1]<Y[X=0]", 
                   b = "Y[X=1]>Y[X=0]", 
                   c = "(Y[X=1]==0) & (Y[X=0]==0)", 
                   d = "(Y[X=1]==1) & (Y[X=0]==1)",
                   a1 = "M[X=1]<M[X=0]", 
                   b1 = "M[X=1]>M[X=0]", 
                   c1 = "(M[X=1]==0) & (M[X=0]==0)", 
                   d1 = "(M[X=1]==1) & (M[X=0]==1)",
                   a2 = "Y[M=1]<Y[M=0]", 
                   b2 = "Y[M=1]>Y[M=0]", 
                   c2 = "(Y[M=1]==0) & (Y[M=0]==0)", 
                   d2 = "(Y[M=1]==1) & (Y[M=0]==1)"
                   ),
    subsets = list(TRUE, TRUE,  TRUE,  TRUE),
    sims = 8000,
    fit = fit, 
    refresh = 1000,
    iter = 8000)

write_rds(test_corr, "saved/test_corr.rds")

test_flat <- 
  diagnose_strategies(
    analysis_model = model,
    data_strategies = data_strats(2),
    given = gen_observed_data(4,4,4,4),
    queries = list(ATE = "Y[X=1]-Y[X=0]", 
                   a = "Y[X=1]<Y[X=0]", 
                   b = "Y[X=1]>Y[X=0]", 
                   c = "(Y[X=1]==0) & (Y[X=0]==0)", 
                   d = "(Y[X=1]==1) & (Y[X=0]==1)",
                   a1 = "M[X=1]<M[X=0]", 
                   b1 = "M[X=1]>M[X=0]", 
                   c1 = "(M[X=1]==0) & (M[X=0]==0)", 
                   d1 = "(M[X=1]==1) & (M[X=0]==1)",
                   a2 = "Y[M=1]<Y[M=0]", 
                   b2 = "Y[M=1]>Y[M=0]", 
                   c2 = "(Y[M=1]==0) & (Y[M=0]==0)", 
                   d2 = "(Y[M=1]==1) & (Y[M=0]==1)"
                   ),
    subsets = list(TRUE, TRUE,  TRUE,  TRUE),
    sims = 8000, fit = fit, refresh = 1000, iter = 8000)


write_rds(test_flat, "saved/test_flat.rds") 
}

test_corr <- read_rds("saved/test_corr.rds") 
test_flat <- read_rds("saved/test_flat.rds") 

# key thing to note is for $all_x1[[1]] a2 = b2
# and for $all_x1[[2]] a1 = b1 
# so in both cases shareas of a and b remains contstant


```

For intuition consider first the case with flat prior data:

* if we examine two cases in different quadrants *on the diagonal* (one $X=Y=0$ case and one $X=Y=1$ case) :
  * if  we find that $M$ is the same in the two cases then we  increase our belief that $X$ has no effect at all on $Y$ and reduce our confidence that $X$ had a positive or a negative effect on $Y$. Since we are looking on the regression line however, our confidence that $X$ had a *positive* effect on $Y$ is more strongly reduced, producing a  posterior centered on a small negative effect. 
  * Conversely if we see that $M$ is different in the two cases, then we have a correlation of the same sign between both $X$ and $M$ and between $M$ and $Y$  cases. We increase our confidence that $X$ mattered for $Y$ in general an din particular that it had a positive effect, resulting in a posterior centered on a small positive ATE.

* if we examine two cases in different quadrants *off the diagonal* (one $X=0, Y=1$ case and one $X=1, Y=0$ case) :
  * if  we find that $M$ is the same in the two cases then we  increase our belief that $X$ has no effect at all on $Y$ and reduce our confidence that $X$ had a negative effect  $Y$ (producing a  posterior centered on a small positive effect). 
  * Conversely if we see that $M$ is different in the two cases, then we have a correlation (of different signs) between both $X$ and $M$ and between $M$ and $Y$  cases. We increase our confidence that $X$ mattered for $Y$ in general and in particular that it had a negative effect, resulting in a posterior centered on a small negative ATE.

* If we examine data conditioning on the value of $X$ but with variation on $Y$ (for instance $X=1, Y=0$ and   $X=1, Y=1$) 
data patterns do not discriminate between X having a positive effect on Y an d X having a negative effect on Y. However variation in $M$ is more consistent with $M$ being responsive to $X$ and thus the chances that $X$ matters, positively or negatively, overall. In teh case with flat data this changes beliefs on positve and negative effects but not the difference between them. The ate then remains unchanged. FLAG work though intuition  more

* If we examine data conditioning on the value of $Y$ but with variation on $X$ (for instance $X=0, Y=1$ and   $X=1, Y=1$) 
data patterns do not discriminate between X having a positive effect on Y an d X having a negative effect on Y. However variation in $M$ is more consistent with $M$ being responsive to $X$ and thus the chances that $X$ matters, positively or negatively, overall. In teh case with flat data this changes beliefs on positve and negative effects but not the difference between them. The ate then remains unchanged.

For correlated data similar logics apply, but the effects are stronger for evidence on the regression line. The reason is that given correlated data we believe there are more units with positive effects than negative effects. When we find evidence against causal relations on the regression line that reduces our confidence for types with positive effects more than for types with negative effects and teh difference between teh shares with positive effects and negative eeffects smaller due to the fact that the beliefs in the shares with negative effects is not so strongly reduced. Wehen we find evidence for causal relations this magnifies teh difference between beliefs in shares positive and shares negatives; these effects are magnified however since our confidence for the positive effects change more than for the negative effects, since we are examining cases on the regression line. Conversely when examining cases of teh regression line, the two forces offset each other. 



```{r cinfer, include = FALSE}
# Make sense given process tracing approach

long_data <-  expand_data(gen_observed_data(4,2,2,4), model)

if(do_diagnosis){
  updated <- update_model(model, long_data, iter = 14000, chains = 12) %>%
  set_parameters(param_type = "posterior_mean")
  write_rds(updated, "saved/ch13_longupdated.rds") 
}
updated <- read_rds("saved/ch13_longupdated.rds") 

get_parameters(updated)

conditional_inferences(updated, query = "Y[X=1]!=Y[X=0]", given = "!is.na(X) & !is.na(Y) & !is.na(M)") %>%
  arrange(X, Y, M)
```






## Explorations

Using the same basic logic we can explore performance across a wider range of strategies that might be employed under different conditions.

The qualitative case-selection literature has identified a range of possible strategies for choosing cases for in-depth analysis. These include, for instance, selecting for variation on $X$ (KKV), selecting for variation on $Y$, selecting cases on the regression line (Seawright and Gerring, Lieberman), and selecting off the regression line (Seawright and Gerring, Seawright 2017, Lieberman).[These authors view selection off the regression line as best for arriving at inductive insight. We address this strategy primarily to show the contrast with the on-the-line strategy.] While we have not seen it advocated elsewhere, we might add to this list the strategy of selecting cases that are representative in their $X,Y$ values of the larger set of cases from which we are selecting. (FLAG: insert proper citations). While it is difficult to clearly distinguish these strategies from each other with a small initial set of $X, Y$ cases, we can do so readily if we start with a larger set of $X,Y$ cases.

In Table \@ref(caseselectlots), we show the results of diagnoses of each of these five classes of strategies, assuming that we will be process-tracing 6 cases. In each diagnosis we start with 500 $X, Y$ cases, with 100 $X=Y=0$ cases, 200 $X=Y=1$ cases, 130 $X=0, Y=1$ cases, and 70 $X=1, Y=0$ cases. The regression line here represents a positive association. We work with the same model and priors as above. Importantly, this  means that the results we show here are not *general* evaluations of these strategies, but contingent on this particular model and set of priors. And that is precisely our point: optimal case-selection will always hinge on our model, a claim that we demonstrate further below.


### Procedure

The procedure requires as inputs  (i) a causal model, (ii) any data we have already observed, and (iii) the causal query we seek to answer. 

The general intuition is that we can use the causal model and any previously observed data to estimate what observations we are more or less likely to make under a given case-selection strategy, and then figure out how far off from the (under the model) true estimand we can expect to be under the strategy, given whatever causal question we seek to answer. 

Suppose that we want to estimate the average treatment effect of $X$ on $Y$ in a population and have initially observed $X$ and $Y$ data on a set of cases. 

Suppose that we are now considering gathering process-tracing evidence for one of these cases to inform our estimate of the ATE. There are many different case-selection strategies we might pursue, and each of these can give rise to different possible data and thus to different possible conclusions. What should we do?

**DAG**. We start, as always, with a DAG representing our beliefs about which variables we believe to be direct causes of other variables. For the current illustration, suppose that we are operating with a simple mediation model, $X \rightarrow M \rightarrow Y$. 

**Given data.** If we have already observed something in a set of cases, we can use this information to condition our strategy for searching for further information. For instance, if we have observed $X$'s and $Y$'s value in a set of cases, we might select cases for process tracing based on their values of $X$ and $Y$. Further, what we have already observed in the cases may constrain what possible data we could end up with once we have collected the additional (process tracing) data. 

**Priors**. As when conducting mixed-method inference, we can set qualitative restrictions and/or differential quantitative weights on the (possibly conditional) nodal types in the model. And we can indicate our uncertainty over the latter, by setting the $\alpha$ parameters of the relevant Dirichlet distributions. For the current example, let us define restrictions at both the $M$ and $Y$ nodes, positing beliefs that $X$ never has a negative effect on $M$ and that $M$ never has a negative effect on $Y$. Let us further assume that we have flat priors over the remaining nodal types and posit similar assignment propensities for all types (no unobserved confounding).

**Query**. We define our query. This might, for instance, be the share of cases in the population in which $X$ has a positive effect on $Y$; or it might be $X$'s average effect on $Y$. We can use the general procedure to identify case-selection strategies for any causal query that can be defined on a DAG. And, importantly, the optimal case-selection strategy may depend on the query. For instance, the best case-selection strategy for estimating the average causal effect of $X$ on $Y$ may not be the same as the best strategy for figuring out for what proportion of the population $X$ has a positive effect on $Y$.

**Define one or more strategies**. A strategy is defined, generically, as the search for data on a given set of *nodes*, in a given *number* of cases randomly selected *conditional* on some information we already have about potential cases. Let us assume here that our strategy will involve uncovering $M$'s value in 1 case---but we are wondering how to choose this case. Consider four possible strategies, conditional on the $X$ and $Y$ values that we already know. We could do process tracing on a randomly selected $X=1, Y=1$ case, a randomly selected $X=0, Y=0$ case, the $X=1, Y=0$ case, or the $X=0, Y=1$ case. We itemize this set of possible strategies in the first column in Table \@ref(tab:caseselect1).

**Possible data**. For each strategy, there are multiple possible sets of data that we could end up observing. In particular, the data we could end up with will be the $X,Y$ patterns we have already observed plus *either* $M=0$ *or* $M=1$ in the case that our strategy leads us to select for process tracing. We represent the data possibilities (showing just the possible $M$ values) in the second column in Table \@ref(tab:caseselect1). Thus, for instance, for a strategy in which we choose a random $X=1, Y=1$ case, we could end up observing the initial $X,Y$ pattern plus $M=0$ in one of the $X=1, Y=1$ cases, or the initial $X,Y$ pattern plus $M=1$ in one of the $X=1, Y=1$ cases. 

**Probability of the data**. We now calculate a probability of each possible data realization, given the model and the data (the $X$'s and $Y$'s) that we have already observed. In practice, we do this in `CQtools` via simulation. Starting with the model together with our priors, we update our beliefs about $\lambda$ based on the initial $X,Y$ data. This posterior now represents our *prior* for the purposes of the process tracing; it represents what we believe about causal-type share allocations in the population, having seen the $X,Y$ data only. We then use this posterior to draw a series of $\lambda$ values. 

Given that the ambiguity matrix gives us the mapping from causal types to data realizations, we can calculate for each $lambda$ draw the probability of each data possibility given that particular $\lambda$ and the strategy. We then average across repeated $\lambda$ draws. (Since $\lambda$'s are being drawn from our prior, we are automatically weighting more heavily those $\lambda$'s that we believe to be most likely.) We show the data probabilities in the third column of Table \@ref(tab:caseselect1): one probability for each data-possibility given each strategy.

**Posterior variance on estimate given the data**. For each data possibility, we can then use `CQtools` to ask what inference we would get from each data possibility, given whatever query we seek to answer. What we are in fact interested in for the purposes of case selection is the *variance* of the posterior. We indicate in the fourth column of Table \@ref(tab:caseselect1) the posterior variance for each possible data realization.

**Expected posterior variance under each strategy**. The quantity of ultimate interest is the posterior variance that we expect to end up with under each *strategy*. Calculating this expectation is now elementary as we have both the posterior variance arising from each data possibility and the probability of each data possibility (given our prior beliefs and the data already observed). The expected posterior variance is simply an average of the posterior variances under each data possibility, weighted by the probability of each data possibility. The final column provides the expected posterior variances, one for each strategy.


Complicated as the procedure might seem, it can all be done in code using a single command from the `CQtools` package. 


```{r showx1y1diagnosestrategies, eval=FALSE, include = TRUE}

diagnose_strategies(
    
   analysis_model = model,
   
   observed = observed,
   
   data_strategies = list(
     A_online  = list(N=c(1,1), within = TRUE, vars = "M", 
                       conditions = c("X==1 & Y==1", "X==0 & Y==0")),
     B_offline = list(N=c(1,1), within = TRUE, vars = "M", 
                       conditions = c("X==1 & Y==0", "X==0 & Y==1")),
     C_X1Y1    = list(N=2, within = TRUE, vars = "M", 
                       conditions = c("X==1 & Y==1"))),
   
   queries = "Y[X=1] - Y[X=0]"
   
   )

```

```{r ch13diagnosis387, echo = FALSE, message = FALSE}

if(do_diagnosis){
  
diagnosis <- 
  
  diagnose_strategies(
    
   analysis_model = model,
   
   observed = observed,
   
   data_strategies = list(
     A_online =  list(N=c(1,1), withins = TRUE, vars = "M", 
                       conditions = c("X==1 & Y==1", "X==0 & Y==0")),
     B_offline = list(N=c(1,1), withins = TRUE, vars = "M", 
                       conditions = c("X==1 & Y==0", "X==0 & Y==1")),
     C_X1Y1 = list(N=2, withins = TRUE, vars = "M", 
                       conditions = c("X==1 & Y==1")),
     D_random = list(N=2, withins = TRUE, vars = "M", conditions = TRUE)),
   
   queries = "Y[X=1] - Y[X=0]",
   
   sims = 1000
   
   )

write_rds(diagnosis, "saved/ch13diagnosis.rds")

}

diagnosis <- read_rds("saved/ch13diagnosis.rds")
kable(diagnosis$diagnoses_df[, c(1, 4:7)], digits = 4, caption = "The estimand and the expected estimate are the same under each data strategy. ")
```



```{r diagprep, echo = FALSE}
# Models

models <- list(
  chain_model     = make_model("X->M->Y"),

  base_model     = make_model("X->M->Y<-X"),
  
  restricted_model = make_model("X->M->Y<-X") %>%
    set_restrictions(c(decreasing("X", "Y"), 
                       decreasing("X", "M"),
                       decreasing("M", "Y"))),
  
  obs_confound   = make_model("M->X->Y<-M"),
  
  unobs_confound = make_model("X->M->Y<-X; X<->Y"))

gen_observed_data <- function(n00=1, n01=1, n10=1, n11=1,times = 1) {
  data.frame(X = c(0,0,1,1), Y =c(0,1,0,1), M = NA, ns = c(n00, n01, n10, n11)*times) %>%
    tidyr::uncount(ns) %>%
    collapse_data(models$base_model, drop_family = TRUE)}

# observed_data
observed_data <- list(
  regr   = gen_observed_data(12,4,4,12),
  flat   = gen_observed_data(8,8,8,8),
  suff   = gen_observed_data(8,8,4,12),
  necs   = gen_observed_data(12,4,8,8),
  conc   = gen_observed_data(8,4,4,16))


# Data strategies

data_strats <- function(N = 4) {
  if(N%%2 != 0) stop("even Ns only please")
  
  
  out <- list(
    off_01 = list(N=N, withins = TRUE, vars = "M", conditions = "X==0 & Y==1"),
    off_10 = list(N=N, withins = TRUE, vars = "M", conditions = "X==1 & Y==0"),
    on_00  = list(N=N, withins = TRUE, vars = "M", conditions = "X==0 & Y==0"),
    on_11  = list(N=N, withins = TRUE, vars = "M", conditions = "X==1 & Y==1"),
    
    all_on = list(N=c(N,N)/2, withins = TRUE, vars = "M",
                  conditions = list("X==0 & Y==0", "X==1 & Y==1")),
    all_off = list(N=c(N,N)/2, withins = TRUE, vars = "M",
                   conditions = list("X==0 & Y==1", "X==1 & Y==0")),
    all_y0 = list(N=c(N,N)/2, withins = TRUE, vars = "M",
                  conditions = list("X==0 & Y==0", "X==1 & Y==0")),
    all_y1 = list(N=c(N,N)/2, withins = TRUE, vars = "M",
                  conditions = list("X==0 & Y==1", "X==1 & Y==1")),
    all_x0 = list(N=c(N,N)/2, withins = TRUE, vars = "M",
                  conditions = list("X==0 & Y==0", "X==0 & Y==1")),
    all_x1 = list(N=c(N,N)/2, withins = TRUE, vars = "M",
                  conditions = list("X==1 & Y==0", "X==1 & Y==1")))
  
  if(N==2){message("N<4, fully spread strategy not possible"); return(out)}
  
  c(out, list(spread = list(N=c(N,N,N,N)/4, withins = TRUE, vars = "M",
                       conditions = list("X==0 & Y==0", "X==0 & Y==1","X==1 & Y==0", "X==1 & Y==1"))
  ))
}

# Queries

queries <- list(
  ATE = "Y[X=1]-Y[X=0]", 
  ProbPos = "Y[X=1]>Y[X=0]",
  via_M = "(Y[X=1] > Y[X=0]) & (M[X=1] > M[X=0]) & (Y[M=M[X=1]] > Y[M=M[X=0]]) & (Y[X=1, M=M[X=1]] == Y[X=0, M=M[X=1]])"
)

subsets <- list(TRUE, "X==1 & Y==1", "X==1 & Y==1")

# Diagnosis function 

# fit <- fitted_model()

case_selection_diagnose <- function(model, given,  N, sims = 4000, fit, ...)
  
  diagnose_strategies(
    analysis_model = model,
    data_strategies = data_strats(N),
    given = given,
    queries = queries,
    subsets = subsets,
    sims = sims,
    fit = fit,
    iter = 6000,
    chains = 6,
    refresh = 1000,
    ...)

```


```{r diagn, echo = FALSE}  
## Implement

if(do_diagnosis){
  
  {if(!exists("fit")) fit <- fitted_model()}
  
#  for(N in c(2, 4)) { for(m in  1:length(models)) {for(g in 1:length(observed_data)) {	
#  for(N in 4) { for(m in  1:2) {for(g in 1:length(observed_data)) {	
  for(N in 4) { for(m in  2) {for(g in 5) {	
    filename <-  paste0("saved/M_", names(models)[m], "_", N, "_", names(observed_data)[g], ".rds")
    
    print(filename)
    
    write_rds(case_selection_diagnose(models[[m]], observed_data[[g]], N = N, fit = fit), 
              filename)
    
  }}}

  #  for(N in c(2, 4)) { for(m in  1:length(models)) {for(g in 1:length(observed_data)) {	
#  for(N in 4) { for(m in  3:5) {for(g in 1:length(observed_data)) {	
  for(N in 4) { for(m in  4) {for(g in 4:5) {	
    filename <-  paste0("saved/M_", names(models)[m], "_", N, "_", names(observed_data)[g], ".rds")
    
    print(filename)
    
    write_rds(case_selection_diagnose(models[[m]], observed_data[[g]], N = N, fit = fit), 
              filename)
    
  }}}
    
}
```





Graph full set


```{r graph, echo = FALSE}



## Read and display
## PARAMETERS
rds_files <- dir("saved") %>% 
  str_subset("^M.+(chain|restricted|base).+(flat|regr).+rds$")

## UPLOAD MODEL DIAGNOSIS
upload_df <- function(file){
  df <- data.frame()
  toappend  <- read_rds(paste0("saved/",file))$diagnoses_df %>% 
    mutate(
      given = str_match(file,"....(?=\\.rds)"),
      model = str_match(file, "^M_(.*?)_[:digit:]")[,2],
      N     = str_match(file,"[:digit:]")
      ) %>% 
    mutate_if(is.factor, function(x) as.character(tolower(x))) %>% 
    mutate_if(is.character, tolower)
  ifelse(
    "post_var_sd" %in% colnames(toappend),
    TRUE,
    toappend <- mutate(toappend, post_var_sd = NA))
  df <- bind_rows(df, toappend)
}

df <- do.call(rbind,lapply(rds_files, upload_df))

## TIDY UP DF
df$strategy2                            <- df$strategy
df$strategy2[df$strategy == "all_on"]   <- "on_all"
df$strategy2[df$strategy == "all_off"]  <- "off_all"

df <- df %>% mutate(
  strategy22 = word(strategy2,2,sep = "_"),
  strategy21 = word(strategy2,1,sep = "_"),
  strategy21 = case_when(
    strategy21 == "on"  ~ "on diagonal",
    strategy21 == "off" ~ "off diagonal",
    strategy21 == "all" ~ "on x, on y",
    TRUE                ~ strategy21),
  strategy21 = factor(strategy21, levels = c("on diagonal","off diagonal","on x, on y","spread")),
  given      = factor(given, levels = c("flat","regr","conc","necs","suff")),
  model_given = factor(str_c(word(model,1,sep ="_"), "\n", given), 
                       levels = c("chain\nflat",
                                  "chain\nregr",
                                  "restricted\nflat",
                                  "restricted\nregr",
                                  "base\nflat",
                                  "base\nregr"))
  )


df  <- df %>%
  left_join(filter(df,strategy=="prior") %>%  select(given,Query,model,N,prior_var = post_var)) %>%
  filter(strategy != "prior") %>% 
  mutate(gain_prior = (post_var/prior_var-1)*100, 
         gain_prior_upp_ci = ((post_var+2*post_var_sd)/prior_var-1)*100,
         gain_prior_low_ci = ((post_var-2*post_var_sd)/prior_var-1)*100) %>% 
  select(model,Query,starts_with("strategy"),given,N,everything()) %>% 
  arrange(model,Query,strategy,given)

# Iterate over estimands
estimands <- df %>%
  select(Query) %>% 
  distinct() %>% 
  pull
  
# GRAPHS
for (i in 1:length(estimands)) {
print(
  ggplot(df %>%
         filter(Query == estimands[i] & 
         strategy2 != "spread"),
         aes(x=strategy22, y=gain_prior, group = N, colour = N)) +
    # Line at 0 
    geom_hline(yintercept=0, linetype="dashed", size=0.5) +
    # Add point plot
    geom_point(position = position_dodge(width = 0.4)) + 
    # Add error bars
    geom_errorbar(
      width=.1, 
      aes(ymin=gain_prior_upp_ci, ymax=gain_prior_low_ci),
      position = position_dodge(width = 0.4)) +
    # Faceting
    facet_grid(model_given ~ strategy21, scale = "free") +
    # Labeling & Styling
    ylab("Gain in posterior variance over prior (%)") +
    labs(title = paste0("Strategy diagnosis for the ",estimands[i]," estimand"),
         subtitle = "Grouped by strategy & model/given") + 
    theme_bw() + scale_colour_hue(l = 50, c = 50)  + 
    theme(
      legend.title.align = 0.5,
      axis.title.x = element_blank(),
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank())
)
}

```


Starting with the results for the $ATE$ estimand, let us first look at the results for the simple chain model. Where the given data are flat, we see that we do not expect to learn about the $ATE$ from observing $M$ under any case-selection strategy. To be clear, we can certainly learn about the likelihood of *effects*. For instance, if we observe $M=0$ in both an $X=Y=1$ case and in an $X=Y=0$ case, then we know that a causal effect of $X$ on $Y$ is less likely since $M$ is not correlated with either $X$ or $Y$. The trouble is that we become equally less confident in *positive* and in *negative* effects since the given data, implying an $ATE$ of 0, have us believing that both effects are about equally likely: incremental evidence against a positive effect, in this situation, is also incremental evidence against a negative effect.  also neither the original model nor the given data give us any reason to put a higher prior probability on either positive or negative effects (i.e., our prior belief is in an $ATE$ of 0). Thus, evidence of no effect is taken to be evidence, equally, against negative effects and against positive effects. 

However, observe what happens if we are using the chain model and have given data that falls along a positive regression line, rather than being flat. Now, if we observe cases on the regression line (say, that $X=Y=1$ case and the $X=Y=0$ case again), we can learn something. Again, observing $M=0$ in both cases would count against a causal effect of $X$ on $Y$. But *now*, prior to the process tracing, we start out with a belief that the $ATE$ is *positive*, given the positive $X,Y$ correlation. So evidence against any causal effect counts more strongly against a positive effect (the effect we start out thinking is more likely) than against a negative effect, with the result that our posterior on the $ATE$ will fall. 

Note, however, that we do 


we can immediately see striking differences across the three causal models. In particular, we only see clear evidence of expected learning for *any* strategy when operating with the chain or restricted models. For the base model --- in which $X$ may operate directly or indirectly, via $M$, and all nodal types are permitted --- $M$ looks entirely uninformative about average effects. The problem is that the base model is too unconstrained to allow an $M$ data pattern to be informative. We can certainly learn about the likelihood of *effects*. For instance, if we observe $M=0$ in both an $X=Y=1$ case and in an $X=Y=0$ case, then we know that an indirect causal effect of $X$ on $Y$ is less likely since $M$ is not correlated with either $X$ or $Y$. And while there could still be a direct effect, reducing the probability of one pathway carrying the effect reduces our overall confidence that there's an effect. The trouble is that we become equally less confident in *positive* and in *negative* effects since we start out with no prior belief about one of these effects being more likely than the other (i.e., our prior belief is in an $ATE$ of 0). Thus, there can be no change in our beliefs about the *average* effect.


Consider, by contrast, the restricted model, which is identical to the base model but with built-in monotonicity restrictions for all nodes. Now, for an $X=Y=1$ case, only $M=1$ is consistent with a mediated causal effect, thus making $M$ informative about the total effect: if we observe $M=0$, we don't know for sure whether $X$ caused $Y$ since the effect could have run via the direct path, but the probability of an effect is now lower. 

Notably, the chain model also allows for the possibility of learning but *only* if we start out with $X,Y$ data consistent with a strong effect. These given data shift our priors in favor of there being a positive effect of $X$ on $Y$. In turn, they shift our priors in favor of there being an effect of $X$ on $M$ and an effect of $M$ on $Y$, since any $X \rightarrow Y$ effect has to run through these two mediating effects. While these new priors do not operate with the full force of a restriction, they 


## Principles

### Sometimes one case is not enough

For on regression line cases, you can see cases where M correlates positively with X and Y correlates positively to Y; or cases where M is correlated with neither

For off regression line cases, you can see cases where M correlates negatively with X and positively with Y, or vice versa; and you can see cases where M is correlated with neither. 

With no restrictions, you can't actually learn from a single data point  because you have flat priors over X->M and M->Y effects. You're only able to learn from correlations. 




### Different strategies for different estimands

### Where the probative value is

<!-- For the general intuition, recall that the probative value of a process-tracing test hinges on the difference in clue likelihoods associated with the alternative hypotheses in play for a given case. Recall that for different values of $X$ and $Y$ cell, we want to use process tracing to help us distinguish between two specific types that are consistent with the $X, Y$ pattern. Which types are in question varies across $X,Y$ combinations. Table \@ref(tab:FP) illustrates. -->

<!-- | \small      |          $Y=0$           |           $Y=1$             |   -->
<!-- |-------------|:------------------------:|:---------------------------:| -->
<!-- |     $X=0$   |     $b$ or $c$           |       $a$ or $d$            | -->
<!-- |     $X=1$   |     $a$ or $c$           |       $b$ or $d$            | -->
<!-- Table: (\#tab:FP). The ambiguity about types in each $X, Y$ cell. -->

<!-- Thus, in the $X=0, Y=0$ cell, what would be most useful is a clue that has high probative value in distinguishing between an untreated ($X=0$) $b$ type and and an untreated $c$ type. For a case in the $X=1, Y=0$ cell, on the other hand, what matters is how well the clue can discriminate between treated ($X=1$) $a$ and $c$ types. In our notation, it is the difference in $\phi_{jx}$ values for that indicates these cell-specific degrees of leverage.  -->

<!-- To illustrate, consider a situation in which for a given clue we have  $\phi_{b1}$=0.5^[The probability of observing the clue for a $b$ type (positive causal effect) case with $X=1$.]; $\phi_{d1}$=0.5 ^[The probability of observing the clue for a $d$ type (zero causal effect, $Y$ fixed at 1) case with $X=1$.]; $\phi_{b0}$=0.5^^[The probability of observing the clue for a $b$ type (positive causal effect) case with $X=0$.]; and $\phi_{c0}$=0.1^[The probability of observing the clue for a $c$ type (zero causal effect, $Y$ fixed at 0) case with $X=0$.]. In this situation, searching for the clue in $X=Y=1$ cases will yield no leverage since the clue does not discriminate between the two types ($b$ and $d$) that need to be distinguished given $X=Y=1$. Here there is no additional learning about $\lambda_b$ that can be gained from looking for the clue. In contrast, $X=0, Y=0$ cases will be informative since the clue is much better at distinguishing between $b$ and $c$ types---the two types in contention for this kind of case. Thus, although process tracing here does not provide information on the prevalence of positive causal effects ($b$ types) for an $X=Y=1$ case, it does provide information when $X=Y=0$.  -->

<!-- While it is common practice for mixed-method researchers to perform their process tracing "on the regression line," the BIQQ framework suggests that the gains to process tracing for different $X$ and $Y$ values in fact depend on the particular constellations of $\phi$ values for the potentially available clues. More generally, the framework allows one to assess the expected gains from any given case-selection strategy *ex ante* once priors have been specified.  -->




<!-- ## Appendix: Stan uncertainty -->

```{r appendix, echo = FALSE, eval = FALSE}
# manual check of base model with refr data and checking off regression line with data pattern 1

data_check <- M_base_model_2_regr$possible_data_list$all_off[,c(1,3)] %>% expand_data(models$base_model)
update_check <- update_model(models$base_model, data_check)
query_check <- query_model(update_check, "Y[X=1] - Y[X=0]", using = "posteriors")

# Prior check

prior_data <- gen_observed_data(12,4,4,12)  %>% expand_data(models$base_model)
update_prior <- update_model(models$base_model, prior_data, chains = 8, iter = 10000)
query_prior <- query_model(update_prior, "Y[X=1] - Y[X=0]", using = "posteriors")
query_prior


posterior <- query_distribution(update_prior, "Y[X=1] - Y[X=0]", using = "posteriors")

sd(posterior)

n = 8000

resample <- function(posterior, n) {
  sample(1:length(posterior), n, replace = TRUE)
}
resample(posterior)


sds <- replicate(2000, sd(posterior[resample(posterior, n)]))

hist(sds)



iters <- seq(1:12)*1000
sdss <- sapply(iters, function(n) sd(replicate(5000, sd(posterior[resample(posterior, n)]))))

plot(iters, sdss, ylim = c(0,.003 ))
```
