# Mixed methods data strategies {#caseselection}

```{r packagesused13, include = FALSE}
source("_packages_used.R")
do_diagnosis = FALSE
```

:::: {.headerbox data-latex=""}
::: {.center data-latex=""}
:::
With a causal model in hand, together with priors over parameters, we can assess in advance what conclusions we will draw from different observations and assess what kinds of observations are most worth seeking. We draw out the implications of this idea for case selection. We then turn to the problem of choosing between going "wide" and going "deep." We discuss the tradeoffs and communicate an intuition that clue data, even on a small number of cases, can be informative even when there is $X, Y$ data on a very large number of cases, but only if it provides information that cannot be gathered from $X,Y$ data, such as selection into treatment. Simulations suggest that going deep is especially valuable for observational research, situations with homogeneous treatment effects, and, of course, when clues have strong probative value.
::::


A critical decision for scholars employing mixed methods is to determine when and where to look deep and when to look wide. We address this question if two steps, first addressing the problem of case selection and second the question of qualitative/quantitative mixing.  

## Case selection strategies

A host of different strategies have been proposed for selecting cases for in-depth study based on the observed values of $X$, $Y$ data. Perhaps the most common strategy is to select cases in which $X=1$ and $Y=1$ and look to see whether in fact $X$ caused $Y$ in the case in question (using some more or less formal strategy for inferring causality from within-case evidence). But many other strategies have been proposed, including strategies to select cases "on the regression line" or, for some purposes, cases "off the regression line" (e.g., @Lieberman2005nested). Some scholars suggest ensuring variation in $X$ (most prominently, @king1994designing), while others have proposed various kinds of matching strategies. Some have pointed to the advantages of random sampling of cases, either stratified or unstratified by values on $X$ or $Y$ (@FL2008, @HerronQuinn). 

Which cases we should choose will likely depend on the purposes to which we want to put them. 

A matching strategy for instance---selecting cases that are comparable on many features but that differ on $X$---replicates at a small scale the kind of inference done by matching estimators with large-$n$ data. The strategy draws leverage from $X,Y$ variation rather than from within-case information beyond what is available in the measurement of $X$ and $Y$. (FLAG: Citations needed.)

Other treatments seek to use qualitative information to check assumptions made in $X, Y$ analysis: for example, is the measurement of $X$ and $Y$ reliable in critical cases? (FLAG: Citations needed) For such questions with limited resources, it might make sense to focus on cases for which validation plausibly makes a difference to the $X,Y$ inferences: for example influential cases that have unusually extreme values on $X$ and $Y$.^[Note: We can say more about why these would be good choices from a Bayesian perspective, based on the idea that measurement is more likely to be wrong in such cases and shifting them to more typical values would make a big difference.] Similar arguments are made for checking assumptions on selection processes, though we consider this a more complex desideratum since this requires making case level causal inferences and not simply measurement claims.

A third purpose is to use a case to generate alternative or richer theories of causal processes, as in Lieberman's "model-building" mode of "nested analysis" (@Lieberman2005nested). Here it may be cases off the regression line that are of interest.

Weller and Barnes (CITE article) on case selection focus on (a) X/Y relations and (b) whether the cases are useful for hypothesis generation. 

In what follows, we focus on a simpler goal: given existing $X, Y$ data for a set of cases and a given clue (or set of clues) that we can go looking for in the intensive analysis of some subset of these cases, for which cases would process tracing yield the greatest learning about the population-level causal effect of $X$ on $Y$?

The basic insight of this chapter is simple enough: *the optimal strategy for case selection for a model-based analysis are a function of the model we start with and the query we seek to address*, just as we saw for the optimal clue-selection strategy in Chapter \@ref(Clues). Using this strategy yields guidance that is consistent with some common advice but at odds with other advice. The main principles that emerge from the analysis can be summarized as:

* go where the probative value is, and
* sample from $X$ and $Y$ values in proportion to their occurrence in the population,
* invest in collections of cases that provide complementary learning. 

Beyond these general principles, other patterns are more complex and thus more difficult to neatly summarize. The most general message of this chapter is about the general approach: that is, that we can use a causal model to tell us what kinds of cases are likely to yield the greatest learning, given the model and a strategy of inference. We provide a tool for researchers to undertake this analysis, at least for simple problems with $X, Y, K$ data.

Most closely related to our analysis in this chapter is the contribution of @HerronQuinn, who build on @SeawrightGerring2008. While Seawright and Gerring provide a taxonomy of approaches to case selection, they do not provide a strategy for assessing the relative merits of these different approaches.  As we do, @HerronQuinn focus on a situation with binary $X,Y$ data and assess the gains from learning about causal type in a set of cases (interestingly in their treatment causal type, $Z_i$  is called a confounder rather than being an estimand of direct interest; in our setup, confounding as normally understood arises because of different probabilities of different causal types of being assigned to "treatment", or an $X=1$ value). @HerronQuinn assume that in any given case selected for analysis a qualitative researcher is able to infer the causal type perfectly.  

Our setup differs from that in @HerronQuinn in a few ways.  @HerronQuinn paramaterize differently, though this difference is not important.^[@HerronQuinn have a parameter $\theta$ that governs the distribution of data over $X$ and $Y$ and then, conditional on $X,Y$ values, a set of parameters $\psi_{xy}$ that describe the probability of a case's being of a given causal type. We take both $\theta$ and $\psi_{xy}$ to derive from the fundamental distribution of causal types and assignment probabilities. Thus, for example, $\psi_{00}$ from @HerronQuinn corresponds to $\frac{(1-\pi_b)\lambda_b}{(1-\pi_b)\lambda_b + (1-\pi_c)\lambda_c}$ in our notation. The difference in  paramaterization does have implications for interpretations of the priors. For example flat priors over $\theta$ and $\psi$ implies a tighter distribution that a uniform prior over the causal types. In fact @HerronQuinn use priors with greater variance than uniform in any event.] Perhaps the most important difference between our analysis and that in @HerronQuinn  is that we connect the inference strategy to process-tracing approaches. Whereas @HerronQuinn assume that causal types can be read directly, we assume that these are inferred *imperfectly* from evidence and we endogenize the informativeness of the evidence to features of the inquiries. Moreover, not only can we have uncertainty about the probative value of clues, but researchers can learn about the probative value of clues by examining cases.

<!-- As in our baseline model, our ability to make inferences for causal types can differ by type and as a function of $X$.  -->

<!-- Are Herron and Quinn's priors Jeffrey priors? -->

Here we assume that the case selection decision is made after observing the $XY$ distribution and we explore a range of different possible contingency tables. In  @HerronQuinn the distribution from which the contingency tables are drawn is fixed, though set to  exhibit an expected  observed difference in means (though not necessarily a true treatment effect) of 0.2. They assume large $XY$ data sets (with 10,000) units and case selection strategies ranging from 1 to 20 cases.

Another important difference, is that in many of their analyses, @HerronQuinn take the perspective of an outside analyst who knows the true treatment effect; they then assess the expected bias generated by a research strategy over the possible data realizations. We, instead, take the perspective of a researcher who has *beliefs* about the true treatment effect that correspond to their priors, and for whom there is therefore no *expected* bias. This has consequences also for the assessment of expected posterior variance, as in our analyses the expectation of the variance is taken with respect to the researcher's beliefs about the world, rather than being made conditional on some specific world (ATE). We think that this setup is addressed to the question that a researcher must answer when deciding on a strategy: given what they know now, what will produce the greatest reduction in uncertainty (the lowest expected posterior variance)?

Finally, we proceed somewhat differently in our identification of strategies from Herron and Quinn: rather than pre-specifying particular sets of strategies (operationalizations of those identified by @SeawrightGerring2008) and evaluating them, we define a strategy as the particular distribution over $XY$ cells to be examined and proceed to examine *every possible strategy* given a choice of a certain number of cases in which to conduct process tracing. We thus let the clusters of strategies---those strategies that perform similarly---emerge from the analysis rather than being privileged by past conceptualizations of case-selection strategies.

Despite these various differences, our results will agree in key ways with those in @HerronQuinn.


### No general rules

Case selection is about choosing in which cases to make observations. We obviously want to look for evidence in cases where that evidence is likely to be most informative. And the informativeness of a case depends, in turn, on our model and our query.

Although it might be tempting to seek general case-selection rules of the form "examine cases in which $X=1$ and $Y=1$" or "ignore cases in which $X=0$ and $Y=1$," it is easily demonstrated that which cases will be (in expectation) more informative depends on models and queries. 

Suppose we have observed $X$ and $Y$ values for a set of random draw of cases from a population. Suppose further that we know, for this population, that:

  * $X \rightarrow Y \leftarrow K$
  * $\Pr(Y=1|X=0, K = 0) = 1$
  * $\Pr(Y=1|X=1, K = 0) = .5$
  * $\Pr(Y=1|X=0, K = 1) = 0$
  * $\Pr(Y=1|X=1, K = 1) = .9$

One way to read this set of statements is that $X$'s causal effect on $Y$ varies with $K$. We do not know, however how common $K$ is. Thus we do not know either the average effect of $X$ or the probability that $X$ caused $Y$ in a case with particular $X, Y$ values.

What do the above statements tell us about $K$'s *informativeness*? The beliefs above imply that if, $X=Y=1$, then $K$ is a "doubly decisive" clue for assessing whether, in a given case, $X$ causes $Y$. In particular, we see that for an $X=Y=1$ case, observing $K=1$ implies that $X$ caused $Y$: this is because, otherwise, $Y$ would have been 0. We also see that $K=0$, in an $X=1, Y=1$ case implies that $X$ did not cause $Y$ since $Y$ would have still been 1 even if $X$ were 0.  So an $X=Y=1$ case would be a highly informative place to go looking for $K$.

However, if we had a case in which $X=Y=0$, then learning $K$ would be entirely uninformative for the case. In particular, we already *know* that $K=1$ in this case as the statements above exclude the possibility of a case in which $X=Y=0$ and $K=0$. So there would be nothing gained by "looking" to see what $K$'s value is in the case. Moreover, this knowledge is not enough to tell us whether $X=0$ caused $Y=0$.

For the same reason, we can learn nothing from $K$ in an $X=0, Y=1$ case since we know that $K=0$ in such a case. On the other hand, if we chose an $X=1, Y=0$, then $K$ would again be doubly decisive, with $K=0$ implying that $X=1$ caused $Y=0$, and $K=1$ implying that $X=1$ did not cause $Y=0$.

We have chosen extreme values for this illustration --- our beliefs could, of course, allow for gradations of informativeness, rather than all-or-nothing identification --- but the larger point is that beliefs about the way world work can have a powerful effect on the kind of case in which learning is possible. And note that in this example, there is nothing special about where a case lies relative to a (notional) regression line: informativeness in this setup happens to depend on $X$'s value entirely. But, again, this is a particular feature of this particular set of beliefs about the world.

```{r, eval= FALSE, echo = FALSE}
py_k0 = .5
py_k1 = .9

py_k1 / (py_k0 + py_k1)

(1-py_k1) / (1-py_k0 + 1-py_k1)

```

Suppose, now, that we were interested in a population estimand: the average effect of $X$ on $Y$. We can see that this is equal to $\Pr(K=1)\times.9 + (1-\Pr(K-0))\times(-.5)) = 1.4\times Pr(K=1)-.5$. To estimate this estimand, we need only determine the prevalence of $K=1$ in the population. It might seem that this means that it is irrelevant what type of case we choose: why not use pure random sampling to determine $K$'s prevalence. As noted above, however, we have more information about the likely value of $K$ in some kinds of cases than in others. Thus, for this population-level estimand as well, selecting an $X=1$ case will be informative while selecting an $X=0$ case will not be informative.

At the same time, not all $X=1$ cases are equally informative. Should we choose an $X=1$ case in which $Y=1$ or one in which $Y=0$? In both types of case, $K$ is doubly decisive for the probability of causation in the case. However, the two kinds of cases are differentially informative about $K$'s prevalence in the population. For an $X=1, Y=1$ case, we think it moderately likely that $K=1$ (specifically, assuming a prior of $\Pr(K=1)$ we think $\Pr(K=1 | X=1, Y=1) = \frac{.9}{.9+.5}=.64$). For an  $X=1, Y=0$ case, we think $K=1$ is quite unlikely ($\Pr(K=1 | X=1, Y=0) = \frac{.1}{.5+.1}=.17$). In other words, we are much more uncertain about the value of $K$ in the $X=Y=1$ case than in the $X=1, Y=0$ case. In this setup, we would thus expect to learn more about the average treatment effect by choosing to observe $K$ in an on-the-diagonal case than in an off-the-diagonal case.

FLAG: What does "assuming a prior of $\Pr(K=1)$" add to above paragraph?

Specifically, let $\kappa$ denote $\Pr(K=1)$. Suppose that we begin thinking it equally likely that $\kappa=\kappa^H = .5$ and $\kappa=\kappa^L=0$. Suppose, further, that the distribution of $X$ is such that, for any randomly drawn case, $\Pr(X=1) = .5$. We then observe one case with $X=1, Y=1$ and another with $X=1, Y=0$. From that information alone, we can update over $\kappa$. Specifically, conditioning on $X=1$, when we observe the data pattern, $D$, in which $Y=0$ in one case and $Y=1$ in the other, the posterior on $\kappa$ is:

$$p(\kappa = \kappa^L|D) =  \frac{p(D|\kappa^H)}{p(D|\kappa^H)+p(D|\kappa^L)}=\frac{.5}{.5 + .25\times(2\times.5\times.5 + 2\times.9\times.1 + 2\times(.9\times.5 +.1\times.5))}$$

FLAG: Is the above expression not missing the prior?

FLAG: I am not sure what we're trying to get out of this expression, so unsure how to tie this up.


In summary, under the stipulated beliefs about the world, we can learn most about the population $ATE$ by selecting an $X=Y=1$ for study. We can also learn about the case-level effect in such a case as well as in an $X=1, Y=0$ case. If we are interested in the case level estimand for any $X=0$ case, then there are no gains from any case-selection strategy since we know $K$'s value based on $X$ and $Y$'s value.

FLAG: AJ: Check my rewrite of last statement. 

There is nothing preferable in general about an $X=1$ case. Under a different set of beliefs about the world, we would expect to learn more in an $X=Y=0$ than in an $X=Y=1$ case. Suppose, for instance, that we have a model in which: 

  * $X \rightarrow Y \leftarrow K$
  * $\Pr(Y=1|X=0, K = 0) = .5$
  * $\Pr(Y=1|X=1, K = 0) = 0$
  * $\Pr(Y=1|X=0, K = 1) = .5$
  * $\Pr(Y=1|X=1, K = 1) = 1$

In this world, we learn nothing from observing a case in which $X=Y=1$---since we already know that $K=1$. In contrast, if $X=Y=0$, then if we learn that $K=1$, we know that, were $X=1$, $Y$ would have been 1; and if instead we observe $K=0$, we know that $Y$ would have (still) been 0 if $X$ were 1. Now, $K$ is doubly decisive for an $X=Y=0$ case but unhelpful for an $X=Y=1$ case. 

The two off-diagonal cases may also be different in the opportunities for learning that they present. Suppose that you knew that: 

  * $X \rightarrow Y \leftarrow K$
  * $\Pr(Y=1|X=0, K = 0) = 0$
  * $\Pr(Y=1|X=1, K = 0) = .5$
  * $\Pr(Y=1|X=0, K = 1) = 1$
  * $\Pr(Y=1|X=1, K = 1) = .5$

For an $X=1, Y=0$ case, if you observe $K=1$ you know that if $X$ were 0, $Y$ would have been 1; but if $K=0$, you know that if $X$ were 0, $Y$ would still have been 0. In that case $K$, would be doubly decisive for $X=1$ causing $Y=0$.  But in an $X=0, Y=1$, we already know $K=0$ before we go looking.   
  
Say instead that you knew that: 

  * $X \rightarrow Y \leftarrow K$
  * $\Pr(Y=1|X=0, K = 0) = .5$
  * $\Pr(Y=1|X=1, K = 0) = 0$
  * $\Pr(Y=1|X=0, K = 1) = .5$
  * $\Pr(Y=1|X=1, K = 1) = 1$

It may be better to select a case that is not "like" the cases you want to make inferences about.

FLAG: Not getting the point here. How is this different from the setup two up from here?


### Specific case walk through

We introduce a flexible model-based approach to comparing the prospective learning from alternative case-selection strategies. To help explore the intuition behind this strategy, we start by walking through a simplified setup.

```{r selectioncode, echo = FALSE}

# We define a model
model <- make_model("X->M->Y")  %>%
 set_restrictions(c("(Y[M=1]<Y[M=0])", "(M[X=1]<M[X=0])")) %>%
 set_parameter_matrix() %>%
 set_parameters(type = "flat")

# We imagine some preexisting data we have observed
data  <-  data.frame(X = c(0,0,0,1,1,1), M = NA, Y = c(0,0,1,0,1,1))
observed <-  collapse_data(data, model, drop_family = TRUE)

# We can then imagine what data we might observe if we examine M inside some subset of cases
A_on_regression_line <- 
  make_possible_data(
    model, 
    observed = observed, vars = "M", 
    N = c(1,1), 
    condition = c("X==1 & Y==1", "X==0 & Y==0"), 
    prefix = "A")
names(A_on_regression_line)[-c(1:2)] <- c("A1", "A2", "A3", "A4")

B_off_regression_line <- 
  make_possible_data(model, observed = observed, vars = "M", N = c(1,1), 
                     condition = c("X==1 & Y==0", "X==0 & Y==1"), prefix = "B")
names(B_off_regression_line)[-c(1:2)] <- c("B1", "B2", "B3", "B4")

C_X1Y1_only <- 
  make_possible_data(model, observed = observed, vars = "M", N = 2, 
                     condition = c("X==1 & Y==1"), prefix = "C")
names(C_X1Y1_only)[-c(1:2)] <- c("C1", "C2", "C3")

```

Consider a situation in which our model is $X \rightarrow M \rightarrow Y$. Suppose, further, that we restrict the nodal types such that $X$ cannot have a negative effect on $M$, and $M$ cannot have a negative effect on $Y$, with flat priors over all remaining nodal types. Imagine then that we begin by collecting only $X,Y$ data on six cases and obtain the following data pattern:

```{r observed, echo = FALSE}
kable(observed)
```

These $X,Y$ data already give us some information about the causal effect of $X$ on $Y$. Yet, we want to learn more by examining some subset of these cases more deeply -- and, specifically, by colelcting data on $M$ for two of these cases. Which cases should we select? We are considering two strategies, each conditional on $X$ and $Y$ values:

* Strategy $A$ chooses two cases on the regression line, one randomly drawn from the $X=Y=0$ cell and one randomly drawn from the $X=Y=1$ cell
* Strategy $B$ chooses off the regression line, one randomly drawn from the $X=1, Y=0$ cell and one randomly drawn from the $X=0, Y=1$ cell

How can we evaluate these strategies prospectively? 

We start by recognizing that different strategies yield different *possible* data patterns. For instance, Strategy $A$ (on the line) could possibly give us a data pattern that includes the observation $X=0, M=0, Y=0$. Yet Strategy $A$ cannot possibly yield a data pattern that includes the observation $X=1, M=0, Y=0$ --- because it does not involve the inspection of $M$ in an $X=1, Y=0$ case --- whereas Strategy $B$ (off the line) *can* yield a pattern that includes this observation. And neither strategy can possibly yield a pattern that includes both $X=1, M=0, Y=0$ and $X=0, M=1, Y=0$. 

In Table \@ref(tab:chselillustration), we represent the full set of possible data patterns that can arise from each strategy, with the possible data patterns for Strategy $A$ or $B$ labeled $A1, A2$, etc. or $B1, B2$, etc., respectively. As we can see, there are four possible data patterns from each, representing the 4 different combinations of $M$ values we might find across the two cases selected for deeper investigation. In the comparison presented here, none of the possible data patterns overlap across strategies.

The next step is to grapple with the fact that not all of the possible data realizations for a given strategy are equally *likely* to emerge. We represent the data probabilities (calculated by CQTools) near the bottom of the table. How likely a data pattern is to emerge will depend on the model, any restrictions or priors we have built into the model, and any updating of beliefs that arises from the pure $X,Y$ data. Note, for instance, that data pattern $A3$ is much more likely to emerge than the other data patterns possible under Strategy $A$. This is for two reasons. One is that $A3$ involves $M$ covarying with $X$ and $Y$, a pattern consistent with $X$ having an effect on $Y$ -- since, in this model, $X$ can only affect $Y$ if it affects $M$ and if $M$ effects $Y$. Data patterns $A1$ and $A4$ have $M$ constant between the two cases, even as $X$ and $Y$ vary; this is a pattern inconsistent with $X$ having an effect on $Y$. $A3$, then, is more likely than $A1$ or $A4$ because the restrictions on the model plus the evidence from the $X,Y$ data make us believe that $X$ *does* have an average effect on $Y$. Second, we believe $A3$ is more probable than $A2$ because of the model's restrictions: the model allows positive effects of $X$ on $M$ and of $M$ on $Y$ (a way of generating $A3$), but rules out negative intermediate effects (a way of generating $A2$).

Finally, each possible data realization will (if realized) generate (possible) updating of our beliefs about the query of interest. In the second-to-last row of the Table \@ref(tab:chselillustration), we can see the mean of the posterior distribution (for the $ATE$ of $X$ on $Y$) under each data pattern. What is of particular interest in assessing case-selection strategies is the *variance* of the posterior distribution, which represents our *uncertainty*. This posterior variance on the $ATE$, for each data pattern, is represented in the table's final row. We can see that our level of posterior uncertainty varies across possible data realizations. We operationalize the expected learning under each case-selection strategy as the *expected* reduction in posterior variance. 

FLAG: Need to redo Table to conform to text: 2 strategies only.


```{r dtypes, echo = FALSE}
# Combine data types
df <- A_on_regression_line[,-2] %>% 
  merge(B_off_regression_line[,-2], by = "event", all = TRUE) %>%
  merge(C_X1Y1_only[,-2], by = "event", all = TRUE) %>%
  dplyr:::mutate_if(is.numeric, ~replace(., is.na(.), 0))

# table(duplicated(t(df)))
```


```{r illustration, echo = FALSE}

if(do_diagnosis){

# We define a model
model <- make_model("X->M->Y")  %>%
 set_restrictions(c("(Y[M=1]<Y[M=0])", "(M[X=1]<M[X=0])")) %>%
 set_parameter_matrix() %>%
 set_parameters(type = "jeffreys")
  
 strategies <- 
   list(A_on_regression_line, B_off_regression_line, C_X1Y1_only)

 	write_rds(strategies, "saved/14_illustration_strategies.rds")
  
 
 estimates_dbs <- lapply(strategies, function(s) 
    make_estimates_database(model, observed, possible_data = s, queries = "Y[X=1]-Y[X=0]")
    )
    
	write_rds(estimates_dbs, "saved/14_illustration_estimates.rds")

# The reference model is the original model updated with the XY data we've already seen.
 reference_model <- update_model(model, data)
 
	write_rds(reference_model, "saved/14_reference_XMY_updated_model.rds")

 probabilities <- lapply(strategies, function(s) 
#    average_data_probabilities(reference_model, s, using = "posteriors", sims = 3000)

   apply(reference_model$posterior_distribution, 1, function(pars) {
  					make_data_probabilities(model = reference_model, pars = pars, possible_data = s, normalize = TRUE)
  			})
   )
   
	write_rds(probabilities, "saved/14_illustration_probabilities.rds")

}

  reference_model <- read_rds("saved/reference_XMY_updated_model.rds")
	probabilities   <- read_rds("saved/illustration_probabilities.rds")
  strategies      <- read_rds("saved/illustration_strategies.rds")
  estimates_db    <- read_rds("saved/illustration_estimates.rds")
  
  digits <- 3

  df <- rbind(df, 
          c("Probability", round(unlist(probabilities),digits)),
          c("Posterior mean",   (unlist(lapply(estimates_db, function(j)   unlist(unlist(lapply(j, function(k) k$mean))))))),
          c("Posterior variance",   round(unlist(lapply(estimates_db, function(j)   unlist(unlist(lapply(j, function(k) k$mean^2))))), digits)))
```

```{r chselillustration, echo = FALSE}  
kable(df, caption = "Each column shows a possible distribution of data that can be generated from a given strategy. We calculate the probability of each data possibility, given the data seen so far, and the posterior variance associated with each one.", digits = 2)
```

 
From the probability of each data type (given the model and the $X,Y$ data seen so far) and the posterior variance given each data realization, the implied *expected* variance is easily calculated as a weighted average. The expected posterior variances for our two strategies are summarized in Table \@ref(tab:exppostvar):

```{r exppostvar, echo = FALSE}
expected_posterior_var <- sapply(1:3, function(i) 
  sum(probabilities[[i]]*unlist(lapply(estimates_db[[i]], function(x) x$sd^2)))
 )
kable(data.frame(Strategy = c("Online", "Offline", "X=1, Y=1"), Variance = expected_posterior_var), digits = 3)
```


In this example, we see that we would expect to be better off---in the sense of having less posterior uncertainty---by focusing our process-tracing efforts on the regression line than off the regression line. We save an account of the intuition underlying this result for the discussion of our more extensive set of simulations below. 

The key takeaway here are the core elements of our model-based approach to assessing case-selection strategies:

1. Derive from the model the full set of possible data patterns under each case-selection strategy being assessed
2. Calculate the probability of each data pattern given the model (with any priors or restrictions), the prior $X,Y$ data, and the strategy
3. Generate a posterior distribution on the query of interest for each data pattern 
4. Use the probability of and posterior variance under each data pattern toalculate the expected posterior distribution on the query of interest for each strategy



```{r test, include = FALSE}
# source("_packages_used.R")

model <- make_model("X -> M -> Y")
  
gen_observed_data <- function(n00=1, n01=1, n10=1, n11=1,times = 1) {
  data.frame(X = c(0,0,1,1), Y =c(0,1,0,1), M = NA, ns = c(n00, n01, n10, n11)*times) %>%
    tidyr::uncount(ns) %>%
    collapse_data(model, drop_family = TRUE)}


# Data strategies

data_strats <- function(N = 4) {
  if(N%%2 != 0) stop("even Ns only please")
  list(
    all_on = list(N=c(N,N)/2, withins = TRUE, vars = "M",
                  conditions = list("X==0 & Y==0", "X==1 & Y==1")),
    all_off = list(N=c(N,N)/2, withins = TRUE, vars = "M",
                   conditions = list("X==0 & Y==1", "X==1 & Y==0")),
    all_x1 = list(N=c(N,N)/2, withins = TRUE, vars = "M",
                   conditions = list("X==1 & Y==0", "X==1 & Y==1")),
    all_y1 = list(N=c(N,N)/2, withins = TRUE, vars = "M",
                   conditions = list("X==0 & Y==1", "X==1 & Y==1"))
  )
}

# Queries

# if(!exists("fit")) fit <- fitted_model()

# Diagnosis function 

if(do_diagnosis){
test_corr <- 
  diagnose_strategies(
    analysis_model = model,
    data_strategies = data_strats(2),
    given = gen_observed_data(6,2,2,6),
    queries = list(ATE = "Y[X=1]-Y[X=0]", 
                   a = "Y[X=1]<Y[X=0]", 
                   b = "Y[X=1]>Y[X=0]", 
                   c = "(Y[X=1]==0) & (Y[X=0]==0)", 
                   d = "(Y[X=1]==1) & (Y[X=0]==1)",
                   a1 = "M[X=1]<M[X=0]", 
                   b1 = "M[X=1]>M[X=0]", 
                   c1 = "(M[X=1]==0) & (M[X=0]==0)", 
                   d1 = "(M[X=1]==1) & (M[X=0]==1)",
                   a2 = "Y[M=1]<Y[M=0]", 
                   b2 = "Y[M=1]>Y[M=0]", 
                   c2 = "(Y[M=1]==0) & (Y[M=0]==0)", 
                   d2 = "(Y[M=1]==1) & (Y[M=0]==1)"
                   ),
    subsets = list(TRUE, TRUE,  TRUE,  TRUE),
    sims = 8000,
    fit = fit, 
    refresh = 1000,
    iter = 8000)

write_rds(test_corr, "saved/test_corr.rds")

test_flat <- 
  diagnose_strategies(
    analysis_model = model,
    data_strategies = data_strats(2),
    given = gen_observed_data(4,4,4,4),
    queries = list(ATE = "Y[X=1]-Y[X=0]", 
                   a = "Y[X=1]<Y[X=0]", 
                   b = "Y[X=1]>Y[X=0]", 
                   c = "(Y[X=1]==0) & (Y[X=0]==0)", 
                   d = "(Y[X=1]==1) & (Y[X=0]==1)",
                   a1 = "M[X=1]<M[X=0]", 
                   b1 = "M[X=1]>M[X=0]", 
                   c1 = "(M[X=1]==0) & (M[X=0]==0)", 
                   d1 = "(M[X=1]==1) & (M[X=0]==1)",
                   a2 = "Y[M=1]<Y[M=0]", 
                   b2 = "Y[M=1]>Y[M=0]", 
                   c2 = "(Y[M=1]==0) & (Y[M=0]==0)", 
                   d2 = "(Y[M=1]==1) & (Y[M=0]==1)"
                   ),
    subsets = list(TRUE, TRUE,  TRUE,  TRUE),
    sims = 8000, fit = fit, refresh = 1000, iter = 8000)


write_rds(test_flat, "saved/test_flat.rds") 
}

test_corr <- read_rds("saved/test_corr.rds") 
test_flat <- read_rds("saved/test_flat.rds") 

# key thing to note is for $all_x1[[1]] a2 = b2
# and for $all_x1[[2]] a1 = b1 
# so in both cases shareas of a and b remains contstant


```

<!-- For intuition consider first the case with flat prior data: -->

<!-- * if we examine two cases in different quadrants *on the diagonal* (one $X=Y=0$ case and one $X=Y=1$ case) : -->
<!--   * if  we find that $M$ is the same in the two cases then we  increase our belief that $X$ has no effect at all on $Y$ and reduce our confidence that $X$ had a positive or a negative effect on $Y$. Since we are looking on the regression line however, our confidence that $X$ had a *positive* effect on $Y$ is more strongly reduced, producing a  posterior centered on a small negative effect.  -->
<!--   * Conversely if we see that $M$ is different in the two cases, then we have a correlation of the same sign between both $X$ and $M$ and between $M$ and $Y$  cases. We increase our confidence that $X$ mattered for $Y$ in general an din particular that it had a positive effect, resulting in a posterior centered on a small positive ATE. -->

<!-- * if we examine two cases in different quadrants *off the diagonal* (one $X=0, Y=1$ case and one $X=1, Y=0$ case) : -->
<!--   * if  we find that $M$ is the same in the two cases then we  increase our belief that $X$ has no effect at all on $Y$ and reduce our confidence that $X$ had a negative effect  $Y$ (producing a  posterior centered on a small positive effect).  -->
<!--   * Conversely if we see that $M$ is different in the two cases, then we have a correlation (of different signs) between both $X$ and $M$ and between $M$ and $Y$  cases. We increase our confidence that $X$ mattered for $Y$ in general and in particular that it had a negative effect, resulting in a posterior centered on a small negative ATE. -->

<!-- * If we examine data conditioning on the value of $X$ but with variation on $Y$ (for instance $X=1, Y=0$ and   $X=1, Y=1$)  -->
<!-- data patterns do not discriminate between X having a positive effect on Y an d X having a negative effect on Y. However variation in $M$ is more consistent with $M$ being responsive to $X$ and thus the chances that $X$ matters, positively or negatively, overall. In the  case with flat data this changes beliefs on positive and negative effects but not the difference between them. The ate then remains unchanged. FLAG work though intuition  more -->

<!-- * If we examine data conditioning on the value of $Y$ but with variation on $X$ (for instance $X=0, Y=1$ and   $X=1, Y=1$)  -->
<!-- data patterns do not discriminate between X having a positive effect on Y an d X having a negative effect on Y. However variation in $M$ is more consistent with $M$ being responsive to $X$ and thus the chances that $X$ matters, positively or negatively, overall. In the  case with flat data this changes beliefs on positive and negative effects but not the difference between them. The ate then remains unchanged. -->

<!-- For correlated data similar logics apply, but the effects are stronger for evidence on the regression line. The reason is that given correlated data we believe there are more units with positive effects than negative effects. When we find evidence against causal relations on the regression line that reduces our confidence for types with positive effects more than for types with negative effects and the  difference between the  shares with positive effects and negative effects smaller due to the fact that the beliefs in the shares with negative effects is not so strongly reduced. When we find evidence for causal relations this magnifies the  difference between beliefs in shares positive and shares negatives; these effects are magnified however since our confidence for the positive effects change more than for the negative effects, since we are examining cases on the regression line. Conversely when examining cases of the  regression line, the two forces offset each other.  -->



```{r cinfer, include = FALSE}
# Make sense given process tracing approach

long_data <-  expand_data(gen_observed_data(4,2,2,4), model)

if(do_diagnosis){
  updated <- update_model(model, long_data, iter = 14000, chains = 12) %>%
  set_parameters(param_type = "posterior_mean")
  write_rds(updated, "saved/ch13_longupdated.rds") 
}
updated <- read_rds("saved/ch13_longupdated.rds") 

get_parameters(updated)

conditional_inferences(updated, query = "Y[X=1]!=Y[X=0]", given = "!is.na(X) & !is.na(Y) & !is.na(M)") %>%
  arrange(X, Y, M)
```



### Case selection from causal models: a simulation-based approach

In this section, we generalize and more extensively illustrate the model-based approach by applying it to a wide range of models, queries, and case-selection strategies. We generate the results displayed here using `CausalQueries` together with `CQTools`.

In all scenarios examined here, we imagine a situation in which we have already observed some data (the values of some nodes from the causal model in some set of cases) and must now decide in which cases we should gather additional data. To simplify the setup, we will be assuming throughout that we are considering gathering additional observations in cases for which we already have *some* data. In other words, we are deciding which subset of the cases --- among those we have already gathered some data on --- we should investigate more *deeply*. (This is distinct from the question of "wide vs. deep", where we might decide to observe cases we have not yet seen at all.) 

The general intuition of the case-selection approach that we develop here is that we can use our causal model and any previously observed data to estimate what observations we are more or less likely to make under a given case-selection strategy, and then figure out how far off from the (under the model) true estimand we can expect to be under the strategy, given whatever causal question we seek to answer. 

We proceed as follows:

**DAG**. We start, as always, with a DAG representing our beliefs about which variables we believe to be direct causes of other variables. For the current illustrations, we consider two different DAGS: a simple mediation (or "chain") model, $X \rightarrow M \rightarrow Y$, and a two-path model, $X \rightarrow M \rightarrow Y \leftarrow X$.

**Restrictions or priors**. As when conducting mixed-method inference, we can set qualitative restrictions and/or differential quantitative weights on the (possibly conditional) nodal types in the model. We can also indicate our uncertainty over the latter, by setting the $\alpha$ parameters of the relevant Dirichlet distributions. For the current example, we start by setting flat priors over all nodal types and assume no unobserved confounding.

**Given data.** If we have already made observations of any of the model's nodes in some set of cases, we can use this information to condition our strategy for searching for further information. For instance, if we have observed $X$'s and $Y$'s value in a set of cases, we might select cases for process tracing based on their values of $X$ and/or $Y$. And, importantly, what we have already observed in the cases will affect the inferences we will draw when we observe additional data, including how *informative* a particular new observation is likely to be. For the present examples, we assume that we have already observed $X$ and $Y$ in a set of cases and found a positive correlation.

**Query**. We define our query. This could, for instance, be $X$'s average effect on $Y$ or it might be the probability that $X$ has a negative effect on $Y$ in an $X=1, Y=0$ case. We can use the general procedure to identify case-selection strategies for any causal query that can be defined on a DAG. And, importantly, the optimal case-selection strategy may depend on the query. The best case-selection strategy for answering one query may not be the best case-selection strategy for another query. 

**Define one or more strategies**. A strategy is defined, generically, as the search for data on a given set of *nodes*, in a given *number* of cases that are randomly selected *conditional* on some information we already have about potential cases. In the simulations below, our strategy will always involve uncovering $M$'s value in 1 or 2 cases. What we are wondering is how to choose these one or two cases for deeper analysis. 

**Possible data**. For each strategy, there are multiple possible sets of data that we could end up observing. In particular, the data we could end up with will be the $X,Y$ patterns we have already observed plus some pattern of $M$ observations. 

**Probability of the data**. We then calculate a probability of each possible data realization, given the model (with any restrictions or priors) and any data that we have already observed. In practice, we do this in `CQtools` via simulation. Starting with the model together with our priors, we update our beliefs about $\lambda$ based on the previously observed data. This posterior now represents our *prior* for the purposes of the process tracing. In the analyses below, we use the already-observed $X,Y$ correlation to update our beliefs about causal-type share allocations in the population. We then use this posterior to draw a series of $\lambda$ values. 

Given that the ambiguity matrix gives us the mapping from causal types to data realizations, we can calculate for each $lambda$ draw the probability of each data possibility given that particular $\lambda$ and the strategy. We then average across repeated $\lambda$ draws. Since $\lambda$'s are being drawn from our prior, we are automatically weighting more heavily those $\lambda$'s that we believe to be most likely.

**Posterior on estimate given the data**. For each data possibility, we can then use `CQtools` to ask what inference we would get from each data possibility, given whatever query we seek to answer, as well as the variance of that posterior. Examining the inferences from possible data-realizations, as we do below, can help us understand how the learning unfolds for different strategies. 

**Expected posterior variance under each strategy**. The quantity of ultimate interest is the posterior variance that we expect to end up with under each strategy. The expected posterior variance is simply an average of the posterior variances under each data possibility, weighted by the probability of each data possibility. We operationalize the expected learning under a strategy as the expected *reduction* in posterior variance arising from that strategy.


#### Models, queries, and strategies

Here we describe the dimensions along which we vary the features of the simulations.

*Models*

We illustrate the approach using a set of relatively simply models with core structural features that we believe will be fairly common in applied settings. The four structures that we examine are:

* Chain model: an $X \rightarrow M \rightarrow Y$ model, where $M$ is a mediator
* Confounded model: a model with $X \rightarrow Y$ and with $M$ as a confounder, pointing into both $X$ and $Y$
* Moderator model: an $X \rightarrow Y \leftarrow M$, where $M$ is a moderator
* Two-path model: a model in which $X \rightarrow M \rightarrow Y \leftarrow X$, meaning that $X$ can affect $Y$ both through a direct path and indirectly via $M$

For each of these causal structures, we consider both an unconstrained version (all nodal types permitted) and a monotonic version in which we use restrictions to exclude negative effects throughout the model.


*Queries*

We also examine a range of queries under each model:

* $ATE$: what is the average effect of $X$ on $Y$ for the population?
* Probability of positive causation: what is the probability that $Y=1$ because $X=1$ for a case randomly drawn from the population of $X=1, Y=1$ cases
* Probability of negative causation: what is the probability that $Y=1$ is due to $X=0$ for a case randomly drawn from the population of $X=0, Y=1$ cases
* Probability of an indirect effect: defined only for the two-path models, we estimate the probability that the effect of $X$ on $Y$ operates through the indirect path. More precisely, we ask, for an $X=1, Y=1$ case in which $X=1$ caused $Y=1$, what the probability is that that effect would have occurred if $M$ were held fixed at the value it takes on when $X=1$.^[In code, this somewhat complicated query is expressed as `"Y[X=0, M=M[X=1]]==Y[X=1, M=M[X=1]]"`, `given "(X == 1 & Y == 1) & (Y[X=1]>Y[X=0])"`.]

<!-- $X$'s effect on $M$  causal effect along the indirect pathway, through $M$. We calculate two indirect effects, one conditional on $X=0$ and one conditional on $X=1$. Each of these effects is defined as the effect on $Y$ of the change in $M$ that *would* result from a change in $X$ (from 0 to 1 or from 1 to 0, respectively), but with $X$ held constant (at 0 or 1, respectively). -->

*Strategies*

Finally, for each model-query combination, we assess the contributions of 7 strategies for selecting cases for process tracing, with inferences from the $X,Y$ data alone serving as our baseline. In the figures below, the strategies run along the $X-$axis of each graph and can be interpreted as follows:

* **Prior**: beliefs are based on $X,Y$ data only. 
* **1 off**: data on $M$ is sought in one case in the $X=1, Y=0$ cell
* **1 on**: data on $M$ is sought in one case in the $X=1, Y=1$ cell
* **2 off**: data on $M$ is sought in one $X=0, Y=1$ case and one $X=1, Y=0$ case
* **2 pos**: data on $M$ is sought for two cases in the $X=1, Y=1$ cell 
* **2 on**: data on $M$ is sought in one $X=1, Y=1$ case and one $X=0, Y=0$ case
* **fix $X$**: a strategy in which we seek $M$ in two cases in which a causal condition was present, with $X$ fixed at 1, one with $Y=0$ and one with $Y=1$
* **fix $Y$**: a strategy in which we seek $M$ in two cases in which a positive outcome was observed, with $Y$ fixed at 1, one with $X=0$ and one with $X=1$

These are all "pure" strategies in the sense that the number of units for which data on $M$ is sought in each cell is fixed. One could also imagine random strategies in which a researcher chooses at random in which cells to look. For example, if we choose one point at random, we are randomly choosing between a case on the regression line and a case off the line. The performance of a random strategy will be a weighted average of the pure strategies over which the random strategy is randomizing.

For all simulations, we assume prior $X,Y$ data of $N=6$, with a weak positive relationship (2 $X=1, Y=1$ cases, 2 $X=0, Y=0$ cases, and 1 case in each off-diagonal cell). And it is *from* these original 6 cases that we are selecting our cases for process tracing.


#### Results

The main results are shown in Figure \@ref(fig:caseselectionest) and Figure \@ref(fig:caseselectionvar). For each model-strategy-query combination, we figure out (a) all of the possible data realizations, (b) what inferences would be made on the query from each data realization, (c) how likely each data-realization is to arise given the strategy, the model, and the prior data, and (d) the resulting expected reduction in variance from each strategy. Result (d) is our measure of expected learning. 

The two figures take two different approaches to representing the value of alternative strategies.  In Figure \@ref(fig:caseselectionest), we examine the informativeness of strategies by showing how much our inferences depend on what we observe within the cases. Generally, a larger spread across points (for a given model-query-strategy combination) represents a greater opportunity for learning from the data. However, as expected learning is also a function of how likely each data realization is, we represent the probability of each potential inference via shading of the points. In Figure \@ref(fig:caseselectionvar) we directly plot expected learning, operationalized as the expected reduction in posterior variance. 

In the remainder of this section, we walk through the results and suggest, often tentatively, interpretations of some of the more striking patterns. We caution that reasoning one's way through expected learning for different model-query-strategy combinations, given a particular pattern in the prior data, can be tricky --- hence, our recommendation that researchers simulate their way to research-design guidance, rather than relying on intuition. 

<!-- These figuresare close connected of course, with a higher dispersion of estimates in Figure \@ref(fig:caseselectionest) implying a greater reduction in variance in Figure \@ref(fig:caseselectionvar). -->




```{r caseselectionest, fig.height = 7, fig.width = 10, fig.cap = "Inferences given observations", warning = FALSE, echo = FALSE} 

read_rds("saved/Case_selection_1.rds")
```


```{r caseselectionvar, fig.height = 7, fig.width = 10, fig.cap = "Reduction in variance on ATE and PC given strategies", echo = FALSE, warning = FALSE} 

read_rds("saved/Case_selection_2.rds")

```



<!-- We see also the reduction in variance for other estimands, notably the effects of $X$ on $M$, the effects of $M$ on $Y$ as well as the direct and indirect effects for our two-path models. We display these to help with interpretation of the main results but they also usefully demonstrate the applicability of the procedure to different types of queries we might be concerned with. -->

<!-- #### Priors: From the $X,Y$ data only -->


<!-- <!-- we see our pure prior on each query before we have seen the $X,Y$ correlation --- i.e., based strictly on the DAG and flat priors over nodal types. The story here is straightforward. We start out believing that, for the $X \rightarrow M$ relationship, the population is evenly divided across the four types, $\theta^M_{00}, \theta^M_{01}, \theta^M_{10}, \theta^M_{11}$ (so 0.25 share of each). Likewise for $\theta^Y$ in relation to the $M \rightarrow Y$ relationship. We can get a positive $X,Y$ effect from a causal type involving $\theta^M_{01}$ and $\theta^Y_{01}$ or from a causal type involving $\theta^M_{10}$ and $\theta^Y_{10}$. Each of those causal types has a 0.125 probability, implying a 0.25 share of positive effects. A mirror-image of that logic gets us to a 0.25 share of negative effects, with the remaining 0.5 of the population comprising causal types that yield null effects (involving either a null effect at one or both steps in the chain). Thus, the expected effect of $X$ on $Y$ for a randomly drawn case from the population, given nothing but the model, is 0.  --> -->

<!-- <!-- moving across the row, we see our prior beliefs for cases with more specific features. For cases with $X=1, Y=1$, our priors imply a positive effect 25\% of the time.^[This belief arises from a straightforward calculation involving the different combinations of nodal types that could generate $X=1, Y=1$ and their prior probabilities.] As we have flat priors over $M$'s type and its effects on $Y$, conditioning $M$'s value leaves this belief unaffected. By a parallel logic, the model and flat priors imply the belief that 0.25 of all $X=0, Y=1$ cases to involve a negative effect of $X$ on $Y$. --> -->

<!-- We turn first to the chain model. Before examining inferences from data, it is worth noting what a chain model with flat priors implies about the queries of interest. We start out believing that, for the $X \rightarrow M$ relationship, the population is evenly divided across the four types, $\theta^M_{00}, \theta^M_{01}, \theta^M_{10}, \theta^M_{11}$ (so 0.25 share of each). Likewise for $\theta^Y$ in relation to the $M \rightarrow Y$ relationship. We can get a positive $X,Y$ effect from a causal type involving $\theta^M_{01}$ and $\theta^Y_{01}$ or from a causal type involving $\theta^M_{10}$ and $\theta^Y_{10}$. Each of those causal types has a 0.125 probability, implying a 0.25 share of positive effects. A mirror-image of that logic gets us to a 0.25 share of negative effects, with the remaining 0.5 of the population comprising causal types that yield null effects (involving either a null effect at one or both steps in the chain). Thus, the average effect of $X$ on $Y$, given nothing but the model, is 0. Moreover, for an $X=1, Y=1$ case, the probability of a (positive) causal effect is also 0.25, as it is for a negative effect in an $X=0, Y=1$ case. -->

<!-- Now, we introduce data. In the "prior" column of each graph in Figure \@ref(fig:caseselectionest), we see our inferences from the initial set of 6 $X,Y$ observations -- that is, *prior* to process tracing. In this model, as there is no confounding, a positive $X,Y$ correlation is evidence of a higher relative share of cases in which $X$ has a positive effect on $Y$. This implies a positive average treatment effect as well as a greater than 0.25 chance of causation for an $X=1, Y=1$ case since the $X,Y$ pattern suggests that there are more cases where $X$ has a positive effect on $Y$ than there are cases in which $Y$ is fixed at 1.  -->

<!-- Notably, however, our beliefs for an $X=0, Y=1$ case are unaffected by observing the positive correlation. To return for a moment to our handy $a,b,c,d$ typology, the reason is that the positive $X,Y$ correlation affects our beliefs about the share of $b$ types relative to $a,c,$ and $d$ types (with reference to the overall $X \rightarrow Y$ effect), but it is uninformative about the relative shares of $a$ vs. $d$ types, the two types in contention for an $X=0, Y=1$ case. In the $X,Y$ data, the $X=0, Y=1$ case is equally likely to be an $a$ or a $d$, which is exactly what our prior belief was about a random $X=0, Y=1$ case.^[Put differently, the observed $X,Y$ correlation is equally consistent with all $X=0, Y=1$ cases being $a$'s, with them all being $d$'s, or with anything in between.] -->

<!-- (Again, conditioning on $M$'s value has no effect since we still have no information to shift us off our flat priors relating to $M$'s causes or effects.) -->

#### $N=1$ strategies, unconstrained models

Suppose that we can only conduct process tracing (observe $M$) for a single case drawn from our sample of 6 $X,Y$ cases. Should we choose a case from on or off the regression line implied by the $X,Y$ pattern?  In Figure \@ref(fig:caseselectionest), we can see that for all unconstrained models, our inferences are completely unaffected by the observation of $M$ in a single case, regardless of which case-selection strategy we choose and regardless of the query of interest. We see only 1 point plotted for the two $N=1$ strategies for all unconstrained models and all queries because the inference is the same regardless of the realized value of $M$. In Figure \@ref(fig:caseselectionvar), we see, in the same vein, that we expect 0 reduction in expected posterior variance from these $N=1$ strategies: they cannot make us any less uncertain about our estimates because the observations we glean cannot affect our beliefs.

To see why, let's first consider the on-the-line strategy. Not having observed $M$ previously, we still have flat priors over the nodal types governing $X$'s effect on $M$ and $M$'s effect on $Y$. That is to say, we still have no idea whether $X$'s positive effect on $Y$ (where present) more commonly operates through a chain of positive effects or a chain of negative effects. Thus, the observation of, say, $M=1$ in an $X=1, Y=1$ case is equally consistent with a positive $X \rightarrow Y$ (to the extent that effect operates via linked positive effects) and with no $X \rightarrow Y$ effect (to the extent positive effects operate through linked negative effects). Observing $M=1$ in an $X=1, Y=1$ case therefore tells us nothing about the causal effect in that case and, thus, nothing about the average effect either. 

Similarly, we have no idea whether $X$'s negative effect on $Y$ (where present) operates through a positive-negative chain or a negative-positive chain, making $M=1$ or $M=0$ in an $X=1, Y=0$ case both equally consistent with a negative or null $X \rightarrow Y$ effect, yielding no information about causation in the case. By a similar logic, observing $M=1$ in the $X=1, Y=1$ case is uninformative about negative effects in an $X=0, Y=1$ case, and observing $M=1$ in an $X=1, Y=0$ case tells us nothing about positive effects in an $X=1, Y=1$ case. 

The same logic applies to drawing inferences from $M$ as a moderator or to learning from $M$ about indirect effects. In the absence of prior information about effects, one case is not enough.


#### $N=1$ strategies, monotonic models

The situations changes, however, when we operate with models with montonicity restrictions. Now we can see that our inferences on the queries do generally depend on $M$'s realization in a single case and that we expect learning. For many model-query combinations, the two $N=1$ strategies perform comparably, but there are situations in which we see substantial differences. 

Most notably, in a chain model with negative effects ruled out by assumption, we learn almost nothing from choosing an off-the-line case: this is because we already know from the model itself that there can be no $X \right arrow Y$ effect in such a case since such an effect would require a negative effect at one stage. The only learning that can occur in such a case is about the prevalence of positive effects (relative to null effects) at individual stages ($X \rightarrow M$ and $M \rightarrow Y$), which in turn has implications for the prevalance of positive effects (relative to null effects) of $X$ on $Y$. Likely for similar reasons, in the monotonic two-path model, an on-the-line case is much more informative than an off-the-line case about the $ATE$ and about the probability that the effect runs via the indirect path. 

Interestingly, however, the on-the-line strategy is not uniformly superior for an $N=1$ process-tracing design. We appear to learn significantly more from an off-the-line case than an on-the-line case when estimating the share of positive effects in the population of $X=1, Y=1$ cases and operating with a monotonic confounded or two-path model. At first, these results seem surprising: why would we not want to choose an $X=1, Y=1$ case for learning about the population of $X=1, Y=1$ cases? One possible reason is that, in the on-the-line case, one data realization is much more likely then the other, while we are more uncertain about what we will find in the off-the-line case. For instance, in the confounding model with montonicity, in an $X=1, Y=1$ case we would learn about the prevalence of confounding from seeing $M=0$ (where confounding cannot be operating since negative effects are excluded) as opposed to $M=1$; but we do not expect to see $M=0$ when both of its children ($X$ and $Y$) take a value of 1 while negative effects are excluded. In an $X=1, Y=0$ case, however, $M=0$ and $M=1$ are about equally likely to be observed, and we can learn about confounding from each realization. We can see these differences in relative data probabilities from the shadings in the graphs, where we have more even shading for the possible inferences from the one-off strategy than for the one-one strategy. 

The general point here is that we expect to learn more from seeking a clue the more uncertain we are about what we will find, and some case-selection strategies will give us better opportunities to resolve uncertainty than do others. 


#### $N=2$ strategies, unconstrained models

Next, we consider the process tracing of *two* of our cases 6 cases. Now, because we are observing $M$ in two cases, we can learn from the variation in $M$ across these cases---or, more specifically, from its covariation with $X$ and with $Y$. This should matter especially for unconstrained models, where we start out with no information about intermediate causal effects (e.g., whether they are more often positive or more often negative). Thus, when we only process trace one case, we cannot learn about causal effects in the cases we process-trace since we don't know how to interpret the clue. In contrast, if we observe $M$ in two or more cases, we *do* learn about causal effects for those cases because of the leverage provided by observing covariation between the process-tracing clue and other variables.

<!-- Thus, while we have pointed out that observing $M$ within a case can lend $M$ probative value for *other* cases in which we might choose to observe $M$, $M$ has no probative value *for* that one case.  -->

We assess the expected gains from 5 $N=2$ stragies: examine two off-the line cases, one $X=1, Y=0$ case and one $X=0, Y=1$ case; examine two on-the-line cases, an $X=Y=0$ case and an $X=Y=1$ case; examine two treated, positive outcome ($X=Y=1$) cases; select on $X$ by examining two $X=1$ cases with different $Y$ values; and select on $Y$ by examining two $Y=1$ cases with different $X$ values. 

A key message of these results is that, with 2 cases, the performance of each strategy depends quite heavily on the model we start with and what we want to learn. For instance, when estimating the $ATE$, the on-the-line strategy in which we disperse the cases across cells (two-on) clearly outperforms both the dispersed off-the-line strategy (two-off) and an on-the-line strategy in which we concentrate on one cell (two-pos) *if* we are working with an unconstrained chain model, and the off-the-line strategy is clearly the worst-performing of the three. The differences in learning about the $ATE$ are more muted, however, for an unconstrained confounded model, and the two-pos strategy does *better* than the other two for a two-path model.

If we seek to learn about the probability of positive causation in an $X=1, Y=1$ case, then there is little difference between two-off and two-pos, with two-on performing best. We also see that two-pos has lost its edge in an unconstrained two-path model, with no strategy offering leverage. When estimating the probability of a *negative* effect for an $X=0, Y=1$ case, we see that the two-off strategy performs best for the chain model, *but* that the two-pos strategies offers the greatest leverage in a two-path model. Finally, when estimating the probability of an indirect positive effect in an unconstrained two-path model, we get the most from a two-on strategy, though the two-off strategy does moderately well.

In general, selecting conditional on a fixed value of $X$ or $Y$ (while dispersing on the other variable) does not do particularly well in unconstrained models, and it does not usually matter much which variable we fix on. There are exceptions, however. Perhaps most strikingly, in a two-path unconstrained model, we do relatively well in estimating the probability of an indirect positive effect when we fix $Y$ but stand to learn nothing if we fix $X$. Interestingly, fixing $Y$ fairly well dominates fixing $X$ across all model-query combinations shown, given the prior data pattern we are working with.

This pattern is particularly interesting in light of canonical advice in the qualitative methods literature. King, Keohane, and Verba [-@king1994designing] advise selecting for variation on the explanatory variable and, as a second-best approach, on the dependent variable. And they warn sternly against selection for variation on both at the same time. But note what happens if we follow their advice. Suppose we decide to select for variation on $X$, ignoring $Y$. We might end up with a pair of informative on-the-line or off-the-line cases. But we might just as easily (depending on the joint $X,Y$ distributon in the population) end up with a fairly uninformative $X=0, Y=1$, $X=1, Y=1$ pair. King, Keohane, and Verba's advice makes sense if all we are interested in is examining covariation between $X$ and the $Y$: then we can learn from forcing $X$ to vary and letting $Y$'s values fall where they may. However, seeking leverage from a mediator is a different affair. As our simulations indicate, for at least some common causal structures, we actually stand to learn the most from observing that mediator when we choose to examine it across a set of cases selected for variation in both $X$ and $Y$.

<!-- We plot four points for each strategy (for each query), representing the inference we would draw from four possible realizations of $M$. In most analyses, the inferences are essentially the same for pairs of data-realizations, so only two dots are apparent.^[For instance, whether we see $M=1$ in the $X=0, Y=0$ case and $M=0$ in the $X=1, Y=1$ case or vice versa, the inference is the same for a model with no restrictions.] While unlabelled, the higher dot represents the inference we would draw if $M$ varies with $X$; the lower dot represents the inference we would draw if $M$ is constant across the two cases. -->

<!-- Let's consider the on-the-line strategy first. Consider what we infer if we observe $M$ covary (either positively or negatively) with $X$ in these cases. What we are seeing, then, is that the mediator covaries across cases with both the cause and the outcome, which constitutes evidence consistent with a causal effect of $X$ on $Y$. Since the observed pattern of covariation is consistent with a positive $X \rightarrow Y$ effect (via linked positive effects), our belief about the average effect of $X$ on $Y$ goes up. For the same reason, so too does our belief about the probability of causation in an $X=1, Y=1$ case.  -->

<!-- <!-- We see an even bigger boost in our estimate of the effect in an $X=1, Y=1$ case conditional on $M$ also being 1. Two things are generating this upward updating. First, just as for the unconditional estimate and the estimate conditional on $X=1, Y=1$, the process tracing generates a stronger belief that $X$ generally has a positive effect on $Y$. Second, as for the $N=1$ process tracing, seeing $M=1$ in two on-the-line cases (which we already believe are mostly positive-effect cases) makes $M$ informative as a clue in favor of a positive effect---indeed, more informative than if we'd only seen $M=1$ in one on-the-line case. --> 

<!-- Now, supppose we we observe $M=1$ or $M=0$ in both on-the-line cases. This data pattern represents clear evidence *against* positive effects. We have observed that $M$ is unresponsive to changes in $X$, and that $Y$ varies without $M$ varying---null effects at both stages of the chain. Now we see that our posterior on the $ATE$ drops to about $0$. We also see a sharp downward movement in the probability of causation for an $X=1, Y=1$ case, representing our updated belief that there are few positive effects.  -->

<!-- In Figure \@ref(fig:caseselectionvar), we see the consequences for the expected reduction in posterior variance: the large potential swings in inferences, depending on what we observe, generate the potential for quite a lot of learning. Likewise, in Figure \@ref(caseselectionvar2), we can see the learning about effects at each stage: the relatively large reduction in expected posterior variance in our estimates of the average effect of $X$ on $M$ and of $M$ on $Y$. -->

<!-- <!-- Interestingly, whereas conditioning additionally on $M=1$ generated in a *higher* estimated effect when we observed $M$ correlated with $X$ and $Y$, it generates a *lower* estimated effect when we observe $M=1$ in both cases. One reason is likely that we have updated our belief about $\theta^M$ such that we now believe that $X=1, M=1$ cases are more likely to be $\theta^M_{11}$'s (no effect) than $\theta^M_{01}$'s. We have not similarly updated with respect to $X=1, M=0$ cases and the $\theta^M_{00}$'s than $\theta^M_{10}$ shares. Thus, for an $X=1,Y=1$ case, additionally specifying that $M=1$ implies an even higher likelihood that the case contains a null effect.  --> 

<!-- Moving down the queries, we also see that process-tracing the on-the-line cases yields leverage on negative causal effects in an *off-the-line* case. In particular, though not directly indicated on the graphs, finding evidence consistent with positive effects in those cases leads us to update in favor of stronger *negative* effects as well, while finding $M$ constant in both cases leads to a downard shift in the probability of negative causation. Why is this? -->

<!-- Our answer is somewhat speculative, but we think there are two things going on. One is simply that the correlations between $M$ and $X$ and between $M$ and $Y$ across the two process-traced cases constitutes evidence of *effects*, period, at the intervening steps---a necessary condition for *any* $X \rightarrow Y$ effect, positive or negative. So, for an $X=0, Y=1$ case, a stronger belief that $X$ has some effect on $M$ and that $M$ has some effect on $Y$ implies a stronger belief in a negative $X \rightarrow Y$ effect.  -->

<!-- The second thing that is probably going on is constraint imposed by prior beliefs. Having seen the $X,Y$ data, we have previously formed a belief that the *average* effect of $X$ on $Y$ is about 0.05. When we find evidence in favor of positive effects, we update to a set of beliefs that are a compromise between the new data and prior beliefs. On the one hand, we upwardly update on the $ATE$. On the other hand, our beliefs are constrained from moving *too* far away from an $ATE$ of 0.05. Recall that, in a binary setting, the average effect in a population is the share of cases with positive effects minus the share with negative effects. Thus, the conservative influence of our priors means that, as we upwardly update on positive effects, we also upwardly update our beliefs about the prevalence of *negative* effects, though not by as much. The result is that our beliefs about the $ATE$ do not shift as much as they would if we had had flat priors going into the process tracing. By more deeply anchoring our beliefs about average effects, then, the prior $X,Y$ evidence means that any evidence in favor of positive effects also constitutes evidence in favor of negative effects.  -->

<!-- To put this point another way, while we learn *directly* about positive effects from process-tracing the on-the-line cases since these are cases in which positive effects can possible occur. But we then learn *indirectly* about negative effects from these cases *via* the prior-imposed constraint on the average effect. -->

<!-- We also consider an off-the-line $N=2$ strategy, selecting for process tracing one $X=1, Y=0$ case and one $X=0, Y=1$ case. Again, the two types of data-realization we consider are that $M$ varies with $X$ across the two cases (higher point)  or that $M$ is constant in both cases (lower point). One thing we can observe immediately is that we get less leverage from this strategy if we are interested in either the $ATE$ or positive effects. In Figure \@ref(fig:caseselectionest), looking at both the $ATE$ and probability of causation in an $X=1, Y=1$ case, the difference in our posterior beliefs between the two data-realizations is smaller for this strategy than it is for the two data-realizations of the on-the-line strategy. We likewise see less reduction in posterior expected variance for these queries with an off-the-line strategy. However, we learn more from the off-the-line strategy about the probability of *negative* causation.  -->

<!-- Unpacking this further, we know that we can learn directly about *negative* $X \rightarrow Y$ effects through process-tracing off-the-line cases. If we see $M$ covary with $X$ across these two cases---which also means covarying (though in the opposite direction) with $Y$---we have found evidence in favor of the operation of negative $X \rightarrow Y$ effects. Hence the upward movement in our posterior on the probability of negative causation. We also upwardly update our posterior on positive effects in an $X=1, Y=1$ case, though by a smaller amount. Again, some of this movement is likely attributable to the constraint imposed by the prior generated from the $X,Y$ data; and some of this updating is likely attributable to the fact that we have found evidence against null intermediate effects.  -->

<!-- <!-- Curiously, under this strategy and data-realization, our beliefs about effects conditional on $X=0, M=1, Y=1$ move in the *opposite* direction from our beliefs conditional just on $X=0,Y=1$. Why might this be? This is likely because of the way in which the process-tracing has lent probative value to $M$. The pattern of $M$ data is most consistent with any negative $X \rightarrow Y$ effect operating via a *positive* $X \rightarrow M$ effect followed by a *negative* $M \rightarrow Y$ effect. Yet the intermediate effects would have to have the opposite sign for $X=0, M=1, Y=1$ to be consistent with an $X \rightarrow Y$ effect. In other words, $M=1$ rules out the likeliest way in which an $X=0, Y=1$ case could contain an effect.  -->

<!-- If on the other hand we observe $M$ constant in both off-the-line cases, we have found evidence against the operation of negative effects. Hence the the strong downward movement of our posterior on the probability of negative causation. Likewise, we see (weaker) upward movement in our posterior on the $ATE$ and, for reasons discussed earlier, in our posterior on positive effects. In sum, an off-the-line strategy generates "direct" learning about negative effects and indirect, weaker learning about positive effects. We see the corresponding learning about average effects at each stage in the causal chain in Figure \@ref(caseselectionvar2). -->

<!-- The remaining puzzle is why the learning about the $ATE$ from the off-the-line strategy is weaker than under the on-the-line strategy. This is at first glance puzzling in that the two strategies generate symmetrical opportunities to observe an $M$ pattern that is consistent or inconsistent with causal effects. We believe there are two reasons, both relating to where the direct and indirect learning falls. -->

<!-- The first, more mechanical reason is that we start, given the $X,Y$ data, with a higher and higher-variance posterior on the prevalence of positive effects as compared to negative effects. So there is simply more scope for the data to influence our beliefs about positive effects, and the on-the-line cases are those in which we can learn "directly" about positive effects. In support of this view, note how much bigger the swing in beliefs, conditional on the data pattern, we see for the probability of positive causation than for the probability of negative causation. -->

<!-- The second reason is prevalence: given the $X,Y$ correlation, we believe that there are more cases out there that potentially contain positive effects than there are cases that potentially contain negative effects. Thus, *whatever* it is we learn about positive effects is going to have a bigger impact on our beliefs about the population as a whole than is whatever we learn about negative effects. Since on-the-line cases yield the most direct learning about positive effects, they are therefore the most informative about the population. -->

<!-- Note, importantly, that this finding is consistent with recommendations in the literature to investigate on-the-line cases [@Lieberman2005nested; @goertz2017multimethod],^[We set aside here Lieberman's recommendation to choose off-the-line cases for discovery-oriented "model-building" because we see that enterprise as distinct from the estimation endeavor to which our analysis is oriented.] but for a very different reason. Lieberman and Goertz emphasize the importance of examining cases where the theorized mechanism connecting $X$ to $Y$ might potentially play out, arguing that this would be in the set of cases that conform to theory in their $X,Y$ values. We would emphasize that, in the causal-model framework, one *can* learn about both overall effects and intermediate effects from off-the-line cases. The reason why on-the-line cases is simply that they are more *representative* about the universe about which we aim to learn (assuming we want to learn about the population from which the $X,Y$ pattern is drawn). -->

<!-- Finally, we consider $N=2$ case-selection strategies that fix the value of $X$ or the value of $Y$ in both cases, while seeking variation on the other variable. The "fix $X=1$" strategy selects one $X=1, Y=0$ case and one $X=1, Y=1$ case. The "fix $Y=1$" strategy selects one $X=0, Y=1$ case and one $X=1, Y=1$ case.  Across queries, we see that these two strategies generate roughly equivalent opportunities for learning. Interestingly, we also see that these strategies generates significantly *less* learning than selecting on *both* $X$ and $Y$, whether on-the-line or off-the-line.  -->

<!-- Why? When we select on-the-line or off-the-line, we are building in variation in both $X$ and $Y$ across the cases. In contrast, when we select on a value for $X$ or a value for $Y$, while letting the other variable vary, we are building in variation in only one variable. This means that the "signal" we get from the $M$ pattern is likely to be more ambiguous across the cases we end up with. For instance, if we examine an  $X=1, Y=1$ case and an $X=1, Y=0$ case, suppose we observe $M=1$ in both cases. That $M$ pattern is consistent with $X$ having an effect on $M$, but inconsistent with $M$ have an effect on $Y$. And what if, instead, we observe $M=1$ in the first case and $M=0$ in the second case? Then we have evidence of an $M \rightarrow Y$ effect but not of an $X \rightarrow M$ effect. In both scenarios, there is no strong implication for the higher-level $X \rightarrow Y$ causal queries we are interested in.   -->
<!-- We can see this logic reflected in the expected posterior variance reductions for the $ATE$ as compared to the $ATE$ at the individual causal stages. Looking at Figure \@ref(caseselectionvar2), we see that we actually expect to learn a good deal about the $X \rightarrow M$ effect and about the $M \rightarrow Y$ effect from a fix-$X$ or fix-$Y$ strategy. The problem is that we expect rarely to learn about both stages *at the same time* -- hence the limited learning about the $X \rightarrow Y$ effect.  -->

<!-- <!-- We note that the same difficulty would arise if we were to select for variation in $X$ or $Y$. If we select for variation in $X$, for instance, we could end up with an $X=0, Y=1$ case and an $X=1, Y=1$ case, for which $M$ will be uninformative. We could get lucky and end up with an informative pair ($X=0, Y=0$ and $X=1, Y=1$), but a strategy of selection conditional on one variable leaves that matter to chance.  -->




#### Moderator models

What if $M$ appears in the model as a moderator, rather than a mediator, of $X$'s effect on $Y$: $X \rightarrow Y \leftarrow M$? 

Intuitively, we might think that we can learn from examining a moderator because of the fact that a moderator is another potential cause. So, for instance, if we see $M$ constant while $Y$ varies (or vice versa) this suggests evidence against $M$'s effect --- and, we might conjecture, *for* $X$'s effect on $Y$. 

Our simulations, broadly, speak against this logic. We see in Figure \@ref(fig:caseselectionest) that our inferences hardly budge conditional on different realizations of the moderator variable, regardless of case-selection strategy. We also see in Figure \@ref(fig:caseselectionvar) that we get very little reduction in expected posterior variance.^[There seems to be a bit of learning about the probability of causation, but attention to the $y$-axis scale reveals that the reductions in expected posterior variance here are very small.] The key reason, we think, is the non-rivalry of causes. Given no restrictions on $Y$'s nodal types and no prior evidence pertaining to $M$/$X$ interactions, evidence for or against $M$'s effect on $Y$ has little impact on our beliefs about $X$'s effect.

#### Two-path models

We turn now to a two-path model in which $X$ can have both an indirect effect on $Y$ via $M$ and a direct effect on $Y$. We first examine a version with flat priors over all nodal types and then versions in which we impose some monotonicity restrictions and priors over the two paths. We note that, for a model with multiple paths, we may be interested in an additional, mechanismic query, such as the effect operating along a *particular* path.

*Two-path model with flat priors.* In the two-path row of graphs, we see the results of the case-selection exercise for a two-path model with flat priors. In Figure \@ref(fig:caseselectionvar), we see that most case-selection strategies yield no prospect of learning about average effects of $X$ on $Y$ or about the probability that $X$ caused $Y$ in a given case. 

The reason, we think, is the combination of the double pathway and uninformtativeness of priors about intermediate effects. We go in with a flat distribution across nodal types $\theta^M$ and nodal types $\theta^Y$. We then acquire some information on the $X \rightarrow Y$ effect from the $X,Y$ data, but those effects are equally consistent with all possible pathways---direct, indirect, or combined. Suppose we then find evidence that cuts against the operation of the indirect pathway: $M=1$ is observed in two on-the-line cases with different $X$ values. So we downgrade the probability of $\theta^M_{01}$ relative to $\theta^M_{11}$. However, we are still able to preserve our prior on the effect of $X$ on $Y$ by updating our beliefs about $\theta^Y$ in a manner that places greater weight on direct effects. To put the point differently: high flexibility in our lower-level beliefs impedes learning about higher-level quantities from observations at the lower level. This is true regardless of which quadrants we select our cases from.

There is, however, one case-selection strategy that stands out: the selection of two $X=1, Y=1$ cases. The learning here relatively weak, but it is notable that our inferences about the $ATE$ and the probability of (especially negative) causation are affected by what we observe on $M$ across the two cases. It is not always easy to "intuit" one's way through updating on a causal model, and we are not sure why the 2-case $X=1, Y=1$ strategy performs better here. The learning about the $X \rightarrow Y$ causal effects is still minimal, however, even for this "optimal" strategy.

What if instead of the  $X \rightarrow Y$ causal effect, we are interested in the *pathway* along which that effect unfolds? Here $M$ can be informative, as we can see from the results in Figure \@ref(fig:caseselectionvar2), displaying expected reductions in posterior variance for a range of additional queries. Moving down the graph, we first see learning emerge for the intermediate causal effects, of $X$ on $M$ and of $M$ on $Y$. Indeed, even observing $M$ in a single case yields leverage on these two effects (since, for instance, $M=1$ in an $X=1$ case is inconsistent with a negative $X \rightarrow M$ effect in that case). As for our other queries, it is the on-the-line cases that are most informative about the intermediate effects.

We also see that we can learn from $M$ about the pathway through which $X$'s effect on $Y$ operates. Focusing on the indirect effect rows (one conditioning on $X=1$, one conditioning on $X=0$), we see that we need at least 2 cases to learn about these effects. We also see that, for $N=2$, we learn more about them from on-the-line than off-the-line cases. It is striking, however, that we still can learn about the mechanism through which a positive effect unfolds *by studying cases in which only a negative effect is possible*. 

*Two-path model with monotonicity restrictions.* The situation changes considerably if we come in with informative beliefs at the lower level. In the two-path-monotonic set of graphs, we see results for a two-path model in which we have imposed stronger beliefs about the direction of causation at the lower level, ruling out negative $X \rightarrow M$ and negative $M \rightarrow Y$ effects. Now we see in Figure \@ref(fig:caseselectionvar) that we can learn from all of the strategies. Perhaps most striking is that we can learn even just from $N=1$ strategies in this model. Observing $M=0$ in an $X=1, Y=1$ case decisively rules out the indirect pathway as a mechanism of causation, thus making a positive effect overall less likely in the case than if we observed $M=1$. 

We also see that our $N=2$ strategies are all informative. They are generally more weakly informative than for the chain model, reflecting the fact that we are still getting some countervailing updating of beliefs about the direct pathway. But the restrictions make $M$ sufficiently informative -- our beliefs about causation along the indirect pathway can shift by a sufficient magnitude -- that the updating about the direct pathway cannot fully offset it. So evidence for (against) the operation of the indirect pathway registers, on net, as evidence for (against) the operation of a total $X \rightarrow Y$ effect.

In Figure \@ref(fig:caseselectionest), we can also observe asymmetries in the learning that we do not see in the other models. That is, our beliefs move further in absolute-value terms from the baseline given by the pure $X,Y$ data depending on what we find. This asymmetry is driven by the asymmetry of the restrictions. In excluding intermediate negative effects along the indirect path, we have made the search for $M=1$ a "hoop" test for the operation of this path: it *might* be operating if $M=1$ (generating a small upward shift in the causal estimand) but is certainly not operating if $M=0$ (generating a larger downward shift). We derive further information here from the shading -- representing the probability of each data-realization -- which we see is also asymmetrical. In general, we see that the data pattern that would generate the larger shift in beliefs away from the prior is *less* likely to occur (a mathematically necessary feature -- otherwise, we could not logically hold the prior belief that we hold).

Another interesting result is that, under these restrictions, it is the *off*-the-line strategy that holds the potential for the greatest learning. We believe that the reason relates, again, to the asymmetry introduced by the restrictions. In an off-the-line pair of cases, we know that the indirect path cannot be fully operating: any $X \rightarrow Y$ effect here must be negative, but the restrictions only allow for a positive indirect effect. Another way to put this is that an off-the-line case is one in which we start out with the lowest prior on the mediating effects along the indirect path: it's the kind of case in which we least expect an effect of $X$ on $M$ or an effect of $M$ on $Y$. This creates an opportunity for an especially strong "surprise": in particular, if we observe $X=0, M=0, Y=1$ and $X=1, M=1, Y=0$ in the two off-the-line cases, we have found evidence in *favor* of one of the two intermediate effects, the $X \rightarrow M$ effect. Now, this pattern also constitutes evidence against the operation of an $M \rightarrow Y$ effect; but as we have said, we did not particularly expect to find such an effect in off-the-line cases, so this negative finding does not affect our beliefs much. So we see a large boost in our belief about the average effect of $X$ on $M$, a small drop on the estimated average effect of $M$ on $Y$, a considerable boost in our belief in the average indirect effect of $X$ on $Y$, and a concomitant upward shift in our beliefs about the $ATE$. 

Observing $M$ vary in the opposite direction across the two off-the-line cases --- so positively varying with $Y$ but not $X$ --- generates strong downward movement in our beliefs about the indirect effect and the $ATE$. It seems that the hit to the $X \rightarrow M$ effect matters more for the indirect (and total) effect than does the boost to the $M \rightarrow Y$ effect.

Looking on the line, in the monotonic two-path model, also yields opportunities to learn, but weaker ones. If we see $M$ vary in concert with $X$ and $Y$ across the one-the-line cases, we certainly shift our beliefs in indirect effects and the $ATE$ upward. But the shift is smaller than in the off-the-line comparison because these are cases in which we started out with a stronger expectation that the indirect path would be operating and, in turn, that $M$ would covary with $X$ and $Y$. There is simply less opportunity for surprise.

Notably, when we turn to our pathway query (indirect effects), the monotonicity restrictions do not make nearly as a big difference to the possibility of learning as they do for the $X \rightarrow Y$ effects. We also see that the on-the-line strategy performs best for the mediation query, in constrast to the $ATE$ query, as it does for the pathway query in the unrestricted version of the model.

*Two-path model with weak priors on direct path* Finally, we consider a model in which we have no restrictions but adopt priors under which the direct path is believed to be relatively weak. To think about what this means, recall that the limits to learning from $M$ in the two-path model with no informative priors or restrictions derive from the lack of constraint on countervailing updating about the direct path as we learn about the indirect path. When we start with a greater prior weight on the indirect than the direct path, however, we are dampening this countervailing updating. We should expect therefore to see more opportunities for learning from $M$.

As we see in Figure \@ref(fig:caseselectionvar), the priors do not make a large difference, but they do allow for strategies beyond the strategy of selecting two $X=1, Y=1$ cases to provide leverage on the $ATE$ and, even more so, on the probability of positive and negative causation. The second-most promising strategy for the $ATE$ is the on-the-line strategy, followed by the off-the-line strategy. As with the simple chain model, we see that the off-the-line strategy performs best for the negative probability of causation query, while the on-the-line strategy performs best for the probability of positive causation --- and, in fact, dominates the $X=1, Y=1$ strategy for that query.


<!-- FLAG: Cannot yet see why we learn more from the restricted 2-path model than from the flat 2-path. Yes, we can learn from n=1 about the indirect path. But in both cases, we can still have compensating updating on the direct path for anything we learn about the indirect path. In fact, the prior probability of positive effects on each path is very similar across the two models. Ignoring interactions, for indirect path, b share is like 1/8 in flat model, and 1/9 in restricted, vs. 1/4 for direct path. Is it just that the learning about the indirect path is stronger with restricted priors because M starts out as informative about the indirect path? So the learning about the indirect path more easily swamps the prior? -->

<!-- FLAG: Incorporate as needed above and below any implications of Jeffreys priors. -->

<!-- FLAG: Let's move the 2-pos strategy to the right of the 2 on strategy in the graphs.  -->




## Wide or Deep

We continue our journey through the space of research-design choices by thinking about the tradeoff between intensive (deep) and extensive (wide) analysis. Suppose that we have identified those cluesthat will be most informative and those cases in which it would be most valuable to conduct process tracing, given our beliefs about the world. A further question that we face is the quintessential dilemma of *mixing* methods: what mixture of quantitative and qualitative evidence is optimal? We have argued in in Chapter \ref(mixing) that the distinction between quantitative and qualitative inference is, in a causal-model framework, without much of a difference. But here we frame a more precise question: given finite resources, how should we trade off between studying a larger number of cases at a given level of intensiveness, on the one hand, and drilling down to learn more intensively about some subset of the cases in our sample? How should we decide between going "wide" and going "deep"?

Just as with the selection of clues and cases, how much we should expect to learn from going wide versus going deep will *depend* on how we think the world works, as expressed in the causal model with which we start and as shaped by the by the data that we have seen at the point of making the wide-versus-deep decision. 

As in our discussion of case-selection, we examine here queries commonly associated with large-$N$, quantitative strategies of analysis (such as average treatment effects) as well as queries commonly associated with more case-oriented, qualitative approaches (queries about causal pathways and about causal effects at the case level). The analysis in this chapter makes clear the opportunities for integration across these lines of inquiry. We show that investing in-depth process tracing will sometimes make sense even when one aims to learn about average effects in a population. Likewise, collecting $X, Y$ data can sometimes help us draw inferences that will aid in case-level explanation. Particular kinds of case-level information can teach us about populations, and understanding population-level patterns can help us get individual cases right. 


###  Walk-through of a simple comparison

To build up our intuitions about how the optimal mix of strategies might depend on how the world works, let us explore a simple comparison of wide and deep strategies. We focus here on the question of how much we can learn from drilling deeper, given an initial set of $X,Y$ data and beliefs about the world.

Imagine a world in which we have a large amount of data on $X$ and $Y$ (2000 observations), and we see that $X$ and $Y$ are perfectly correlated. We might be tempted to infer that $X$ causes $Y$. And, if $X$ were randomly assigned, then we might be able to justify that inference. Suppose, however, that our data is observational and, in particular, we were aware of an observed confound, $M$, that might determine both $X$ and $Y$. In that situation, the effect of $X$ on $Y$ is not identified. As shown by @manski1995identification,  different priors could support beliefs about that effect anywhere between 0 and 1. From @pearl2009causality, we also know that if the right causal model is $X \rightarrow Y \leftarrow M \rightarrow X$, then data on $M$ would allow the effect of $X$ on $Y$ to be identified. We could estimate the effect of $X$ on $Y$ for $M=0$ and for $M=1$ and take the average. 

Say though that we had a choice between gathering a lot *more* data on $X$ and $Y$ or gathering a *little* data on $M$ for a subset of cases---just 20 in this illustration. Which should we do? Is 20 cases enough to probe the causal model to see whether the correlation between $X$ and $Y$ is spurious or not?

We get an intuition for the answer by imagining the inferences we might draw in 3 extreme cases and compare these to the base case. Figure \@ref(fig:widedeepXYMX) illustrates. The figures are generated by forming a model with $X\rightarrow Y \leftarrow M \rightarrow Y$, strong priors that $\Pr(M=1)=0.5$, and flat priors on all other nodal types. In other words, in our priors we think that $M$ is equally likely to be a 0 or 1 but do not make assumptions about how it is related to $X$ and $Y$. We first update the model with the $X,Y$ data --- and then choose between going wider and going deeper. 




```{r widedeepXYMX, echo = FALSE, fig.width = 10, fig.height = 8, fig.cap = "Posteriors on the ATE given different wide or deep data patterns."}

if(do_diagnosis){
  
  M <- make_model("X -> Y <- M -> X") %>% set_priors(node = "M", alpha = 100)

  dfM2 <- dfM1 <- df <-
    data.frame(X = rep(0:1, 1000), Y = rep(0:1, 1000), M = NA)  
  
  dfM1$M[1:20] <- rep(0:1, each = 10)
  dfM2$M[1:20] <- rep(0:1)
  
  out <- list(
    base = update_model(M, df, keep_transformed = TRUE, iter = 5000) %>% 
      query_distribution("Y[X=1]  - Y[X=0]", using = "posteriors"),
    wide = update_model(M, rbind(df, df), keep_transformed = TRUE, iter = 5000) %>% 
      query_distribution("Y[X=1]  - Y[X=0]", using = "posteriors"),
    deep1 = update_model(M, dfM1, keep_transformed = TRUE, iter = 5000) %>% 
      query_distribution("Y[X=1]  - Y[X=0]", using = "posteriors"),
    deep2 = update_model(M, dfM2, keep_transformed = TRUE, iter = 5000) %>% 
      query_distribution("Y[X=1]  - Y[X=0]", using = "posteriors"),
        deep1check = update_model(M, dfM1[1:20,], keep_transformed = TRUE, iter = 5000) %>% 
      query_distribution("Y[X=1]  - Y[X=0]", using = "posteriors")
  )
  
  write_rds(out, "saved/ch13_wide_deep_histograms.rds")
  }

out <- read_rds("saved/ch13_wide_deep_histograms.rds")
par(mfrow = c(2,2))
hist(out[[1]], main = "(1) 2000 XY observations", xlab = "ATE", xlim = 0:1, freq = FALSE)
hist(out[[2]], main = "(2) Add another 2000 XY observations", xlab = "ATE", xlim = 0:1, freq = FALSE)
hist(out[[3]], main = "(3) Add 20 M observations (uncorrelated)", xlab = "ATE", xlim = 0:1, freq = FALSE)
hist(out[[4]], main = "(4) Add 20 M observations (correlated)", xlab = "ATE", xlim = 0:1, freq = FALSE)

```


Panel 1 in  Figure \@ref(fig:widedeepXYMX) shows our posterior distribution over the average causal effect from observation of the base data: 2000 cases with $X$ and $Y$ perfectly correlated. The distribution is quite wide, despite the strong correlation, because the posterior includes our uncertainty over the nature of confounding.  Our estimate for the $ATE$ is `r round(mean(out$base), 2)` but with a posterior standard deviation of  `r round(sd(out$base), 4)`. There is positive weight on all positive value of the $ATE$.

How can we improve on this estimate?

One possibility would be to go wide and collect $X,Y$ data on an additional 2000 cases. Panel 2 displays our posterior on the average causal effect with the addition of these 2000 cases. We assume that the new data also display a perfect $X,Y$ correlation, like the first set of data. Again, we could not imagine data that more strongly confirms a positive relation, and now we have twice as much of it. What we see, however, is that investing in gathering data on 2000 additional cases does not help us very much. The mean of our posterior on the $ATE$ is now `r round(mean(out$wide), 2)`, with a standard deviation of  `r round(sd(out$wide), 4)`. So the updating is very slight.

Suppose that, for the cost of gathering $X,Y$ data on an additional 2000 cases, we could drill down on a random subset of 20 of the original 2000 cases and observe $M$ in those cases. What might we learn? 

Because we start out with a flat prior on how $M$ will relate to $X$ and $Y$, we display inferences for two possible realizations of that pattern. In Panel 3, we show the updating if $M$ turns out to be uncorrelated with both $X$ and $Y$. The mean of our posterior on the $ATE$ now rises to `r round(mean(out$deep1), 2)`, and the posterior standard deviation shrinks dramatically, to  `r round(sd(out$deep1), 4)`. Greater depth in a relatively small number of cases is enough to convince us that the $X,Y$ relationship is not spurious.  

Panel 4 shows inferences from the same "going deep" strategy but where the $M$ turns out to be perfectly correlated with $X$ and $Y$. Now our estimate for the ATE shifts downward to `r round(mean(out$deep2), 2)`, with a posterior standard deviation of  `r round(sd(out$deep2), 4)`. 

In other words, in this setup, what we observe from our "going deep" strategy can have a big impact on our inferences. One reason we stand to learn so much from process-tracing so few cases is that the process-tracing speaks to relationships about which we start out knowing so little: $M$'s effect on $X$ and $M$'s effect on $Y$, effects on which the $X,Y$ data themselves shed no light. 

It is also interesting to note that we cannot learn as much by updating *only* using information from the 20 cases for which we have full $X$, $M$, $Y$ data. Were we to use only the subset with this complete data --- ignoring the other 1880 cases --- and observe $M$ to be uncorrelated with $X$ and $Y$, the mean of our posterior on the ATE would be `r round(mean(out$deep1check), 2)` with a posterior standard deviation of  `r round(sd(out$deep1check), 4)` (not graphed). The breadth provided by those 1880 $X,Y$-only cases thus adds a great deal. While observing an uncorrelated $M$ in 20 cases allows us to largely rule out $M$ as a cause of any $X,Y$ correlation, observing a strong $X,Y$ correlation over a large number of cases provides evidence that $X$ in fact affects $Y$.

We use this example to illustrate a simple but stark point: there will be situations in which the expected gains from collecting more data on the same cases and from collecting the same data on more cases will be different, sometimes very different. The model and the prior data shape the tradeoff. In this particular setup, it is the confounding together with the large number of prior $X,Y$ observations that makes depth the better strategy. Once we have learned from 2000 $X,Y$ observations, data of the same form from more cases will not change beliefs. Yet going deep---even if only in a few cases---provides information on parameters we know nothing about, helping us interpret the $X,Y$ correlation in causal terms.


### Results from simulations

While the results in the last section are striking, they depend upon particular realizations of the data under each strategy. When selecting strategies we, of course, do not know how the data will turn out. Our problem becomes, as in the case-selection analyses, one of figuring out the *expected posterior variance* from different strategies. 

```{r morn1, fig.width = 6, fig.height = 5, echo = FALSE, message = FALSE, fig.cap = "Distributions of posterior variances from many simulated datasets"}
path = "rep/wide_deep_2"
results <- 
  list(
    Chain = read_rds(paste0(path, "/chain.rds")),
    Confounded = read_rds(paste0(path, "/confounded.rds")),
    Confounded_large_n = read_rds(paste0(path, "/confounded_large.rds")),
    Restricted = read_rds(paste0(path, "/restricted.rds")),
    Base = read_rds(paste0(path, "/base.rds")),
    Chain_homogeneous = read_rds(paste0(path, "/chain_homog.rds")),
    Chain_with_priors = read_rds(paste0(path, "/chain_with_prior.rds"))
) %>% bind_rows(.id = "Model") %>%
    mutate(Model = ifelse(Model == "Confounded_large_n", "Confounded", Model))

r2 <- results %>% filter(Model == "Confounded" & wide_n %in% c(800, 1600) & deep_n %in% c(0, 100) & Case.estimand == "FALSE" & Given == "-")

r2 %>%
  ggplot(aes(sd^2)) +
  geom_histogram() + 
  facet_grid(deep_n ~ wide_n, labeller = label_both) + xlab("Posterior variance on the ATE")
```


```{r morn2, echo = FALSE, message = FALSE}
r2 %>% group_by(wide_n, deep_n) %>% summarize(expected = mean(sd^2))  %>% kable(caption = "Expected posterior variance on the ATE", digits = 4)

```


The more general, simulation-based approach that we introduce here is parallel to the approach for case-selection. The steps of this procedure are:

1. We posit a causal model, with any priors or restrictions.
2. We specify the data we have in hand. Those should be data on some but not all nodes in the model for some number of cases.
3. We then specify a set of mixing strategies to assess. A strategy, in this context, is defined as any combination of collecting data on the same nodes for a given number of additional cases (randomly drawn from the population) and collecting data on new nodes for a random sample of the original set of cases.
4. For each strategy, we define the set of possible data realizations. Whereas for case-selection, the structure of the possible datasets will be the same for all strategies, possible data patterns in wide-versus-deep analyses involve much greater complexity and will vary in structure across strategies.
5. As for case-selection, we use the model and prior data to calculate the probability of each data possibility under each strategy.
6. Again, as for case-selection, we update the model using each possible data pattern to derive a posterior distribution. 
7. We then average the posterior distributions across the possible data patterns under a given strategy, weighted by the probability of each data pattern.

We illustrate the last two steps in Figure \@ref(fig:morn1), using a similar setup to the one above. We again posit the same $X \rightarrow Y \leftarrow M \rightarrow X$. We assume that we have started with $X,Y$ observations for 800 cases. We then choose among drawing another 800 cases for $X,Y$ analysis, going deeper in 100 of our cases, or both. What Figure \@ref(fig:morn1) shows is the distribution of *posterior variances* that we obtain from many possible data draws under each strategy.  In the top row, we are comparing two levels of breadth, with no depth. Moving to the bottom row, we introduce depth in 100 cases, for each level of breadth. We see here a distinct shift in the distributon to the left --- meaning, a reduction in the expected posterior variance -- when we drill down to gather data on $M$. In contrast, we see no such shift in the expected posterior variance when we increase the number of cases for which we gather data on $X$ and $Y$. For greater precision, we report the expected value for each distribution of posterior variances in Table \@ref(tab:morn2). The pattern here is consistent with the analysis above, although here we have averaged across all *possible* data patterns arising from the strategies, rather than focusing on just a couple of data possibilities. 

This approach can be applied to *any* model, with *any* query, and *any* prior set of data. 

We now explore strategies for a range of models and present the results in compact form in Figure \@ref(morn2). For the purposes of this illustration, we focus on the $ATE$ and the probability of positive causation as queries, and the results of the exercise would likely differ for different queries. On each graph, we plot 3 lines, each representing a different number of "wide" ($X,Y$-data only) cases, ranging from 100 to 1600. Each line then shows how the expected posterior variance changes as we add "depth," going from observing $M$ in 0 cases to observing it in 50 and then 100 cases.

Starting with the chain model ($X \rightarrow Y$), we see that there are gains to be reaped from both breadth and depth, for both queries. We can also get a sense of the nature of the tradeoffs. One thing we observe is that, for the $ATE$, the gains to breadth are greater the less depth we have. Once we have process-traced 100 cases, the difference in the posterior variance we expect between having $X,Y$ data only on those 100 cases and having $X,Y$ data on an additional 300 cases is very modest compared to the situation in which we have done no process tracing. Likewise, the gains to depth diminish with breadth: the slope of the depth line flattens as we add $X,Y$ cases. Further, there appear to be diminishing returns to each strategy on its own terms: note, for instance, how the move from 50 to 100 process-tracing cases deliver less expected learning than does the move from 0 to 50.

We can also how the tradeoff varies by estimand. For the $ATE$, the gains to going deep in an additional 50 cases look roughly equivalent to those of expanding the dataset by 100 cases. However, for the probability of positive causation, the scales tip quite dramatically in favor of depth: The gains to depth in 50 cases are more than double the gains of broadening by as many as 300 cases. 

```{r morn3, fig.width = 10, fig.height = 7, echo = FALSE, message = FALSE, fig.cap = "Expected posterior variance over multiple models with multiple data strategies."}

read_rds("saved/wide_deep_fig_book.rds")

```


 
In the chain model we see similar posterior variance from the following three strategies: 

1. Wide: 400w
2. Deep: 100w + 100d
3. Mixed: 200w + 50 d

For difference prices of $m$ and $d$ any of these may be best. In particular if the cost of w is $1: 

* strategy 2, mixed is better if deep costs between $2 and $4
* if deep costs more than $4 then go all wide
* If deep costs < $2 then go all deep


Common set

* Chain
* Chain monotonic
* Moderator
* Moderator monotonic
* Two path
* Two path monotonic
* Confounded
* Confounded monotonic

Mediation estimand for two path models

## Principles

*  Qualitative and quantitative data can act as partial substitutes for assessing causal effects.
*  The *relative* marginal gains from going wider and going deeper vary with the study design. 
* Optimal strategies might involve going deep in a subsample of cases only.




