# Evaluating models {#evaluation}

***

Model based inference takes the model seriously. But deep down we know that all of these models are wrong, in myriad ways. We examine strategies for figuring out whether a model is likely doing more harm than good.

***


```{r packagesused15, include = FALSE}
source("_packages_used.R")
do_diagnosis = FALSE
library(DeclareDesign)
```


Throughout this book we have maintained the conceit that you believe your model. But it is also obvious that even the most non-parametric-seeming models depend on substantive assumptions and that these are almost certainly wrong. The question then is not how much you believe your model (or whether you really believe what you say you believe) but whether your model is useful is some sense. How can we evaluate the usefulness of our models?


## Five Strategies

Sometimes a model is just not able to fit the observed data well. 

Imagine a situation in which researchers believe that the effect of $X$ on $Y$ runs entirely through $M$, positing a model of the form $X \rightarrow M \rightarrow Y$. Imagine that, however, that the true model is $X \rightarrow  M \rightarrow Y \leftarrow X$, with $X$ *only* having a direct effect on $Y$. The problem with the posited model, then, is that it is too restrictive: it does not allow for a direct effect that is in fact operating.

<!-- AJ: "Restrictedness" seems a bit off here insofar as the "unrestricted" model is also restricted in essentially assuming no operative $X -> M$ arrow, as well as no negative direct effects.  -->

We are perfectly able to update using this overly restrictive model and the data --- but the updated model can produce wildly inaccurate causal inferences.  In Figure \@ref(modelsch15), we show the results of an analysis in which the data are generated from a true model of the form $X \rightarrow M \rightarrow Y \leftarrow X$, with an average effect of $X$ on $Y$ of $1/3$ but no effect of $X$ on $M$ or $M$ on $Y$, and thus no indirect effect. 

In the figure, we show the inferences on the average treatment effect for two different updated models, both starting with flat priors: the more restricted, $X \rightarrow M \rightarrow Y$ model and the less restricted,
$X \rightarrow M \rightarrow Y \leftarrow X$ model. We represent the true average effect with the vertical line in each graph.

As we can see, the more restrictive model that excludes direct effects generates a posterior credibility interval that excludes the right answer. So, if we go into the analysis with the restricted model, we have a problem. 

But will we notice?

In the remainder of this section, we explore a range of diagnostics that researchers can undertake to evaluate the usefulness of their models or to compare models with one another: checking assumptions of conditional independence built into a model; attending to computational clues; checking the model's fit; comparing model likelihoods; and "leave-one-out" cross-validation.

```{r modelsch15, echo = FALSE}
n = 200

if(do_diagnosis){
  
 unrestricted_model <- 
    make_model("X -> M -> Y <- X") %>%
    set_parameters(
      statement = c("M[X=1] != M[X=0]", "(Y[X=1] < Y[X=0])"),
      parameters = c(0, 0)
      ) 

 simulate_data(unrestricted_model, using = "parameters", n = n) %>%
    write_rds("saved/ch15_XMYdata.rds") 
  
 make_model("X -> M -> Y") %>%
   update_model(data, keep_fit = TRUE) %>%
   write_rds("saved/ch15_restricted_model.rds") 

 unrestricted_model %>%
   update_model(data, keep_fit = TRUE) %>%
   write_rds("saved/ch15_unrestricted_model.rds") 

}

data               <-   read_rds("saved/ch15_XMYdata.rds") 
restricted_model   <-   read_rds("saved/ch15_restricted_model.rds") 
unrestricted_model <-   read_rds("saved/ch15_unrestricted_model.rds") 

if(do_diagnosis){
  
  ateXY1 <- query_distribution(restricted_model, "Y[X=1] - Y[X=0]", using = "posteriors")
  ateXY2 <- query_distribution(unrestricted_model, "Y[X=1] - Y[X=0]", using = "posteriors")

 write_rds(data.frame(restricted_ate = ateXY1, unrestricted_ate = ateXY2),
           "saved/ch15_ates.rds")
}

ates <- read_rds("saved/ch15_ates.rds") 

```

```{r 15badmodels, echo = FALSE, fig.cap = "A restricted model yields a credibility interval that does not contain the actual average effect."}
par(mfrow = 1:2)
hist(ates$restricted_ate, xlim = c(-.1, .6), main = "Restricted model", xlab = "Effect of X on Y") 
abline(v = .33, col = "red")
hist(ates$unrestricted_ate, xlim = c(-.1, .6), main = "Unrestricted model", xlab = "Effect of X on Y")
abline(v = .33, col = "red")

```


<!-- * $X$ correlates with $Y$ -->
<!-- * $X$ does not correlate with $M$  -->
<!-- * $M$ does not correlate with $Y$.  -->

<!-- These data are inconsistent with the model: under this model, if $X$ doesn't cause $M$ and $M$ doesn't cause $Y$, then there is no other way for $X$ to cause $Y$.  -->

### Check conditional independence

First, even before engaging in updating, we can look to see whether the data pattern is consistent with our causal model. In particular, we can check whether there are inconsistencies with the Markov condition that we introduced in Chapter 2:  that every node is *conditionally independent* of its non-descendants, given its parents. In this case, if the restricted model is right, then given $M$, $Y$ should be independent of $X$. 

Is it? 

One way to check is to assess the covariance of $X$ and $Y$ given $M$ in the data. Specifically, we regress $Y$ on $X$ for each value of $M$, once for $M=1$ and again for $M=0$; a correlation between $X$ and $Y$ at either value of $M$ would be problematic for the conditional independence assumption embedded in the restricted model. 

Note that this form of diagnostic test is a classical one in the frequentist sense: we start by hypothesizing that our model is correct and then ask whether the data were unlikely given the model.



```{r, echo = FALSE}
rbind(
  estimatr::lm_robust(Y~X, data = filter(data, M==0)) %>% tidy %>% filter(term == "X"), 
  estimatr::lm_robust(Y~X, data = filter(data, M==1)) %>% tidy %>% filter(term == "X")) %>%
  mutate(M = 0:1) %>%
  select(M, estimate, std.error, p.value) %>% 
  kable(digits = 3, caption =  "Regression coefficient on $X$ given $M=0$ and $M=1$")

```

We report the regression coefficients on $X$ in the table below. It is immediately apparent that we have a problem. At both values of $M$, there is a strong correlation between $X$ and $Y$, evidence of a violation of the Markov condition implied by the restricted model.^[In applying the Markov condition, we also need to take into account any unobserved confounding. For instance, suppose that there was an unobserved confounder of the relationship between $M$ and $Y$ in the $X \rightarrow M \rightarrow Y$ model. Then we would *not* expect $Y$ to be independent of $X$ conditional on $M$. Intuitively, the data pattern we described could be consistent with such a model in which on average $X$ does not cause $M$ but still *in those cases in which $X=M$* we have $Y=X$. Another way to think about this is that $M$ now acts as a collider between $X$ and another unobserved cause of $Y$; so conditioning on $M$ introduces a correlation between $X$ and this unobserved cause, and thus between $X$ and $Y$.] 

<!-- AJ: I do not follow the sentence starting with "Intuitively" in the above. -->




ROUGH TEXT:

approach 2 -- say actual confound is q~=0; but model  assumes q = 0. Draw data from priors, draw data; given data type (001, 100 etc) plot (a) the posterior distribution under no confounding nad (b) the distribution of estimands that gave rise to the data.  


(Verma and Pearl, 1990) identify conditions under which we can check some independence assumptions.


 Pearl (1995) gives conditions for assessing for discrete data whether $Z$ has a  direct effect on $Y$. (Involves inequalities)
 
Evans (Graphical methods for inequality constraints in marginalized DAGs) generalizes the instrumental inequality.
 


### Computational clues

Second we may be lucky and run into computation issues. In this example there is a good chance that when you run the restricted  model `stan` will throw an error (`Bulk Effective Samples Size (ESS) is too low, indicating posterior means and medians may be unreliable.`)


<!-- AJ: Needs elaboration: why is this error a function of problems with the model? -->
<!-- MH: I agree; do feel free to look into it; I will try -->
<!-- AJ: I have done some digging and not found anything helpful on this -->


### Bayesian $p$ value: Is the data unusual given your model?


THIS IS NOT COMPARING TO ANOTHER MODEL
QUESTION: DO CHARACTERISTICS FO THE ACTUAL DATA LOOK LIKE WHAT SORT OF DATA WE EXPECT TO SEE GIVEN OUR MODEL

Approaches using simulated data from the posterior predictive distribution  are described in @gabry2019visualization.

Using tools in the `bayesplot` package, the graphs below show how typical the data we observe is for the model that correctly assumes a direct effect and the model that incorrectly excludes it. 

First we look just at the distribution of the outcome $Y$ to see how the actual distribution compares to the predicted distribution. We see here not too much difference. 

<!-- AJ: I am having trouble seeing the progression here. -->

```{r ch15simulations, echo = FALSE}

if(do_diagnosis){
  
 sims <- 1000
 n <- nrow(data)
 
 # Outcome prediction
 ############################################# 
 t(replicate(sims, make_data(restricted_model, n = n, 
                        param_type = "posterior_draw")$Y)) %>%
  write_rds("saved/ch15_replicates_restricted.rds") 

  t(replicate(sims, make_data(unrestricted_model, n = n, param_type = "posterior_draw")$Y)) %>%
  write_rds("saved/ch15_replicates_unrestricted.rds") 
  
  # Distribution of cor X, Y in data
  #############################################
  replicate(sims, {
    df <- make_data(restricted_model, n = n, param_type = "posterior_draw")
    cor(df$X, df$Y)}) %>%
  write_rds("saved/ch15_replicates_restricted_b.rds") 

    replicate(sims, {
    df <- make_data(unrestricted_model, n = n, param_type = "posterior_draw")
    cor(df$X, df$Y)}) %>%
  write_rds("saved/ch15_replicates_unrestricted_b.rds") 

  }


```


```{r, fig.cap = "Distribution of" }
x <- data.frame(
  unrestricted = read_rds("saved/ch15_replicates_unrestricted_b.rds"),
  restricted   = read_rds("saved/ch15_replicates_restricted_b.rds") 
)
reshape2::melt(x) %>%
ggplot(aes(x=value, fill=variable)) + geom_density(alpha=0.25)
```


How does the researcher's model do?

```{r distributio113, eval = FALSE, fig.cap= "Posterior distribution of test statistics under researcher's model"}

replicates_restricted <-   read_rds("saved/ch15_replicates_restricted.rds") 

bayesplot::ppc_stat(data$Y, replicates_restricted)

```

How would we evaluate the less constrained ("true") model?

```{r distribution2, echo = FALSE, eval = FALSE, fig.cap = "Distribution of $Y$ under the (data updated) true model"}
# Updated alternative model
replicates_unrestricted <-   read_rds("saved/ch15_replicates_unrestricted.rds") 
bayesplot::ppc_stat(data$Y, replicates_unrestricted)
```


```{r ch15predY, fig.cap = "Predicted outcomes"}

x <- data.frame(
  unrestricted = read_rds("saved/ch15_replicates_unrestricted.rds") %>% apply(2, mean),
  restricted   = read_rds("saved/ch15_replicates_restricted.rds") %>% apply(2, mean) 
)
reshape2::melt(x) %>%
ggplot(aes(x=value, fill=variable)) + geom_density(alpha=0.25)

```


<!-- Bayesian information criterion (BIC)^[$BIC = \ln(n)k - 2\ln(\hat{L})$ where $\hat{L}$ is the maximized likelhood, $k$ is the number of parameters, and $n$ the number of data points.] -->
<!-- BIC involves a penalty for more parameters -->


<!-- ```{r, echo = FALSE} -->

<!-- if(do_diagnosis) { -->
<!-- observed_data <- collapse_data(data, restricted_model) -->

<!-- likely <- function(model, s, posterior = TRUE) { -->
<!--   if(posterior)  pars <- model$posterior_distribution[1:s,] -->
<!--   if(!posterior) pars <- get_prior_distribution(model, s)[1:s,] -->
<!--   L <- apply(pars, 1,  function(par) make_data_probabilities(model, pars = par, observed_data, normalize = FALSE)) -->
<!--   mean(L) -->
<!--   } -->

<!-- s <- 500 -->
<!-- L_prior         <- likely(restricted_model, s, posterior = FALSE) -->
<!-- L_restricted    <- likely(restricted_model, s) -->
<!-- L_unrestricted  <- likely(unrestricted_model, s) -->

<!-- df <- data.frame(L = (c(L_unrestricted/L_prior, L_restricted/L_prior, L_unrestricted/L_restricted))) -->

<!-- rownames(df) <- c("Unrestricted / Prior", "Restricted / Prior", "Unrestricted / Restricted") -->

<!-- write_rds(df, "saved/ch15_likelihoodsdf.rds")  -->
<!-- } -->

<!-- df <- read_rds("saved/ch15_likelihoodsdf.rds")  -->

<!-- kable(df, col.names = "Bayes factors", digits = 2, caption = "Posterior odds: the relative likelihood of one model over another") -->
<!-- ``` -->

<!-- Compare likelihoods of the data under different models -->
<!-- Check look package for rstan -->


### Leave-one-out cross-validation

A further class of model-validation methods involve cross-validation. Rather than asking how well the updated model predicts the data used to update it, cross-validation uses the data at hand to estimate how well the model is likely to predict data that have not yet been seen. 

One approach to cross-validation is the "leave-one-out" (LOO) algorithm. In a LOO approach, we update the model using all data points except for one; ask how well the model performs in predicting the left-out observation; repeat for every data point in the dataset; and aggrgeate predictive performance across all iterations. 

In the standard implementation, the measure of predictive success for each data point is the residual error, typically on an outcome variable. In a \CausalQuery\ framework, however, we may be interested in predictions over the joint realization of all nodes, not just a single "outcome." Thus, instead of calculating residual error as the measure of predictive success for each data point, we calculate the posterior probability of each data point, using the model updated with all of the other observations. We then aggregate by taking the product of the posterior probabilities across all data points.

The LOO estimate of out-of-sample predictive fit, for a dataset with $n$ observations, is then:

$$\prod_1^np(y_i|y_{-i}, \text{model})$$
where $y_{-i}$ is the data pattern with observation $y_i$ left out, and $y_i$ represents the values of all nodes of interest for observation $i$. 

First, we implement LOO cross-validation of the two models using 200 data points generated from the unrestricted model. The likelihood of the data under the restricted model is `r  formatC(loo_restricted$loo_likelihood, format = "e", digits = 2)`
the likelihood if  `r  formatC(loo_unrestricted$loo_likelihood, format = "e", digits = 2)` under the unrestricted models. Thus, the unrestricted model represents an estimated improvement in out-of-sample prediction on the order of   `r  formatC(loo_unrestricted$loo_likelihood/loo_restricted$loo_likelihood, format = "e", digits = 2)`. 

We can visualize the pattern in Figure XXXX, where we plot the likelihood of each possible data type under the restricted model against the likelihood of that data type under the unrestricted model. Looking at the scales of the two axes---which is much more compressed on the horizontal than on the vertical---one can see that the restricted model is not able to differentiate as much across the data types as the unrestricted. 

<!-- AJ: What are these likelihoods exactly? They are informed by updating, I assume? But updating using what data exactly -- what's left out? Generally unsure if my text above is right.-->

Notably, the restricted model is not able to "learn" from the data about the (*in fact*, operative) relationship between $X$ and $Y$. We can see that, for any given $X, M$ combination, the two possible values of $Y$ are predicted with essentially the same likelihood. The restricted model also seems to have "learned" that different values $X,M$ combinations are differentially likely, even though they are not under the true model. The unrestricted model, on the other hand, essentially divides the data types into two groups: those with a positive $X,Y$ correlation and those with a negative $X,Y$ correlation and has correctly (given the true model) learned that the former is more likely than the latter.

In Figure XXXX, we then see how the likelihoods of each data type line up with the actual count of each data type. As we can see, the unrestricted model updates to likelihoods that fit the actual data pattern well while the restricted model does not.

<!-- AJ: The x-axis labels confuse me. What's the meaning of a data count for a restricted or unrestricted model? The data are what they are.  -->







```{r}
#Leave one out

loo_lik <- function(data, model, sims){
  
  short_data <- collapse_data(data, model)
  rows       <- 1:nrow(short_data)
  
  case_likelihood <- lapply(rows, 
        function(j) {
        loo_data <-  mutate(short_data, count = ifelse(count>0 & rows == j, count-1, count))
        loo_left <-  mutate(short_data, count = ifelse(count>0 & rows == j, 1, 0))
        loo_post <-  update_model(model, loo_data, data_type = "compact")
        loo_prob <-  apply(loo_post$posterior_distribution[1:sims, ], 1, function(pars)
                     get_data_probs(model, loo_left, parameters = pars)[j])
        mean(loo_prob)}
        )
  
  list(data = short_data, model = model, case_likelihood = case_likelihood, 
       loo_likelihood = prod((case_likelihood %>% unlist)^(short_data$count)))
  }
  
if(do_diagnosis){
loo_lik(data, restricted_model, 1000) %>% write_rds("saved/ch15_loo_restricted.rds")
loo_lik(data, unrestricted_model, 1000) %>% write_rds("saved/ch15_loo_unrestricted.rds")
}

loo_restricted   <- read_rds("saved/ch15_loo_restricted.rds") 
loo_unrestricted <- read_rds("saved/ch15_loo_unrestricted.rds")  

plot(loo_restricted$case_likelihood, loo_unrestricted$case_likelihood)
text(loo_restricted$case_likelihood, loo_unrestricted$case_likelihood, loo_restricted$data$event)

par(mfrow = c(1,2))
plot(loo_restricted$data$count, loo_restricted$case_likelihood, col = "red")
text(loo_restricted$case_likelihood, loo_unrestricted$case_likelihood, loo_restricted$data$event)

plot(loo_unrestricted$data$count, loo_unrestricted$case_likelihood)

with(loo_restricted,
     {
  plot(loo_restricted$data$count, case_likelihood, col = "red")
  text(case_likelihood, case_likelihood, loo_restricted$data$event)
     }
)

```

DO THE SAME BUT THE OTHER WAY AROUND, WHERE DATA COMES FROM TEH RESTRICTED MODEL

ALS DO WITH CONFOUNDING X<-->Y X->M->Y





### Sensitivity: Qualitative inference

You need to provide priors in order to get Bayesian updating off the ground. But seeing your priors reported as hard cold numbers may make them feel like forced confessions. What if you just don't believe them?

Even in this case you might still engage in a form of "qualitative inference." There is a literature on probabilistic causal models that assesses the scope for inferences when researchers provide ranges of plausible values for parameters (perhaps intervals, perhaps only signs, positive negative, zero), rather than specifying a probability distribution. For a comprehensive treatment of qualitative algebras, see @parsons2001qualitative. Under this kind of approach, a researcher might willing to say that they think some probability $p$ is not plausibly  greater than .5,  but unwilling to make a statement about their beliefs about where in the $0$ to $0.5$ range it lies. Such incomplete statements can be enough to rule our classes of conclusion.

Consider first process tracing models for which you are unsure Say for instance in our running example 

<!-- AJ: Am having some trouble reading off the code what exactly is varying along the axes of the graph. Seems to be something related to M10, M01, Y10, Y01, but not sure how this maps to the graph. Can you explain? -->


```{r, fig.cap = "Possible inferences on `X caused Y` given observation of M=0 and M=1 given different possible models. The highest inference when M=0 is lower than the lowest inference when M=1", echo = FALSE}

model <- make_model("X -> M -> Y") 

adverse1 <- adverse2 <- seq(0, .25, .05)

out <-  
  sapply(adverse1, function(a1) {
      sapply(adverse2, function(a2) 
  
  query_model(
  set_parameters(model, 
                 param_names = c("M.10", "M.01", "Y.10", "Y.01"), 
                 parameters = c(a1, .5+a1, a2, .5+a2)),
  query = c("Y[X=1] > Y[X=0]"),
  given = c("X==1 & Y==1 & M==0", "X==1 & Y==1 & M==1"),
  using = "parameters",
  expand_grid = TRUE)$mean, simplify = FALSE)})
out2 <- do.call(rbind, out) %>% data.frame
names(out2) <- c("M0", "M1")
out2 <- arrange(out2,M0, M1)

plot(out2$M1, xlim = c(0,1), type = "n", xlab = "possible ranges", ylim = c(0, nrow(out2)), ylab = "")
segments(out2$M0, 1:nrow(out2), out2$M1, 1:nrow(out2), col = "red")

```

- A graph showing how some conclusions changes as we relax one of the restrictions.

How much do our conclusions depend on qual restrictions?

- How do conclusions differ if we drop all restrictions



 


## Evaluating the Democracy-Inequality model

** GENERATE A TABLE SHOWING HOW PIMD DOES ON 5 CRITEREA ** 



### Prior check
In a second iteration of the analysis, we show what happens if we loosen the monotonicity restriction on $I$'s effect on $M$. Here we  consider negative effects of $I$ on $M$ *unlikely*, rather than impossible, and we consider null and positive effects somewhat likely. We refer to these priors as "quantitative priors" in the sense that they place a numerical value on beliefs rather than a logical restriction. Here, we set our prior on $\theta^M$ as: $p(\theta^M=\theta^M_{10})=0.1$, $p(\theta^M=\theta^M_{00})=0.25$, $p(\theta^M=\theta^M_{11})=0.25$, and $p(\theta^M=\theta^M_{01})=0.4$. We show the results for the inferences given different findings in tables \ref{tab:HK8cases1quant} and \ref{tab:HK8cases2quant}. The mapping into expected posterior variance associated with each strategy is shown by the numbers in parentheses in  Table \ref{CaseLearn}.

```{r casestable, echo = FALSE, eval = FALSE}
I0D12 <- some_results(i=0, d=1, example = example_a2)
I0D12 <- round(I0D12, 3)
kable(cases_table(I0D12,  cases = cases1), caption = "\\label{tab:HK8cases1quant} Four cases with low inequality and  democratization. Question of interest: Was low inequality a cause of democracy? Table shows posterior beliefs for different data for 4 cases given information on $M$ or $P$. Data from Haggard and Kaufman (2012). Analyses here use priors assuming quantitative restrictions.")
I1D12 <- some_results(i=1, d=1, example = example_b2)
I1D12 <- round(I1D12, 3)
kable(cases_table(I1D12, cases = cases2), caption = "\\label{tab:HK8cases2} Four cases with high inequality and  democratization. Question of interest: Was high inequality a cause of democratization? Table shows posterior beliefs for different data for 4 cases given information on $M$ or $P$. Data from Haggard and Kaufman (2012). Analyses here use priors assuming quantitative restrictions.")

```


The results differ in various modest ways. However, the biggest difference we observe is in the degree to which the mobilization clue matters when we are looking for negative effects of inequality. As discussed, if we assumed monotonic positive effects of inequality on mobilization and monotonic positive effects of mobilization on inequality, then the mediator clue is uninformative about the indirect pathway since that pathway can only generate a positive effect. However, if we allow for the possibility of a negative effect of inequality on mobilization, we now make $M$ informative as a mediator even when the effect of inequality that we are interested in is negative: it is now possible that inequality has a negative effect on democratization via a negative effect on mobilization, followed by a positive effect of mobilization on democratization. So now, observing whether mobilization occurred adds information about whether a negative effect could have occurred via the mobilization pathway. 

Moreover, it is possible for the two effects of observing $M$ on our beliefs to work in opposite ways. What we learn from observing $M$ about the $I \rightarrow M \rightarrow D$ pathway may push in a different direction from what we learn from observing $M$ about the direct $I \rightarrow D$ pathway. We see this dynamic at work in a case with low inequality and democratization. Where we are only learning about $M$ as a moderator of $I$'s direct effect (monotonicity assumption in place), observing $M=0$ shifts our beliefs in favor of $I$'s negative effect. But where we are learning about $M$ as both mediator and moderator, observing $M=0$ shifts our beliefs *against* $I$'s negative effect. The reason for this latter result is straightforward: if $I=0$ and we then see $M=0$, then we have just learned that inequality's possible indirect negative effect, running via the mobilization pathway, has *not* in fact occurred; and this has a considerable downward effect on our beliefs in an overall negative effect of inequality. This learning outweighs the small positive impact of observing $M=0$ on our confidence that $I$ had a direct negative effect on $D$.

We see these differences most clearly in the cases of Albania (as compared to Mexico) and Nicaragua (as compared to Taiwan). Under priors fully constrained to monotonic causal effects, we saw that the mediator clue, $M$, made only a small difference to our inferences. However, if we allow for a negative effect of $I$ on $M$, even while believing it to be unlikely, observing mobilization in Albania and Nicaragua makes us substantially more confident that inequality mattered, and differentiates our conclusions about these cases more sharply from our conclusions about Mexico and Taiwan, respectively. 



