# Evaluating models {#evaluation}


:::: {.headerbox data-latex=""}
::: {.center data-latex=""}
:::
Model based inference takes the model seriously. But deep down we know that all of these models are wrong, in myriad ways. We examine strategies for figuring out whether a model is likely doing more harm than good.
::::
<br>

```{r packagesused15b, include = FALSE}
source("_packages_used.R")
do_diagnosis = FALSE
library(DeclareDesign)
```


Throughout this book we have maintained the conceit that you believe your model. But it is also obvious that even the most non-parametric-seeming models depend on substantive assumptions and that these are almost certainly wrong. The question then is not how much you believe your model (or whether you really believe what you say you believe) but whether your model is useful is some sense. How can we evaluate the usefulness of our models?


## Five Strategies

Sometimes a model is just a poor representation of underlying causal processes. Fortunately it can sometimes be possible to figure out your model is sending you in the wrong direction. We describe five strategies.

Imagine a situation in which researchers believe that the effect of $X$ on $Y$ runs entirely through $M$, positing a model of the form $X \rightarrow M \rightarrow Y$. Imagine, however, that the true causal process is one in which $X$ affects $Y$ *directly*;  $X$ has no effect on $M$, so that there is no indirect effect of $X$ on $Y$; and $M$ never has a negative effect on $Y$. 

The problem with the posited model, then, is that it represents overly strong beliefs about independence relations: it does not allow for a direct effect that is in fact operating.

<!-- AJ: "Restrictedness" seems a bit off here insofar as the "unrestricted" model is also restricted in essentially assuming no operative $X -> M$ arrow, as well as no negative direct effects.  -->

We are perfectly able to update using this too-strong $X \rightarrow M \rightarrow Y$ model and the data --- but the updated model can produce wildly inaccurate causal inferences. We show this using set of 200 observations simulated from a "true" model with direct effects only and an average effect of $X$ on $Y$ of $1/3$.^[Specifically, the data-generating model is $X \rightarrow M \rightarrow Y \leftarrow X$, with no effect of $X$ on $M$ and thus no indirect effect, and no negative effect of $X$ on $Y$, and an otherwise flat parameter space.] In the left panel of Figure \@ref(fig:modelsch15), we show the estimated average treatment effect of $X$ on $Y$ when using these data to update the $X \rightarrow M \rightarrow Y$ model.

In the righthand panel of the figure, we show the inferences using the same data but a model that makes weaker assumptions by allowing for direct effects: a $X \rightarrow M \rightarrow Y \leftarrow X$ model. With both models, we start with flat priors over nodal types.^[Note that the "true" model here involves restrictions on nodal types (no $X \rightarrow M$ effects and no negative $X \rightarrow Y$ effects) while the analysis models are both unrestricted.] 

We represent the true average effect with the vertical line in each graph.

As we can see, the weaker (i.e., more permissive) model performs OK: the true effect falls well within the posterior predictive distribution on the ATE. However, the stronger model, which excludes direct effects, generates a posterior predictive distribution that essentially excludes the right answer. So, if we go into the analysis with the stronger model, we have a problem. 

But will we notice?

In the remainder of this section, we explore a range of diagnostics that researchers can undertake to evaluate the usefulness of their models or to compare models with one another: checking assumptions of conditional independence built into a model; checking the model's fit; using "leave-one-out" cross-validation; and assessing model sensitivity.

```{r modelsch15, echo = FALSE}
n = 200

if(do_diagnosis){
  
 unrestricted_model <- 
    make_model("X -> M -> Y <- X") %>%
    set_parameters(
      statement = c("M[X=1] != M[X=0]", "(Y[X=1] < Y[X=0])"),
      parameters = c(0, 0)
      ) 
 
 restricted_model <- make_model("X -> M -> Y")%>%
    set_parameters(
      statement = c("M[X=1] < M[X=0]", "(Y[M=1] <= Y[M=0])"),
      parameters = c(0, 0)
      ) 
 
 data <- simulate_data(unrestricted_model, using = "parameters", n = n) 
     write_rds(data, "saved/ch15_data.rds") 
  
 XMYdata <- simulate_data(restricted_model, using = "parameters", n = n) 
     write_rds(data, "saved/ch15_XMYdata.rds") 
  
 restricted_model %>%
   update_model(data, keep_fit = TRUE) %>%
   write_rds("saved/ch15_restricted_model.rds") 

 unrestricted_model %>%
   update_model(data, keep_fit = TRUE) %>%
   write_rds("saved/ch15_unrestricted_model.rds") 

}

data               <-   read_rds("saved/ch15_data.rds") 
XMYdata            <-   read_rds("saved/ch15_XMYdata.rds") 
restricted_model   <-   read_rds("saved/ch15_restricted_model.rds") 
unrestricted_model <-   read_rds("saved/ch15_unrestricted_model.rds") 

if(do_diagnosis){
  
  ateXY1 <- query_distribution(restricted_model, "Y[X=1] - Y[X=0]", using = "posteriors")
  ateXY2 <- query_distribution(unrestricted_model, "Y[X=1] - Y[X=0]", using = "posteriors")

 write_rds(data.frame(restricted_ate = ateXY1, unrestricted_ate = ateXY2),
           "saved/ch15_ates.rds")
}

ates <- read_rds("saved/ch15_ates.rds") 

```

```{r 15badmodels, echo = FALSE, fig.cap = "A restricted model yields a credibility interval that does not contain the actual average effect."}
par(mfrow = 1:2)
hist(ates$restricted_ate, xlim = c(-.1, .6), main = "Stronger model", xlab = "Effect of X on Y") 
abline(v = .33, col = "red")
hist(ates$unrestricted_ate, xlim = c(-.1, .6), main = "Weaker model", xlab = "Effect of X on Y")
abline(v = .33, col = "red")

```


<!-- * $X$ correlates with $Y$ -->
<!-- * $X$ does not correlate with $M$  -->
<!-- * $M$ does not correlate with $Y$.  -->

<!-- These data are inconsistent with the model: under this model, if $X$ doesn't cause $M$ and $M$ doesn't cause $Y$, then there is no other way for $X$ to cause $Y$.  -->

### Check conditional independence

First, even before engaging in updating, we can look to see whether the data pattern is consistent with our causal model. In particular, we can check whether there are inconsistencies with the Markov condition that we introduced in Chapter 2:  that every node is *conditionally independent* of its non-descendants, given its parents. In this case, if the stronger model is right, then given $M$, $Y$ should be independent of $X$. 

Is it? 

One way to check is to assess the covariance of $X$ and $Y$ given $M$ in the data. Specifically, we regress $Y$ on $X$ for each value of $M$, once for $M=1$ and again for $M=0$; a correlation between $X$ and $Y$ at either value of $M$ would be problematic for the conditional independence assumption embedded in the stronger model. 

Note that this form of diagnostic test is a classical one in the frequentist sense: we start by hypothesizing that our model is correct and then ask whether the data were unlikely given the model.


```{r, echo = FALSE}
rbind(
  estimatr::lm_robust(Y~X, data = filter(data, M==0)) %>% tidy %>% filter(term == "X"), 
  estimatr::lm_robust(Y~X, data = filter(data, M==1)) %>% tidy %>% filter(term == "X")) %>%
  mutate(M = 0:1) %>%
  select(M, estimate, std.error, p.value) %>% 
  kable(digits = 3, caption =  "Regression coefficient on $X$ given $M=0$ and $M=1$")


```

We report the regression coefficients on $X$ in the table below. It is immediately apparent that we have a problem. At both values of $M$, there is a strong correlation between $X$ and $Y$, evidence of a violation of the Markov condition implied by the stronger model.^[In applying the Markov condition, we also need to take into account any unobserved confounding. For instance, suppose that there was an unobserved confounder of the relationship between $M$ and $Y$ in the $X \rightarrow M \rightarrow Y$ model. Then we would *not* expect $Y$ to be independent of $X$ conditional on $M$. Intuitively, the data pattern we described could be consistent with such a model in which on average $X$ does not cause $M$ but still *in those cases in which $X=M$* we have $Y=X$. Another way to think about this is that $M$ now acts as a collider between $X$ and another unobserved cause of $Y$; so conditioning on $M$ introduces a correlation between $X$ and this unobserved cause, and thus between $X$ and $Y$.] 

Identifying the full set of conditional independence assumptions in  a causal model can be difficult. There are however well developed algorithms for identifying what sets, if any, you need to conditional on to ensure conditional Independence between two nodes given a DAG. `R` users can quickly access such results  using the `impliedConditionalIndependencies` function in dagitty package.

<!-- AJ: I do not follow the sentence starting with "Intuitively" in the above. -->




<!-- ROUGH TEXT: -->

<!-- approach 2 -- say actual confound is q~=0; but model  assumes q = 0. Draw data from priors, draw data; given data type (001, 100 etc) plot (a) the posterior distribution under no confounding nad (b) the distribution of estimands that gave rise to the data.   -->


<!-- (Verma and Pearl, 1990) identify conditions under which we can check some independence assumptions. -->


<!--  Pearl (1995) gives conditions for assessing for discrete data whether $Z$ has a  direct effect on $Y$. (Involves inequalities) -->

<!-- Evans (Graphical methods for inequality constraints in marginalized DAGs) generalizes the instrumental inequality. -->
 


<!-- ### Computational clues -->

<!-- Second, we may be lucky and run into computation issues. In this example there is a good chance that when you fit the stronger  model, `stan` will throw an error:  -->

<!-- > `Bulk Effective Samples Size (ESS) is too low, indicating posterior means and medians may be unreliable.` -->


<!-- AJ: Needs elaboration: why is this error a function of problems with the model? -->
<!-- MH: I agree; do feel free to look into it; I will try -->
<!-- AJ: I have done some digging and not found anything helpful on this -->
<!-- MH: Well that's just great -->
<!-- AJ: Looked a little more - still nothing! Whence did you get a sense there might be something to this? -->


### Bayesian $p-$value: Are the data unexpected given your model?

A third approach asks whether features of the data you observe are in some sense unusual given your model, or more unusual given your model than another model. For instance, if one model assumed no adverse effects of $X$ on $Y$ and no confounding, then a strong negative correlation between $X$ and $Y$ would be unusual, even for the model updated with this data; and this negative correlation would *more* unusual for this model than for a model that allowed for adverse effects.

In fact, this approach is quite classical: we are looking to see whether we should "reject" our model in light of inconistent data.  

<!-- AJ: But it can also be done in a comparative sense as we do below, right? Which is then not about "rejection", right? I've actually adjusted the text two paragraphs up to make the idea of comparison more prominent. This works?-->

An approach for doing this using simulated data from the posterior predictive distribution is described in @gabry2019visualization.^[Tools in the `bayesplot` package can be used to show how typical the data we observe is for different models] 

We consider two test statistics, comparing our stronger to our weaker model. First, we look just at the distribution of the outcome $Y$ to see how the actual distribution in the data compares to the predicteddistribution from the updated model. Second, we look at the actual correlation between $X$ and $Y$ and see how this compares to the predicted distribution. In both cases we calculate a two sided $p$-value by assessing and doubling the share of the mass of the predictive posterior distribution that lies on the more extreme side of the observed data. If the observed data were at the mean of the predicive distribution, then we would have a $p$-value of 1. If it were at the 95th percentile we would have a $p$-value of 0.10. (We note that in this straightforward calculation we assess the probability of the data given the same model that generated the data; approaches could also be used that seek out-of-sample estimates of the  probability of observing the observed data.)

<!-- AJ: Last sentence confuses me. Aren't we using data only from unrestricted_model? -->

For the first test, we see that the predicted distribution of the outcome $Y$ is similar for both updated models; and the actual mean outcome is within the distribution of predicted mean outcomes. The $p$-values for the stronger (0.11) and weaker models (0.14) suggest that the observed mean $Y$ values are about equally likely for both models. 

When it comes to the correlation between $X$ and $Y$, however, the two models perform very differently. The posterior predictive distribution from the stronger model is centered around a $0$ correlation and does not even extend out as far as the observed correlation. The resulting $p$-value is 0, meaning that from the perspective of the stronger model the $X,Y$ correlation in the data is entirely unexpected. A frequentist looking at the observed correlation between $X$ and $Y$ should feel comfortable rejecting the stronger model. The updated weaker model, in contrast, predicts a strong correlation, and the observed correlation is comfortably within the posterior predictive distribution, with a $p$-value of 0.08. 


<!-- AJ: Need to code in the p-values -->

At first blush, the abysmal performance of the stronger model may seem paradoxical. Even after this model has *seen* the $X,Y$ correlations in the data, the model still finds those correlations highly surprising. What keeps the  $X \rightarrow M \rightarrow Y$ model  from learning, however, is the strength of the assumptions it contains. The problem is that $M$ is uncorrelated with $X$ in the true data-generating process, so the stronger model learns that there is no indirect effect. But, at the same time, this model does not *allow* for a direct effect. Despite what would seem to be overwhelming evidence of a systematic $X,Y$ correlation, a causal relationship connecting $X$ to $Y$ remains extremely unlikely given the $X,M$ data pattern and the impossibility of direct effects. The stronger model just can't handle the truth. The weaker model, on the other hand, readily learns about the direct $X \rightarrow Y$ effect.


```{r ch15simulations, echo = FALSE}

if(do_diagnosis){
  
 sims <- 500
 n <- nrow(data)
 
 # Outcome prediction
 ############################################# 
 t(replicate(sims, {
   dd <- make_data(restricted_model, n = n, param_type = "posterior_draw")
   c(Y=mean(dd$Y), C = cor(dd$Y, dd$X))})) %>%  data.frame %>%
  write_rds("saved/ch15_replicates_restricted.rds") 

 t(replicate(sims, {
   dd <- make_data(unrestricted_model, n = n, param_type = "posterior_draw")
   c(Y=mean(dd$Y), C = cor(dd$Y, dd$X))})) %>% data.frame %>%
  write_rds("saved/ch15_replicates_unrestricted.rds") 
  
  }

replicates_restricted   <-   read_rds("saved/ch15_replicates_restricted.rds") 
replicates_unrestricted <-   read_rds("saved/ch15_replicates_unrestricted.rds") 

par(mfrow = c(2,2))
xlimY <- range(c(replicates_restricted$Y, replicates_unrestricted$Y))
xlimC <- range(c(replicates_restricted$C, replicates_unrestricted$C))

## FLAG: ADD p values to others graphs

p_two_sided <- function(Y, y, digits = 2){
  if(y <  mean(Y, na.rm = TRUE)) return(round(2* mean(y > Y, na.rm = TRUE), digits))
  if(y >= mean(Y, na.rm = TRUE)) return(round(2* mean(y <= Y, na.rm = TRUE), digits))
}
p_temp <- p_two_sided(replicates_restricted$Y, mean(data$Y))
hist(replicates_restricted$Y, 
     main = paste("Y mean | stronger: p = ", p_temp),
                  xlim = xlimY, xlab = "predictions") 
     abline(v = mean(data$Y), col = "red")

p_temp <- p_two_sided(replicates_restricted$C, cor(data$Y, data$X))
hist(replicates_restricted$C, 
     main = paste("X, Y covariance | stronger: p =", p_temp), 
     xlim = xlimC, xlab = "predictions"); abline(v = cor(data$Y, data$X), col = "red")

p_temp <- p_two_sided(replicates_unrestricted$Y, mean(data$Y))
hist(replicates_unrestricted$Y, 
     main = paste("Y mean | weaker: p = ", p_temp),
     xlim = xlimY, xlab = "predictions"); abline(v = mean(data$Y), col = "red")

p_temp <- p_two_sided(replicates_unrestricted$C, cor(data$Y, data$X))
hist(replicates_unrestricted$C, 
     main = paste("X, Y covariance | weaker: p = ", p_temp), 
     xlim = xlimC, xlab = "predictions"); abline(v = cor(data$Y, data$X), col = "red")


# FLAG: WOULD LIKE THIS DONE WITH GGPLOT
```


<!-- ## Model likelihoods -->

<!-- Bayesian information criterion (BIC)^[$BIC = \ln(n)k - 2\ln(\hat{L})$ where $\hat{L}$ is the maximized likelhood, $k$ is the number of parameters, and $n$ the number of data points.] -->
<!-- BIC involves a penalty for more parameters -->


<!-- ```{r, echo = FALSE} -->

<!-- if(do_diagnosis) { -->
<!-- observed_data <- collapse_data(data, restricted_model) -->

<!-- likely <- function(model, s, posterior = TRUE) { -->
<!--   if(posterior)  pars <- model$posterior_distribution[1:s,] -->
<!--   if(!posterior) pars <- get_prior_distribution(model, s)[1:s,] -->
<!--   L <- apply(pars, 1,  function(par) make_data_probabilities(model, pars = par, observed_data, normalize = FALSE)) -->
<!--   mean(L) -->
<!--   } -->

<!-- s <- 500 -->
<!-- L_prior         <- likely(restricted_model, s, posterior = FALSE) -->
<!-- L_restricted    <- likely(restricted_model, s) -->
<!-- L_unrestricted  <- likely(unrestricted_model, s) -->

<!-- df <- data.frame(L = (c(L_unrestricted/L_prior, L_restricted/L_prior, L_unrestricted/L_restricted))) -->

<!-- rownames(df) <- c("Unrestricted / Prior", "Restricted / Prior", "Unrestricted / Restricted") -->

<!-- write_rds(df, "saved/ch15_likelihoodsdf.rds")  -->
<!-- } -->

<!-- df <- read_rds("saved/ch15_likelihoodsdf.rds")  -->

<!-- kable(df, col.names = "Bayes factors", digits = 2, caption = "Posterior odds: the relative likelihood of one model over another") -->
<!-- ``` -->

<!-- Compare likelihoods of the data under different models -->
<!-- Check look package for rstan -->


### Leave-one-out (LOO) cross-validation

A further class of model-validation methods involves cross-validation. Rather than asking how well the updated model predicts the data used to update it, cross-validation uses the data at hand to estimate how well the model is likely to predict new data that have not yet been seen. One way to do this is to split the available data, using one subsample to update and then assessing predictions using the other subsample. We focus here, however, on approaches that use *all* of the available data to estimate out-of-sample predictive performance.

One such approach is the "leave-one-out" (LOO) algorithm. In a LOO approach, we update the model using all data points except for one and then ask how well the model performs in predicting the left-out observation. We repeat this for every data point in the dataset to assess how well we can predict the entire dataset. 

Often, the LOO approach is used to predict a particular outcome variable. In a \CausalQuery\ framework, however, we are interested in predictions over the joint realization of all nodes, not just a single "outcome." Thus, we calculate the posterior probability of each data point, using the model updated with all of the other observations. 

<!-- AJ: Is the "predict all nodes" using a sledghammer here?Should we be setting this up in a more query-dependent way, I wonder? Could the same model be good for some kind of queries but not for others? Not quite sure how to think about that -- about *what* we'd be predicting conditional on what the query is, and maybe this isn't a coherent notion. -->

<!-- AJ: Change figure labels to stronger/weaker models. Also, need to make x-axes "data count" (without referring to a model) -->



```{r, echo = FALSE}
#Leave one out


if(do_diagnosis){
  
  get_loo_likelihood(data, restricted_model, 1000) %>% write_rds("saved/ch15_loo_restricted.rds")
    
  get_loo_likelihood(data, unrestricted_model, 1000) %>% write_rds("saved/ch15_loo_unrestricted.rds")

}

loo_restricted   <- read_rds("saved/ch15_loo_restricted.rds") 
loo_unrestricted <- read_rds("saved/ch15_loo_unrestricted.rds")  

plot(loo_restricted$case_likelihood, loo_unrestricted$case_likelihood,
     xlab = "Likelihood | restricted", ylab = "Likelihood | unrestricted")
text(loo_restricted$case_likelihood, loo_unrestricted$case_likelihood, loo_restricted$data$event)

plot_loo <- function(loo, name = "restricted model", cex = .5)
  with(loo, {
  plot(case_likelihood[data$count>0], data$count[data$count>0], 
       xlab =  paste("Case likelihood  |", name), 
       ylab =  "Actual number of cases", type = "n")
  text(case_likelihood, data$count, data$event, cex = cex)
  abline(a=0, b = sum(data$count[data$count>0]), col = "lightgrey")})

par(mfrow = c(1,2))
plot_loo(loo_restricted, name = "restricted model")
plot_loo(loo_unrestricted, name = "unrestricted model")

```

The LOO estimate of out-of-sample predictive fit, for a dataset with $n$ observations, is then:

$$\prod_1^np(y_i|y_{-i}, \text{model})$$
where $y_{-i}$ is the data pattern with observation $y_i$ left out, and $y_i$ represents the values of all nodes of interest for observation $i$. 

<!-- AJ: Any reason not to use a log in the above equation? Would this make some of the outputs easier to read? -->
<!-- MH it would get a more humnan scale but think raw numbers are more intuitive? -->

We implement LOO cross-validation of the stronger and weaker models using 200 observations generated from the same data-generating model employed above. We find that the LOO likelihood of the data under the stronger model is `r  formatC(loo_restricted$loo_likelihood, format = "e", digits = 2)` while
the likelihood is  `r  formatC(loo_unrestricted$loo_likelihood, format = "e", digits = 2)` under the weaker model. Thus, the weaker model represents an estimated improvement in out-of-sample prediction on the order of   `r  formatC(loo_unrestricted$loo_likelihood/loo_restricted$loo_likelihood, format = "e", digits = 2)`. 

We can visualize the pattern in Figure XXXX, where we plot the likelihood of each possible data type under the stronger model against the likelihood of that data type under the weaker model. Looking at the scales of the two axes---which is much more compressed on the horizontal than on the vertical---one can see that the stronger model is not able to differentiate as much across the data types as the weaker. 

<!-- AJ: What are these likelihoods exactly? They are informed by updating, I assume? But updating using what data exactly -- what's left out? Generally unsure if my text above is right.-->

Notably, the stronger model is not able to "learn" from the data about the (*in fact*, operative) relationship between $X$ and $Y$. We can see that, for any given $X, M$ combination, the two possible values of $Y$ are predicted with essentially the same likelihood. The stronger model also seems to have "learned" from chance correlations in the data that different values $X,M$ combinations are differentially likely---even though they are uncorrelated under the true model. The weaker model, on the other hand, basically divides the data types into two groups: those with a positive $X,Y$ correlation and those with a negative $X,Y$ correlation and has correctly (given the true model) learned that the former is more likely than the latter.

In Figure XXXX, we then see how the likelihoods of each data type line up with the actual count of each data type. As we can see, the weaker model updates to likelihoods that fit the actual data pattern well while the stronger model does not.


We can also turn the tables and imagine that the *stronger* model represents the true data-generating process. We  implement LOO cross-validation of the two models using 200 data points generated from the stronger model. In Figure XXXX, we see a comparison of the likelihoods of the data types under the two updated models and note that they are extremely similar. This represents an important asymmetry: the model that makes weaker assumptions performs far better in handling data generated by a "stronger" true model than does the stronger model in learning about a process that violates one of its assumptions. Since the weaker model allows for both direct and indirect effects, the weaker *can* learn about the parameters of the true process in the first situation; but the strong model cannot do so in the second situation because it has by assumption ruled out a key feature of that process (the direct effect). 







```{r, echo = FALSE}

if(do_diagnosis){
  
  get_loo_likelihood(XMYdata, restricted_model, 1000) %>% write_rds("saved/ch15_loo_restricted_2.rds")
    
  get_loo_likelihood(XMYdata, unrestricted_model, 1000) %>% write_rds("saved/ch15_loo_unrestricted_2.rds")

}

loo_restricted_2   <- read_rds("saved/ch15_loo_restricted_2.rds") 
loo_unrestricted_2 <- read_rds("saved/ch15_loo_unrestricted_2.rds")  


par(mfrow = c(1,2))
plot_loo(loo_restricted_2, name = "restricted model")
plot_loo(loo_unrestricted_2, name = "unrestricted model")
    
```

<!-- FLAG: BETTER: PLOT CASES ON BOTTOM AND TWO LIKELIHOODS AS POINTS; HAVE TYPES AS LABELS X AI IF POSSIBLE -->

While it is difficult to see this in Figure XXXX, the stronger model still performs better here than the weaker model. The likelihood of the data under the stronger model is now `r  formatC(loo_restricted_2$loo_likelihood, format = "e", digits = 2)`, compared to 
the likelihood of  `r  formatC(loo_unrestricted_2$loo_likelihood, format = "e", digits = 2)` under the weaker model. Thus, the weaker model represents an estimated loss to out-of-sample prediction on the order of   `r  formatC(loo_unrestricted_2$loo_likelihood/loo_restricted_2$loo_likelihood, format = "e", digits = 2)`. This is not surprising insofar as the stronger model *precisely* models the data-generating process while the extra parameters in the weaker model allow for "learning" from chance features of the data.

These examples display features of estimation of out-of-sample prediction accuracy familiar from a regression context. In a regression framework, adding parameters to a model may improve fit to sample---generating gains to out-of-sample prediction accuracy when the new parameters pick up systematic features of the data-generating process---but run a risk of over-fitting to chance pattenrs in the data. Similarly, in a structural-causal-model framework, for a model with weaker assumptions and more parameters. We saw that the weaker model performed much better when the true process involved direct effects since the extra parameters, allowing for direct effects, captured something "real" going on. But the same model performed slightly worse than the stronger model when there were no direct effects to pick up, such that the extra parameter could only model noise.

<!-- AJ: Last graph here doesn't seem right, with the data-type labels all stacked vertically. And no graph for unrestricted model. -->

<!-- Note on parsimony. It is interesting to note that more complex models have to pay their way: if they do not predict better than simpler models, then simpler models  are favored. For the intuition imagine in fact $X$ and $Y$ are unrelated. Say we had two data points, one with... -->

### Sensitivity

The last approach we consider brackets the question of which model is better and asks, instead: how much do your conclusions depend on the model? You can worry less about your assumptions if the conclusions are not strongly dependent on them.

To illustrate using a process tracing example, consider a situation in which we are unsure about posited parameter values: that is, about the probability of particular effects at particular nodes. It is likely to be the case in many research situations that we are considerably uncertain about how to quantify intuitive or theoretically informed beliefs about the relative likelihood of different effects. 

Suppose, for instance, that we begin with an $X \rightarrow M \rightarrow Y$ model. And suppose, further, that we believe that it is unlikely that $M$ has an adverse effect on $Y$. But we are not sure *how* unlikely that adverse effect is.  (We assume all other modal types are equally likely.) Finally, say that we want to use the observation of $M$ to draw an inference about whether $X=1$ caused $Y=1$ in an $X=Y=1$ case. 

How much does our inference on $X$'s effect on $Y$---when we see $M=0$ or $M=1$---depend on this second stage assumption about the probability of a negative $M \rightarrow Y$ effect?

We answer the question by looking at posterior beliefs for a range of possible values for the relevant parameter, $\lambda^Y_{10}$. In Table  REF, we examine a range of values for $\lambda^Y_{10}$, from 0 to 0.25 (full parity with other types). For each parameter value, we first show the resulting prior belief about the probability that $X=1$ caused $Y=1$. We can see that, before we observe $M$, we think that a positive $X \rightarrow Y$ effect is more likely as a negative $M \rightarrow Y$ effect is more likely. This stands to reason since a negative second-stage effect is one possible process in which a positive $X \rightarrow Y$ effect might occur. And higher values for $\lambda^Y_{10}$ come disproportionately at the expense of types under which $X$ cannot affect $Y$.^[Increasing weight on $\lambda^Y_{10}$ is drawn equally from $\lambda^Y_{00}$, $\lambda^Y_{11}$, and $\lambda^Y_{10}$, with the first two of these three representing null effects.]

In the next two columns, we show the posterior belief we arrive at when we observe $M=0$ and then $M=1$, for each $\lambda^Y_{10}$ assumption. Looking at the last column first, we see that our inference from $M=1$ does *not* depend at all on our beliefs about adverse $M \rightarrow Y$ effects. The reason is that, if we see $M=1$, we already know that $M$ did not have a negative effect on $Y$, given that we also know $Y=1$. Our beliefs are purely a function of the probability that there are positive effects at both stages as compared to the probability of other causal types that could yield $X=M=Y=1$, a comparison unaffected by the probability of a negative $M \rightarrow Y$ effect.  

Our inferences when $M=0$, on the other hand, do depend on $\lambda^Y_{10}$: when we see $M=0$, our belief about a positive $X \rightarrow Y$ effect depends on the likelihood of *negative* effects at both stages. We see, then, that the likelier we think negative effects are at the second stage, the higher our posterior confidence in a positive $X \rightarrow Y$ effect when we see $M=0$. 


```{r, fig.cap = "Possible inferences on `X caused Y` given observation of M=0 and M=1 given different possible models. The highest inference when M=0 is lower than the lowest inference when M=1", echo = FALSE}

model <- make_model("X -> M -> Y") %>%
    set_parameters(
      param_names = c("M.10"), 
      parameters = .25)


adverse2 <- seq(0, .25, .05)

out <-  
  sapply(adverse2, function(a) {
  
  query_model(
  set_parameters(model, 
                 param_names = c("Y.10"), 
                 parameters = c(a)),
  query = c("Y[X=1] > Y[X=0]"),
  given = c("X==1 & Y==1", "X==1 & Y==1 & M==0", "X==1 & Y==1 & M==1"),
  using = "parameters",
  expand_grid = TRUE)$mean})
out2 <- t(out)  %>% data.frame
names(out2) <- c("No data", "M0", "M1")
out2 <- arrange(out2,M0, M1)

kable(cbind(adverse2, out2), col.names = c("$\\lambda^Y_{10}$", "Prior", "$M=0$", "$M=1$"), caption = "Inferences on the probability that $X$ caused $Y$ upon seeing $M=0$ or $M=1$ for a range of possible values of $\\lambda^Y_{10}$")

```


Even though our inferences given $M=1$ do not depend on $\lambda^Y_{10}$, the amount that we *update* if we see $M=1$ *does* depend on $\lambda^Y_{10}$. This is because $\lambda^Y_{10}$ affects our belief, prior to seeing $M$, that $X=1$ caused $Y=1$. Working with a low $\lambda^Y_{10}$ value, we start out less confident that $X=1$ caused $Y=1$, and thus our beliefs make a bigger jump if we do see $M=1$ than if we had worked with a $\lambda^Y_{10}$ higher value. 

However, to the extent that we want to know how our assumptions affect our *conclusions*, the interesting feature of this illustration is that sensitivity depends on what we find. The answer to our query is sensitive to the $\lambda^Y_{10}$ assumption if we find $M=0$, but not if we find $M=1$.  It is also worth noting that, even if we observe $M=0$, the sensitivity is limited across the range of parameter values tested. In particular, for all $\lambda^Y_{10}$ values below parity (0.25), seeing $M=0$ moves our beliefs *in the same direction.* 

We can use the same basic approach to examine how our conclusions change if we relax assumptions about nodal-type restrictions, about confounds, or about causal structure. 

We also note that in cases in which you cannot quantify uncertainty about parameters you might still be able to engage in a form of "qualitative inference." There is a literature on probabilistic causal models that assesses the scope for inferences when researchers provide ranges of plausible values for parameters (perhaps intervals, perhaps only signs, positive negative, zero), rather than specifying a probability distribution. For a comprehensive treatment of qualitative algebras, see @parsons2001qualitative. Under this kind of approach, a researcher might willing to say that they think some probability $p$ is not plausibly  greater than .5,  but unwilling to make a statement about their beliefs about where in the $0$ to $0.5$ range it lies. Such incomplete statements can be enough to rule our classes of conclusion.
 


## Evaluating the Democracy-Inequality model

<!-- ** FLAG: GENERATE A TABLE SHOWING HOW PIMD DOES ON 5 CRITERIA **  -->


```{r pimdmodelsch15, echo = FALSE}

pimd <- make_model("I -> M -> D <- P; I -> D") %>%
  
         set_restrictions(c( 
           "(M[I=1] < M[I=0])",
           "(D[I=1] > D[I=0]) | (D[M=1] < D[M=0]) | (D[P=1] < D[P=0])"))

# make an alternative model that permits negative I->M effects

pimd_negmed <- make_model("I -> M -> D <- P; I -> D") %>%
  
         set_restrictions(c( 
           "(D[I=1] > D[I=0]) | (D[M=1] < D[M=0]) | (D[P=1] < D[P=0])"))

pimd_nonmon <- make_model("I -> M -> D <- P; I -> D")

data <- CausalQueries::democracy_data

data_2 <- read.csv("data/pimd2.csv") %>% #new data we compiled from other sources
  mutate(I2 = I2*1) %>% select(country, P2, I2, M2, D2) %>%
  rename(P = P2, I = I2, M= M2, D = D2)

#data <- read.csv("data/pimd2.csv") %>% 
 # mutate(I2 = I2*1) %>% select(country, P2, I2, M2, D2) %>%
  #rename(P = P2, I = I2, M= M2, D = D2) 


if(do_diagnosis){
  
pimd %>%
   update_model(data, keep_fit = TRUE, iter = 8000) %>%
   write_rds("saved/ch15_pimd_model.rds") 

pimd_negmed %>%
   update_model(data, keep_fit = TRUE, iter = 8000) %>%
   write_rds("saved/ch15_pimd_negmed_model.rds") 

pimd_nonmon %>%
   update_model(data, keep_fit = TRUE, iter = 4000) %>%
   write_rds("saved/ch15_pimd_nonmon_model.rds") 
#unrestricted model saved here

}

if(do_diagnosis)
  pimd_nonmon %>% set_priors(distribution = "jeffreys") %>% 
   update_model(data_2, chains = 4, iter = 6000, keep_transformed = TRUE) %>%
   write_rds("saved/ch15_pimd_nonmon_model_2.rds") 
#unrestricted model saved here


pimd <-   read_rds("saved/ch15_pimd_model.rds") 
pimd_negmed <-   read_rds("saved/ch15_pimd_negmed_model.rds")
pimd_nonmon <-   read_rds("saved/ch15_pimd_nonmon_model.rds")

pimd_nonmon_2 <-   read_rds("saved/ch15_pimd_nonmon_model_2.rds")

```

### Check assumptions of conditional independence

Our model presupposes that $P$ and $I$ are independent and that $P$ and $M$ are independent. Note that the model is consistent with the possibility that, conditional on $D$, there is a correlation betwen $M$ and $P$ or between $I$ and $P$, as $D$ acts as a collider for these pairs of nodes. 

To test these assumptions, we in fact need to depart from the dataset drawn from @haggard2012distributive because these authors only examined cases in which $D=1$, those that democratized. Thus, we cannot use these data to assess the relationships not conditional on $D$ or conditional on $D=0$. We generate observations on all four nodes for a broader set of cases by pulling together measures from the following sources, with the aim of modeling democratization that occurred between 1990 and 2000^[The data that we use to measure mobilization, from @clark2016mobilization, covers only the 1990s.]:

* **Inequality**: We measure inequality, $I$, using the Gini estimates from the University of Texas Inequality Project (@galbraith2019inequality). As we want to measure inequality at the beginning of the period, we take the Gini measure for each country that is closest in time to the year 1989. We then dichotomize the variable using the median value for the period as a cutoff. 

* **Mobilization**: We measure $M$ using the Mass Mobilization Protest Data from @clark2016mobilization. To capture the kinds of mobilization on which redistributive theories of democratization focus, we restrict our focus to protests in the demand categories "land farm issue," "labor wage dispute," "price increases, tax policy," and "political behavior, process." We also include only those gatherings with a size of at least 1000 protesters. We code a country case as $M=1$ if and only if, during the 1990s, it experienced at least one protest that meets both the demand-type and size criteria.

* **Pressure**: We draw on the GIGA Sanctions Dataset to measure international pressure, $P$. Specifically, we code a country case as $P=1$ if and only if the country was the target of democratization-focused sanctions during the 1990-2000 period. 

* **Democratization**: We use dichotomous democracy measures from @cheibub2010revisited, in two ways. First, we filter countries such that our sample includes only those that were not democracies in 1990 ($N=77$). We then use the democracy measure for the year 2000 to determine which countries democratized, coding as $D=1$ those and only those cases that Cheibub and Vreeland code as democracies in that year.


We check the model's assumptions in relation to conditional independencies through a set of simple regression models, displayed in Table \@ref(tab:ch15cipimd). In the first two rows, we examine the simple correlation between $P$ and $I$ and between $P$ and $M$, respectively. We can see from the estimates in the first row that the data pattern is consistent with our assumption of unconditional independence of $I$ and $P$. However, we also see that there *is* an unconditional correlation between $I$ and $M$, something that is excluded by our model.

<!-- MH: Realize we actually have a big iussue here: we only have M,P data in cases with D = 1 so we are in fact condtiioining on D and so we do *not* expect independence  -->

```{r ch15cipimd, echo=FALSE}
independencies <- function(model){

  dagitty::dagitty(paste("dag{", model$statement, "}" )) %>%
  dagitty::impliedConditionalIndependencies()
}
# independencies(pimd)


# dd <- CausalQueries::democracy_data 

rbind(
  lm_robust(P~I, data = data_2) %>% tidy %>% filter(term == "I"), 
  lm_robust(P~M, data = data_2) %>% tidy %>% filter(term == "M"), 
  lm_robust(P~I, data = filter(data_2, M==0)) %>% tidy %>% filter(term == "I"), 
  lm_robust(P~I, data = filter(data_2, M==1)) %>% tidy %>% filter(term == "I"))  %>% 
  #lm_robust(P~M, data = filter(data_2, I==0)) %>% tidy %>% filter(term == "M"), 
  #lm_robust(P~M, data = filter(data_2, I==1)) %>% tidy %>% filter(term == "M"))    
 mutate(Given = c("-", "-", "M = 0", "M = 1")) %>%
  select(Given, estimate, std.error, p.value) %>% 
  mutate(Correlation = c("P,I", "P,M", "P,I", "P,I")) %>%
  select(Correlation, everything()) %>%
  kable(digits = 3, caption =  "Regression coefficients to assess conditional independence")


```



We can dig a little deeper, however. The model *also* implies that $P$ should be independent of $I$ given $M$ --- since $D$ blocks all paths between $P$ and either $I$ or $M$. We test these assumptions in rows 3 and 4 of the table, wgere we examine the conditional independence of $P$ and $I$ given $M=0$ and given $M=1$. Here the evidence is also troubling for our model, as we see a relatively strong negative correlation between $P$ and $I$ when $M=0$, and positive correlation when $M=1$. 

While we cannot identify the correct model from this data pattern, one possible explanation could be that pressure has a direct effect on mobilization, making mobilization the product of inequality and pressure jointly.^[There is, in fact, also a strong positive interaction between $I$ and $P$ in a linear model of $M$.] A model with an arrow running from $P$ to $M$ would make the model consistent with the unconditional correlation between these two variables, the conditional correlation between $P$ and $I$ given $M$ (since $M$ would now be a collider for $I$ and $P$), as well as the unconditional independence of $I$ and $P$. A possible way forward --- which we do not pursue here --- would be to now amend the model and evaluate the revised model against an independent set of data. 


<!-- Substantively, in states with high inequality, mobilization is more common in cases without international pressure ($P=0$) (arising in 0 of 13 cases) than in cases that experience pressure arising in just 1 of 5 cases. Again, our model may be missing a direct link from pressure to mobilization, which would generate the possibility that inequality and pressure interactively affect mobilization. -->

<!-- Unclear what the case numbers above are referring to. For instance, I see 13 cases with I=0, P=0. But what's the 0?-->

<!-- I am thinking we should now bring in models that relax the assumptions that appear not to hold in the CI analysis. For instance, introduce an arrow from $P$ to $M$. -->


### Bayesian $p$-value

We turn next to evaluating the democratization model using the Bayesian $p-$value approach, and for this purpose can return to the data that we coded from @haggard2012distributive's qualitative vignettes. In the two panels of Figure \@ref(fig:ch15pvalue_pimd), we plot the posterior predictive distributions from our updated model for two quantities of interest: the outcome $D$ and the correlation between $I$ and $M$. In each graph, we indicate with a vertical line the mean value for these quantities for the data at hand and report the $p-$value: the probability of the observed data conditional on our model. As we can see both visually and from the $p-$values, the model performs well here as the data that we observe are not at all unexpected under the model.

```{r ch15pvalue_pimd, echo = FALSE}
if(do_diagnosis){
  
 sims <- 500
 n <- nrow(data)
 
 # Outcome prediction
 ############################################# 
 t(replicate(sims, {
   pimd_sim <- make_data(pimd, n = n, param_type = "posterior_draw")
   c(D=mean(pimd_sim$D), C = cor(pimd_sim$I, pimd_sim$M))})) %>%  data.frame %>%
  write_rds("saved/ch15_replicates_pimd.rds") 

  
  }

replicates_pimd <-   read_rds("saved/ch15_replicates_pimd.rds") 

par(mfrow = c(1,2))
xlimD <- range(c(replicates_pimd$D))
xlimC <- range(c(replicates_pimd$C))

p_two_sided <- function(D, d, digits = 2){
  if(d <  mean(D)) return(round(2* mean(d > D), digits))
  if(d >= mean(D)) return(round(2* mean(d <= D), digits))
}

p_temp <- p_two_sided(replicates_pimd$D, mean(data$D))
hist(replicates_pimd$D, main = paste("D mean | pimd: p= ", p_temp)
     , xlim = xlimD, xlab = "predictions"); abline(v = mean(data$D), col = "red")

p_two_sided <- function(Y, y, digits = 2){
  if(y <  mean(Y)) return(round(2* mean(y > Y), digits))
  if(y >= mean(Y)) return(round(2* mean(y <= Y), digits))
}

p_temp2 <- p_two_sided(replicates_pimd$C, cor(data$M, data$I, use = "complete.obs"))
hist(replicates_pimd$C, main = paste("I, M covariance | pimd: p= ", p_temp2),
     xlim = xlimC, xlab = "predictions"); abline(v = cor(data$M, data$I, use = "complete.obs"), col = "blue")

#Some problem with p_temp2 line above. cor(data$M, data$I) returns NA.  Because of missing values??

# FLAG: WOULD LIKE THIS DONE WITH GGPLOT
```


<!-- p_two_sided <- function(Y, y, digits = 2){ -->
<!--   if(y <  mean(Y)) return(round(2* mean(y > Y), digits)) -->
<!--   if(y >= mean(Y)) return(round(2* mean(y <= Y), digits)) -->
<!-- } -->

<!-- hist(replicates_restricted$Y,  -->
<!--      main = paste("Y mean | stronger: p = ", p_temp), -->
<!--                   xlim = xlimY, xlab = "predictions")  -->
<!--      abline(v = mean(data$Y), col = "red") -->

<!-- p_temp <- p_two_sided(replicates_restricted$C, cor(data$Y, data$X)) -->
<!-- hist(replicates_restricted$C,  -->
<!--      main = paste("X, Y covariance | stronger: p =", p_temp),  -->
<!--      xlim = xlimC, xlab = "predictions"); abline(v = cor(data$Y, data$X), col = "red") -->

<!-- p_temp <- p_two_sided(replicates_unrestricted$Y, mean(data$Y)) -->
<!-- hist(replicates_unrestricted$Y,  -->
<!--      main = paste("Y mean | weaker: p = ", p_temp), -->
<!--      xlim = xlimY, xlab = "predictions"); abline(v = mean(data$Y), col = "red") -->

<!-- p_temp <- p_two_sided(replicates_unrestricted$C, cor(data$Y, data$X)) -->
<!-- hist(replicates_unrestricted$C,  -->
<!--      main = paste("X, Y covariance | weaker: p = ", p_temp),  -->
<!--      xlim = xlimC, xlab = "predictions"); abline(v = cor(data$Y, data$X), col = "red") -->


<!-- <!-- AJ: Should we also calculate a p-value? --> 
<!-- AJ: And what about comparing multiple models? -->


### LOO validation

Turning to "leave one out" model assessment, we now consider comparing our base model to models that make weaker assumptions. In one alternative model, we drop the assumption of monotonicity of $M$ in $I$ and instead put a low prior probability on negative $I \rightarrow M$ effects. We also test a maximally flexible model (given the DAG) in which we make no montonocity assumptions for any of the causal effects.

```{r loo_pimdplots, echo = FALSE}
# Leave one out



if(do_diagnosis){
  
  get_loo_likelihood(CausalQueries::democracy_data, pimd, sims = 25, iter = 4000) %>% write_rds("saved/ch15_loo_pimd.rds")
    
  get_loo_likelihood(CausalQueries::democracy_data, pimd_negmed, sims = 25, iter = 4000) %>% write_rds("saved/ch15_loo_pimdnegmed.rds")
  
  get_loo_likelihood(CausalQueries::democracy_data, pimd_nonmon, sims = 25, iter = 4000) %>% write_rds("saved/ch15_loo_pimdnonmon.rds")

}

loo_pimd       <- read_rds("saved/ch15_loo_pimd.rds") 
loo_pimdnegmed <- read_rds("saved/ch15_loo_pimdnegmed.rds")
loo_pimdnonmon <- read_rds("saved/ch15_loo_pimdnonmon.rds")


plot(loo_pimd$case_likelihood[loo_pimd$data$count>0],
     loo_pimdnegmed$case_likelihood[loo_pimd$data$count>0], 
     xlab = "likelihood under monotonic model", 
     ylab = "likelihood under unrestricted model", type = "n")
text(loo_pimd$case_likelihood[loo_pimd$data$count>0],
     loo_pimdnegmed$case_likelihood[loo_pimd$data$count>0], 
     loo_pimd$data$event[loo_pimd$data$count>0], cex = .7)
abline(a = 0, b = 1, col = "lightgrey")

#Is last term above correct?

par(mfrow = c(1,2))
plot_loo(loo_pimd, name = "restricted model")
plot_loo(loo_pimdnegmed, name = "unrestricted model")
plot_loo(loo_pimdnonmon, name = "fully unrestricted model")

```
<!-- AJ: Getting error for LOO plots -- can't find plot_loo function. -->

In Figures \@ref(loo_pimdplots), we show the relationship, for each model, of the likelihood of each data type against the number of cases of that data type in the data. A data type is here defined as a possible combination of realized values on all nodes ($I, P, M$ and $D$).  In each plot, the diagonal line represents equality between the proportion of expected cases under the model and the proportion of actual cases. Just eyeballing the relationships, it appears as though the two stronger models might be performing better than the fully unrestricted model.

More formally, we calculate the LOO likelihood for each model as `r formatC(loo_pimd$loo_likelihood, format = "e", digits = 2)` for the model with monotonicity assumptions throughout, `r formatC(loo_pimdnegmed$loo_likelihood, format = "e", digits = 2)` for the model that relaxes the monotonicity assumption for $I \rightarrow M$, and `r formatC(loo_pimdnonmon$loo_likelihood, format = "e", digits = 2)` for the model with no monotonicity assumptions at all. In other words, we see that the most restrictive model performs best, by about an order of magnitude, followed by the least restrictive model, and then the moderately restrictive one.


### Sensitivity to priors

In our model we assumed that $M$ is monotonic in $I$. How much do conclusions depend on this?  We answer the question by comparing our conclusion to what we would conclude withiout this assumption. Our answers depend on the quantity examined.

We first show results for population inference from a mixed methods analysis. As seen in Table XX our inferences regarging the overall effect of $I$ on $D$ are not very sensitive to this assumption. Our conditional inferences are, however. In particular, in cases with $I=0, D=1$ we are more likely to think that democratization was due to *low* inequality given the  unrestricted model. When we see that in fact there was no mobilization, our attribution increases in the restricted model but decreases in the unrestricted model. In the fully unrestricted model our inferences are not affected at all by observation of $M=0$. 

<!-- FLAG: AJ: Check intuition here! -->

In the non-monotonic model we entertain the possibility that low inequality mattered not just directly but also, perhaps, by inducing protests. However, when you observe no protests, you rule out this possible pathway. In the monotonic model you do not think that democratization could have been produced by low inequality via demonstrations---but nevertheless entertain the possibility of mobilization that is *not* due to inequality, which could nevertheless be the cause of democratization. In this case, observing no mobilization removes a rival, cause of democratization, not a second channel. 

In all, we judge the conditional inferences as very sensitive to the monotonicity assumptions we put in place.

```{r 15pimdgraphs, echo = FALSE, fig.cap = "ATE of I on D."}

if(do_diagnosis){
  
  ate_id  <- query_distribution(pimd, "D[I=1] - D[I=0]", using = "posteriors")
  ate_id2 <- query_distribution(pimd_negmed, "D[I=1] - D[I=0]", using = "posteriors")
 

  q_ate_id1 <- query_model(pimd, "D[I=1] - D[I=0]", using = "posteriors", 
                           given = c(TRUE, "I==0 & D==1", "I==0 & M==0 & D==1"))
  q_ate_id2 <- query_model(pimd_negmed, "D[I=1] - D[I=0]", using = "posteriors", 
                           given = c(TRUE, "I==0 & D==1", "I==0 & M==0 & D==1"))
  q_ate_id3 <- query_model(pimd_nonmon, "D[I=1] - D[I=0]", using = "posteriors", 
                           given = c(TRUE, "I==0 & D==1", "I==0 & M==0 & D==1"))

  out <- cbind(q_ate_id1, q_ate_id2[, c("mean", "sd")], q_ate_id3[, c("mean", "sd")])
  names(out) <- c("q", "Given", "u", "Mean (M1)", "sd (M1)", "Mean (M2)", "sd (M2)", "Mean (M3)", "sd (M3)") 
  out <- out %>% select(-c(q, u))

 write_rds(out, "saved/ch15_ateID.rds")

 }

ateid <- read_rds("saved/ch15_ateID.rds") 

kable(ateid, caption = "Posteriors on the effect of I on D depending on whether the to M relation is assumed to be monotonic (M1) or not (M2)", digits = 2)


if(do_diagnosis)
  q_pimd_2 <- query_model(pimd_nonmon_2, "D[I=1] - D[I=0]", given = c(TRUE, "I==0 & D==1", "I==0 & M==0 & D==1", "I==0 & M==1 & D==1",
                                                                    "I==0 & M==0 & D==1 & P == 1", "I==0 & M==1 & D==1 & P == 1"),
                        using = "posteriors")
if(do_diagnosis)
 write_rds(q_pimd_2, "saved/ch15_q_pimd_2.rds")
                        
read_rds("saved/ch15_q_pimd_2.rds") %>% kable(digits = 2)

```

We now consider case-level analysis. For this setup, we consider negative effects of $I$ on $M$ *unlikely*, rather than impossible, and we consider null and positive effects somewhat likely. We refer to these priors as "quantitative priors" in the sense that they place a numerical value on beliefs rather than a logical restriction. Here, we set our prior on $\theta^M$ as: $p(\theta^M=\theta^M_{10})=0.1$, $p(\theta^M=\theta^M_{00})=0.3$, $p(\theta^M=\theta^M_{11})=0.3$, and $p(\theta^M=\theta^M_{01})=0.3$ in comparison to the 0, 1/3,1/3,1/3 distribution. We show the results for the inferences given different findings in Tables \@ref(tab:HK8cases1quant) and \@ref(tab:HK8cases2quant). 



```{r HK8cases1quant, echo = FALSE, eval = TRUE}


pimd_negmed_1 <- set_parameters(pimd_negmed, label = "10", node = "M", parameters = 0)
pimd_negmed_2 <- set_parameters(pimd_negmed, label = "10", node = "M", parameters = .1)

f_given <- function(i, d) {
  a = paste("I ==", i,  " & D == ", d)
  list(a, paste(a, "& M == 0 & P == 0"), paste(a, "& M == 0 & P == 1"), paste(a, "& M == 1 & P == 0"),   paste(a, "& M == 1 & P == 1"))}

q_01_1 <- query_model(
  pimd_negmed_1, 
  query = "D[I=1] != D[I=0]", 
  using = "parameters", 
  given = f_given(0,1)
) 

q_01_2 <- query_model(
  pimd_negmed_2, 
  query = "D[I=1] != D[I=0]", 
  using = "parameters", 
  given = f_given(0,1)
) 

# Do same for nonmon model

q_01 <- select(q_01_1, Given) %>% 
  mutate(
    Restricted   = q_01_1$mean,
    Unrestricted = q_01_2$mean)


q_01 %>%
  mutate(Example = c(" ", "Mexico (2000)", "Taiwan (1996)", 
                  "Albania (1991)", "Nicaragua (1984)")) %>%
kable(caption = "\\label{tab:HK8cases1quant} Four cases with low inequality and  democratization. Question of interest: Was low inequality a cause of democracy? Table shows posterior beliefs for different data for 4 cases given information on $M$ or $P$. Data from Haggard and Kaufman (2012).")



```

```{r HK8cases2quant, echo= FALSE}

q_11_1 <- query_model(
  pimd_negmed_1, 
  query = "D[I=1] != D[I=0]", 
  using = "parameters", 
  given = f_given(1,1)
) 

q_11_2 <- query_model(
  pimd_negmed_2, 
  query = "D[I=1] != D[I=0]", 
  using = "parameters", 
  given = f_given(1,1)
) 

q_11 <- select(q_11_1, Given) %>% 
  mutate(
    Restricted    = q_11_1$mean,
    Unrestricted  = q_11_2$mean)
q_11 %>%
  mutate(Example = c(" ", "Mongolia (1990)", "Paraguay (1989)", "Sierra Leone (1996)", "Malawi (1994)")) %>%
  kable(caption = "\\label{tab:HK8cases2} Four cases with high inequality and  democratization. Question of interest: Was high inequality a cause of democratization? Table shows posterior beliefs for different data for 4 cases given information on $M$ or $P$. Data from Haggard and Kaufman (2012).")

```


The results differ in various modest ways. In the case in which $I=D=1$, conditional inferences are unaffected by the assumption about the effect of $I$ on $M$. The reason is that, since we still maintain a monotonicity assumption for the direct effect of $I$ on $D$, the only question is whether there was an indirect effect. Since we maintain the assumption of a monotonic effect of $M$ on $D$ we know that there was no (indirect) effect if $M$ were 0. If however $M$ were 1, then $I$ did not have a negative effect on $M$ and the only question is whether $M=1$ because of $I$ or independent of it --- which depends only on the relative size ot $\theta^M_{11}$ and $\theta^M_{01}$.

There are bigger differences when we are looking for negative effects of inequality, but even in this case the ordering of inferences does not change. As before in cases where $M=0$, the monotonicity of M in D is not relevant since htere is no question of a negative effect anyhow. However if $M=1$, then it is possible that $I=0$ caused $D=1$ via mobilization in the model that allows non monotonicity and so we put more weight on the possibility that $I=0$ was the cause. 

We see these differences most clearly in the cases of Albania (as compared to Mexico) and Nicaragua (as compared to Taiwan). Under priors fully constrained to monotonic causal effects, we saw that the mediator clue, $M$, made only a small difference to our inferences. However, if we allow for a negative effect of $I$ on $M$, even while believing it to be unlikely, observing mobilization in Albania and Nicaragua makes us substantially more confident that inequality mattered, and differentiates our conclusions about these cases more sharply from our conclusions about Mexico and Taiwan, respectively. 



